---
title: "Trading social status for genetics in marriage markets: evidence from UK Biobank"
author: |
  Abdel Abdellaoui \thanks{Department of Psychiatry, Amsterdam UMC, University 
  of Amsterdam, Amsterdam, The Netherlands. Email: a.abdellaoui@amsterdamumc.nl},
  Oana Borcan \thanks{School of Economics, University of East Anglia, Norwich, 
  UK. Email: O.Borcan@uea.ac.uk} & 
  David Hugh-Jones \thanks{Corresponding author. School of Economics, 
  University of East Anglia, Norwich, UK. Email: D.Hugh-Jones@uea.ac.uk}
abstract: |
  If social status and genetic variants are both attractive in marriage markets,
  then they will be passed on to subsequent generations together. This process
  provides a 
  new explanation for the surprising persistence of inequality across generations,
  and for observed genetic differences across socio-economic status. We model 
  Social-Genetic Assortative Mating (SGAM) and test for its existence in a large
  genetically-informed survey. We compare spouses of individuals with different 
  birth order, which is known to affect socio-economic status and which is 
  exogenous to own genetic endowments among siblings. Spouses of 
  earlier-born siblings have more genetic variants that predict educational 
  attainment. We provide evidence that this effect is mediated by individuals' 
  own educational attainment and income. Thus, environmental shocks to 
  socio-economic status are reflected in the DNA of subsequent generations. 
  SGAM reveals a new aspect of the inheritance of inequality in contemporary and
  historical societies. 
date: "`r Sys.Date()`"
output: 
  bookdown::pdf_document2:
    toc: false
    latex_engine: xelatex
    number_sections: false
    keep_tex: true
editor_options: 
  chunk_output_type: console
  markdown: 
    wrap: 72
bibliography: bibliography.bib
spacing: double
mainfont: Times
header-includes:
  - \usepackage{subfig}
  - \usepackage{setspace}\doublespacing
  - \usepackage{amsmath}
  - \usepackage{amsthm}
  - \usepackage{placeins}
  - \newtheorem{prop}{Proposition}
---


```{r setup, include = FALSE}

library(Formula)
library(car)
library(drake)
library(Hmisc)
library(dplyr)
library(magrittr)
library(AER)
library(ggplot2)
library(huxtable)
library(broom)
library(fixest)
library(santoku)
library(purrr)
library(scales) # percent should override santoku
library(systemfit)
library(nlWaldTest)

set.seed(27101975)

regression_subset <- function (data) {
  data %>% 
         filter(
           n_sibs.x >= 2, 
           n_sibs.x <= 6, 
           ! is.na(birth_order.x)
         )
}


calc_prop_shared_children <- function (mf_pairs) {
  drake::loadd(parent_child)
  # all pairs where at least 1 parent has a genetic child in the sample
  mf_w_parent <- mf_pairs %>% 
        left_join(parent_child, by = c("ID.m" = "parent_id")) %>% 
        left_join(parent_child, by = c("ID.f" = "parent_id"), 
                  suffix = c(".m", ".f")) %>% 
        filter(! is.na(child_id.m) | ! is.na(child_id.f))
  
  
  n_one_has_kid <- nrow(mf_w_parent)
  n_both_same_kid <- mf_w_parent %>% 
                        filter(child_id.m == child_id.f) %>% # NAs excluded
                        nrow()
  return(list(one = n_one_has_kid, both = n_both_same_kid))
}


pretty <- function (n, digits = 2, ...) {
  formatC(n, digits = digits, big.mark = ",", ...)
}


update_with_birth_order_dummies <- function (fml) {
  update(fml, . ~ . - birth_order.x + factor(birth_order.x))
}


knitr::opts_chunk$set(echo = FALSE)
options(huxtable.long_minus = TRUE)
theme_set(theme_minimal())

drake::loadd(mf_pairs)
drake::loadd(mf_pairs_twice)
drake::loadd(famhist)
drake::loadd(resid_scores)

famhist %<>% left_join(resid_scores, by = "f.eid")
rm(resid_scores)
famhist$EA3 <- famhist$EA3_excl_23andMe_UK_resid

mf_pairs_reg <- regression_subset(mf_pairs_twice)

# for p values
my_stars <- c(`***` = 0.001, `**` = 0.01, `*` = 0.05, `+` = 0.10)

# basic formulae
fml_bo_psea <- list()

fml_bo_psea[[1]] <- EA3.y ~ birth_order.x | factor(n_sibs.x)
fml_bo_psea[[2]] <- EA3.y ~ university.x + birth_order.x | factor(n_sibs.x)
fml_bo_psea[[3]] <- EA3.y ~ first_job_pay.x + birth_order.x | factor(n_sibs.x)
fml_bo_psea[[4]] <- EA3.y ~ first_job_pay.x + university.x + birth_order.x |
      factor(n_sibs.x)

fml_bo_psea <- lapply(fml_bo_psea, Formula::as.Formula)

# common controls
fml_bo_psea <- lapply(fml_bo_psea, update, 
                        . ~ . + EA3.x + par_age_birth.x)
fml_bo_psea <- lapply(fml_bo_psea, update, 
                        . ~ . | . + factor(YOB.x) + factor(birth_mon.x))

# alternative mediators
fml_bo_psea[-1] <- lapply(fml_bo_psea[-1], update,
                            . ~ . + fluid_iq.x + height.x)

# basic tidy args
# tidy_args <- list(se = "cluster", cluster = "couple_id", conf.int = FALSE)
tidy_args <- list(se = "hetero", conf.int = FALSE)
my_note <- "{stars}. Standard errors: robust."
```

```{r TODO-NOTES, eval = FALSE}

Nice chunk from earlier version:
Under this mechanism, shocks to an individual’s social status may alter the identity of his or her spouse, and so be reflected in the genetics of his or her children. Thus, in contrast to a large body of research that treats genetic data as an independentvariableandtracesitseffectsonsocialandeconomicoutcomes,wemakethegeneticsendogenous.[^2]

[^2]: Almost all research in behavioural genetics fits this pattern [see
    e.g. @plomin2008behavioral]. For a review of relevant research in
    economics, see @benjamin2011promises. Some more recent economics
    papers use polygenic scores [@barth2020genetic;
    @papageorge2020genes; @ronda2020family; @sias2020molecular].



# * Why missing all the first job data? - because not everyone does the
#   online followup. We could use "current job" which might be endogenous
#   to your spouse; then do first job as a robustness check.
#   
# 
# * maybe - consider alternative exogenous shocks to income. For example, some # professions are more "cyclical" than others wrt recessions. If we could do 
# predicted income at age 21-25 from business cycle X profession, that might 
# count as exogenous. (Could use an independent source to estimate evolution
# of incomes, e.g. GHS or BHPS)
# 
# * Maybe: number of elder *brothers*? (We don't have this info but we have
#   total number of brothers, so we can interact this with the birth order
#   effect)
# 


# Check prop 2. (Bob question: if there's noise, then shouldn't
# correlation tend to a limit less than 1? And if so, then at that limit,
# parental correlation should equal child correlation).
# - I think the claim doesn't contradict this. But do double-check!


# TODO: We may have to control, rather than residualize, on PCs
# to get accurate p-values, as our N is not huge by the end (or is 2000 
# enough not to worry)

# Connect the model to existing theoretical models of matching 
# (e.g. Gale-Shapley, Becker 1974, econometrics of matching). DHJ.

# Robustness: look at all 33 polygenic scores for families of size 3. Rerun
# regressions excluding family size of 3. Appendix. DHJ.

# For future: consider gender asymmetries - e.g. does male SES matter more?
# Might need some other "good genes" measures e.g. waist-hip ratio.

# Maybe full set of birth order dummies within family size
# but maybe only for family size 2-4 or so?
# 

# * NB: why in mf_pairs_twice do we see differences in own EA3 with birth_order?
#   - just chance? it's only true for families of size 3...


# NOTES on economics literature
# Areas/topics
# * Econometrics of matching
#   - Choo & Siow 2006
#   - Chiappori et al 'fatter attraction'
#     - Quite simple regressions as a result of their basic framework
#     - Run regressions on each of the other person's characteristics;
#       in the linear version of their framework, coefficients should all be
#       proportional
#   - I'm not sure these guys are our key targets though, more likely some
#     people we might have to "appease"...!
# * Inheritance of inequality
# * Assortative mating
#   - a mechanism for inequality
#   - also related to family economics
#   - this might be a good way to set up the paper:
#   - "We bring together two explanations of persistent inequality. (1) Genetics
#     and (2) assortative mating."
#  * Getting comments from Greg Clark would be very useful.
#  * Maybe from Samuel Bowles too?
#  * They have a Science article on IG Wealth Transmission
#  * This special issue of Current Anthropology is important:
#    https://www.journals.uchicago.edu/toc/ca/2010/51/1
#    - they distinguish "material", "somatic" (inc genetic, but also e.g.
#      embodied knowledge), and "relational" wealth
#    - In their response to comments, Mulder et al. acknowledge that maybe 
#      one kind of wealth can help you acquire other kinds, and this is an
#      important area of future research.
#  * Fernandez et al. (Love and Money)
#    - inequality increases marital sorting because it increases the benefits
#      to matching with another skilled type;
#    - they confirm the correlation cross-country
#    - they also predict sorting lowers GDP (not sure how this works)
#    - there is a link with macroeconomics and the demographic transition in
#      this literature
#  * Eika et al. 2019
#    - Empirics of assortative mating in US from 1940s: it's increased at 
#      the bottom but gone down at the top
#    - Important for inequality but changes have led to little net increase
#      in inequality. (A simple accounting framework when you compare "what
#      if people had matched randomly into households")
#    - The motivation "assortative mating is increasing inequality" is 
#      a sufficient one, for this paper - it's a big issue.
#  * Doepke and Tertilt "families in macroeconomics"
#    - parents' fertility decisions are important for econ growth
#    - as is skill formation
#    - I think this is for another paper - e.g. what happens when the rich
#      have more children?
#  * Gould, Moav and Simhon 2008
#    - to read 
#    - relates end of polygyny to econ growth
#    - as above, maybe for another paper
#      
# PEOPLE you could talk to
# * Doepke
# * Clark
# * Fernandez
# * Bowles
```

\normalem

# Introduction

Inequality persists over generations.
Intergenerational mobility is correlated with cross-sectional inequality
[@becker2018theory; @krueger2012rise], which has risen dramatically in
high-income countries [@western2008inequality].[^stats]  At the same time
intergenerational absolute mobility has fallen [@chetty2017fading], although
relative mobility has been stable [@chetty2014united]. Over the long run,
inequality is surprisingly persistent across generations
[@clark2015intergenerational; @solon2018we]. To explain inequality and its
persistence, social scientists have analysed assortative mating (homogamy) in
the marriage market [e.g. @breen2011educational; @greenwood2014marry] and
increasing returns to human capital [e.g. @kaplan2013s; @becker2018theory]. It
follows that how families are formed, and how they transmit traits and assets to
their offspring, are critical for understanding inequality. These processes
have been studied from both socio-economic and genetic angles. While educational
homogamy is well established, genetic assortative mating has been demonstrated
only recently [@hugh2016assortative]. Similarly, wealthy families pass on
advantages to their children through both genetic inheritance and environmental
influence [@Rimfeld_2018; @bjorklund2006origins].[^review]

[^stats]: In the United Kingdom the Gini coefficient has increased from 26% to 34.6% between 1977 and 2020. The United States has seen a 10 percentage point rise up to 43.3% during 1962-2013.

[^review]: See @sacerdote2011nature for a review of the behavioural genetics and
economics literatures on the nature vs nurture debate; for a broader review of
the studies on intergenerational transmission of income see @black2010recent.


This paper examines a plausible, but under-analysed, aspect of the spouse
matching process: that both social status and genetics contribute to a person’s
attractiveness in marriage markets, and as a result, genetics and inherited
social status may become associated in subsequent generations.[^status-SES] For
example, suppose that wealth and intelligence are both attractive qualities in a
spouse. Then wealthy people are more likely to marry intelligent people, and
their children will inherit both wealth, and genetic variants associated with
intelligence. We call this mechanism Social-Genetic Assortative Mating (SGAM).

[^status-SES]: *Social status* refers to characteristics that an individual
  possesses in virtue of their social position. For example, my wealth
  is a fact about me that holds in virtue of my relationship to
  certain social institutions (bank deposits, title deeds et cetera.).
  Other examples include caste, class, income, and educational
  qualifications. *Socio-economic status* (SES) is a specific type of
  social status which exists in economically stratified societies, and
  which refers to a combination of educational attainment,
  occupational class, income and wealth [e.g. @white1982relation].


SGAM may be an important channel for the transmission of inequality. It leads
to a hidden dimension of advantage for privileged families – hidden because most
social science datasets do not include genetic information. This dimension may
help to explain the surprising long-run persistence of inequality
[@clark2015intergenerational]. At the same time, this advantage is not an
exogenous fact of biology, but endogenous to the social structure. Indeed, under
SGAM, shocks to an individual’s social status may be reflected in the genetics 
of his or her children.

Below, we first outline a theoretical framework where attractiveness in the
marriage market is a function of both socio-economic status (SES) and genetic
variants. We show that Social-Genetic Assortative Mating in one generation leads
to increased correlation between social and genetic traits in the offspring
generation. This result provides a new explanation of the association between
genetics and socio-economic status [@belsky2018genetic; @Rimfeld_2018;
@bjorklund2006origins]. While existing explanations have focused on meritocratic
social mobility (genes cause SES), under SGAM causality goes both ways, from
genes to SES and vice versa.

Next, using novel data on matched spouses born between 1935 and 1970 from the UK
Biobank, we empirically test the hypothesis that an individual’s higher social
status attracts spouses with higher genetic potential for educational
attainment. Our genetic measure, the Polygenic Score for Educational Attainment
(PSEA), derives from large-scale genome-wide association studies [@lee2018gene]
and is causally related to educational attainment itself, as well as to
intelligence and labour market outcomes. It is already known that humans mate
assortatively on PSEA, which makes it a likely candidate for detecting SGAM
[@hugh2016assortative; @robinson2017genetic]. We depart from the assumption that
both socio-economic status and genetic traits can enhance the attractiveness of
potential spouses, and are substitutable. This assumption has received support
in recent economic studies of marriage markets, which suggest that people trade
off physical characteristics for higher earnings [e.g. @chiappori2012fatter] or
matching social class [@banerjee2013marry].

The endogeneity of social status is the main challenge in identifying the causal
effect of social status on the spouse’s genetic endowment. For instance,
individuals with high education qualifications tend to also have high
educational attainment genes, and as mentioned above, they may take partners
based on genomic similarity. In order to isolate the causal link from own
socio-economic status to partner genes, we use the “accident of birth”
as a source of exogenous variation in socio-economics status. Specifically, we
use the birth order of individuals in the sample as a “treatment” which affects
their partner choice through a range of mechanisms, of which the most salient
one is own socio-economic status. It is well documented that earlier-born
children enjoy higher parental investment and have better life outcomes,
including measures of socio-economic status such as educational attainment and
occupational status [@black2011older; @booth2009birth; @Lindahl_2008]. At the
same time, birth order is independent of siblings’ genetic endowments, a fact
guaranteed by the biological mechanism involved (the “lottery of meiosis”).

While birth order is plausibly independent of siblings’ genes, it cannot be used
as a valid instrument for socio-economic status, because it may affect partner
choice through alternative mechanisms. Hence, we steer away from a two-stage
procedure and instead rely on a mediation analysis similar to
@heckman2013understanding, who decompose the average treatment effect into
effects of measured and unmeasured consequences of treatment. Specifically, we
estimate a reduced-form model with spouse genes associated with educational
attainment as the dependent variable, and own birth order as the main
independent variable. We then estimate a model which also includes measures of
own socio-economic status. In the latter model, socio-economic status can be
interpreted, under certain assumptions, as a mediator of the effect of birth
order on spouse genetics. We also include controls to balance covariates across
individuals with different birth orders in different cohorts.

We find that a higher own birth order significantly lowers the spouse’s
polygenic score for educational attainment in the reduced-form regressions. When
we include university attendance as a proxy for socio-economic status, birth
order is no longer significant, while university attendance increases the
spouse’s genetic endowment, again at 0.1% significance. A similar pattern holds
when we proxy socio-economic status with a measure of income, although the
sample is reduced. Thus, birth order appears to affect spouse genetics via the
mediator of SES. The results are robust to the inclusion of several own
phenotype traits and a rich set of own genetic traits.

Our paper contributes to several literatures. Firstly, we highlight a novel
mechanism of assortative mating. The economics literature on matching in
marriage markets has typically focused on educational similarities 
[e.g. @pencavel1998assortative] or social class or caste 
[e.g. @abramitzky2011marrying; @banerjee2013marry], but also sorting based on age,
physical traits and ethnicity [@hitsch2010matching]. Matching decisions on the
marriage market have also been shown to follow multiple criteria, with some
degree of substitutability between them.[^others] For instance, @chiappori2012fatter
showed that individuals trade off BMI for partners’ income or education and that
the marginal rate of substitution between these characteristics is different for
males and females.  The genetics literature has focused on genetic assortative
mating (GAM), the phenomenon that people with similar genes marry each other.
Recent research has confirmed the long-standing conjecture that GAM takes place
in contemporary human populations 
[@howe2019genetic; @hugh2016assortative; @robinson2017genetic]. Geneticists have 
also developed the concept of cross-trait assortative mating 
[@Beauchamp_2010;@Sundet_2005], which refers to people with (genes for) e.g.
height marrying people with (genes for) e.g. intelligence. As a result, the two
types of variation become associated. In this paper we bring the two literatures
together, extending the idea of cross-trait assortative mating to encompass both
socially inherited status, and biologically inherited genetic variants. Our
results confirm that individuals with higher social status are more likely to
attract a spouse with higher innate cognitive ability.

[^others]: @oreffice2010anthropometry show that height and BMI are associated with spouse earnings. @dupuy2014personality find spouse matching on multiple independent dimensions, including education, height, BMI and personality.


Secondly, our findings have implications for understanding the sources of
economic inequality. There are several ways assortative mating can affect
inequality [@fernandez2001sorting]. The simplest is that if couples assort with
respect to some characteristic, the resulting households will have more variance
in that characteristic than if couples match randomly. For instance, if rich
people only marry each other, then the wealth distribution of the resulting
households will be more unequal than if they sometimes marry poor people. If the
characteristic is inherited from parents to children, then this will carry over
into higher inter-individual inequality in the next generation. Assortative
mating on one dimension can also affect inequality in other dimensions. For
example, if parents invest in their children’s human capital, and if households
are credit-constrained, then assortative mating on income can increase
inequality in children’s human capital [@fernandez2005love]. The mechanism is
purely environmental, not genetic: some children are raised in a higher-quality
environment (e.g. with more books and educational toys), and variance in these
environments is increased by assortative mating. Research has shown that
assortative mating is associated with higher inequality within generations,
although there is no conclusive evidence that it underpins the inequality trend
over time [@eika2019educational].

A more likely driver of the rise in inequality is the increase in market returns
to human capital [e.g. @kaplan2013s; @eika2019educational]. In this context, the
distribution of human capital is a key contributor to inequality, in addition to
inherited wealth.[^wealth] Of course, one part of human capital is acquired and
the other is genetic. From twin studies, the heritability of occupational class
and educational attainment, i.e. the proportion of variance explained by genetic
differences between individuals, is around 50% [@Tambs_1989]. Genome-wide
Complex Trait Analysis (GCTA) shows that 2-year-old children’s family
socio-economic status can be predicted from their genes [@Trzaskowski_2014].
Studies comparing parent-child income and education associations between
adoptees and non-adoptees show that both post-birth environment and pre-birth
conditions (genetics and to a lesser extent prenatal environment) contribute to
the transmission of wealth and human capital [e.g. @bjorklund2006origins].
Several channels shape the distribution of heritable ability. Parents with
higher ability reap higher market returns, and more successful parents may then
pass both higher socio-economic status and their genes to their children,
leading to an association between the two [@belsky2018genetic]. This mechanism
depends on the level of meritocracy in social institutions
[@branigan2013variation; @Heath_1985]; in a society where social status was
ascribed rather than earned, it could not take effect. Indeed, after the fall of
communism in Estonia, the heritability of SES increased, presumably because
post-communist society allowed higher returns to talent [@Rimfeld_2018].
Social-genetic assortative mating provides a complementary explanation for the
association between genes and SES, one which does not require social meritocracy.

[^wealth]: See @adermon2018intergenerational, @mulder2009intergenerational.


SGAM may also increase social inequality overall, if there are complementarities
between genetic and environmental components of human capital. For example,
higher-ability parents may make more productive investments in children’s human
capital [@cunha2007technology; @cunha2010estimating; @heckman2014economics; @kong2018nature]. @becker2018theory demonstrated how inequality
can arise in a model of intergenerational transmission of human capital where
high-income and human capital parents are able to invest more in their
children’s human capital than low-income parents. Thus, by bringing good genes
and enriched environments together, SGAM may increase inequality in the next
generation.

Lastly, we contribute to a literature in economics that examines the
relationship between genetic and economic variables. @benjamin2011promises is an
early review. Several more recent papers use polygenic scores, in particular
polygenic scores for educational attainment [@barth2020genetic;
@papageorge2020genes; @ronda2020family]. These papers - like the vast majority
of the behavior genetics literature [see e.g. @plomin2008behavioral] - take
genetic endowments as exogenous and examine how they affect individual outcomes,
perhaps in interaction with the environment. We take a different approach by
putting genetics on the "left hand side". Thus, our paper challenges the
assumption, in economics and beyond, that genetic endowment is exogenous to
economic characteristics. While this may be tenable in within-generation
studies, it ceases to hold in intergenerational models. Social-genetic
assortative mating is a causal mechanism going from socio-economic status to
genetic traits. Furthermore, our model shows that the strength of this mechanism
varies with the structure of the society’s marriage market. When both genes and
status are both relevant in marriage markets, then they become associated in the
next generation. When social status is irrelevant and only good genes matter –
or vice versa – then genes and status do not become associated. Social structure
matters for the redistribution of genes, just as for other forms of capital.

The observations behind SGAM are nothing new. That status and physical
attractiveness assort in marriage markets is a commonplace and a perennial theme
of literature. In the Iliad, powerful leaders fight over the beautiful
slave-girl Bryseis. In Jane Austen’s novels, wealth, attractiveness and “virtue”
all make a good match. @marx1844economic wrote “the effect of ugliness, its repelling
power, is destroyed by money.” And Donald Trump claimed: “part of the beauty of
me is that I am very rich.” The literature on mate preference from evolutionary
psychology [@buss1986preferences; @buss1989sex; @buss2019mate] confirms
that attractive mate characteristics include elements of social status (“high
earning capacity,” “professional status”) as well as traits that are partly
under genetic influence (“intelligent,” “tall,” “kind,” “physically
attractive”). Despite this, we have found almost no previous work in genetics or
economics that formalises the mechanism and consequences of SGAM.
@halsey1958genetics showed in a two-class model that social mobility combined
with assortative mating might increase the association between genetics and
social class. @belsky2018genetic offer three reasons for the association between
education-linked genetics and SES, but do not consider SGAM.


```{r literature-notes, eval = FALSE}

# Notes 

# Fernandez et al 2005
# - skills are caused by investment in human capital
# - sorting is inefficient since it increases inequality of parents' capital
# and some people will be credit-constrained.
# 
# Marry Your Like (Greenwood 2014)
# - simple accounting methodology
# - ass mating has increased and has contributed to increase in inequality in the 
#   US, 1960 to 2005
#   
# Eika 2019
# - assortative mating has decreased among highly educated, increased among low
#   educated, 1960s-2010s. Overall, increase only until the 1980s
# - changes in ass mating barely move time trends in household income inequality
# - findings for US + 4 European countries

# Clark 201x - surname groups show higher intergen transmission of wealth 
# than 
# individual parent-child pairs. Also grandparents matter independently. 
# One explanation: underlying "social status" or "social competence" which
# is measured with error by wealth.

# Solon 2018 - reviews evidence on long-run transmissibility. Not
# much evidence for Clark's "0.7". Many alternative explanations.

# Becker-Tomes 1979?
```


# Model

There is a large population, whose members have a single genetic trait
$g_{i}$ and a single social trait $s_{i}$, drawn from distributions $G$
and $S$. The genetic trait could be, for instance, a polygenic score,
which summarizes the effects of many alleles (genetic variants) at
different loci. The social trait is a measure of social status. Broadly
conceived, this means any trait that an individual possesses in virtue
of his or her position in society, rather than as a natural fact. Caste
and class are kinds of social status; so are wealth, income, education
and employment.

Without loss of generality, $EG = ES = 0$ and $Var(G) = Var(S) = 1$.[^3]
People pair according to an attractiveness function

[^3]: Continuous distribution is not strictly required. All that is
    needed is for a set of pairs of positive measure to have different
    values of $G$ and $S$, along a set of attractiveness curves of
    positive measure.

$$
A(g_{i},s_{i}) = f((1-k)g_{i},ks_{i})
$$

where $f$ is smooth and strictly increasing in both its arguments, and
$0 \le k \le 1$. Our sole condition on $G$ and $S$ is that a positive
measure of the population has attractiveness $A$ where the distribution
of $(G, S)|A$ is non-degenerate, i.e. not everybody with attractiveness
$A$ is both genetically and socially identical. We assume that people
marry at random within their attractiveness isoquant. One
microfoundation is that $A$ measures the utility a person provides as a
spouse, and the marriage market works via a Gale-Shapley
[-@gale1962college] matching mechanism. Or, $A$ could represent inputs
into a household production function where male and female inputs are
complementary [@becker1973theory]. Other interpretations are possible:
for instance, different values of $A$ could represent segmented marriage
markets, like different schools or workplaces.

The key parameter is $k$. This describes the working of the society's
marriage market. If $k = 0$, only genetics $G$ are relevant in marriage
markets, and social status $S$ has no effect. That is, the marriage
market is highly egalitarian. Conversely, if $k = 1$, only social status
matters, to the complete exclusion of genetics. This is the equivalent
of a marriage market driven only by "caste" or "class". Realistic
societies are between these extremes.

Figure \@ref(fig:pic-intuition) shows the intuition behind our theory.
The top row shows a *caste marriage market* with $k = 1$. A typical pair
is shown: children have intermediate values of $G$ and $S$ between their
two parents (hollow circle). In this society pairs match only by social
status; genetics plays no role. As a result, while the variance of $G$
shrinks within each status group, genetics remain uncorrelated with
social status in the children's population distribution, shown on the
right. The next row shows a pure *egalitarian marriage market* with
$k = 0$. Parents match only by genetics and ignore social status. Again,
as a result there is no correlation between genetics and social status
in the children's generation. The bottom row shows a (more realistic)
intermediate society, with an intermediate value of $k$. Because both
genetics and social status contribute to attractiveness, matched spouses
typically trade them off against each other. As a result, the
distribution is squeezed along the gradient of $k$, and $G$ and $S$ are
correlated in the children's generation. We next prove this formally.

If $k = 0$, "indifference curves" of attractiveness are vertical lines
in $(G, S)$ space. If $k = 1$, they are horizontal lines. If
$k \in (0,1)$ they are downward sloping curves.

Write $p(i)$ for $i$'s partner. Pairs always have the same
attractiveness:

```{=tex}
\begin{equation}
A(g_{i},s_{i}) = A(g_{p(i)},s_{p(i)}),
\label{eq:same-A}
\end{equation}
```
Each pair has two children. We assume that both children $c(d)$ of
parents $d$, $m$ have

```{=tex}
\begin{align}
g_{c(d)} & =\frac{g_{d}+g_{m}}{2};\label{eq:children-average}\\
s_{c(d)} & =\frac{s_{d}+s_{m}}{2}.\nonumber 
\end{align}
```
This is a strong simplifying assumption; we relax it below. For real
world examples approximated by the model, $S$ could be wealth which is
equally divided between the children; $G$ could be a highly polygenic
trait with many small effects. Write $G_{p}, S_{p}$ to denote the
population variables in the parents' generation; $G_{c}, S_{c}$ for the
children's generation.

```{=tex}
\begin{prop}
\label{prop:prop1}
(i) $Cov(G_{c}, S_{c}) \ge Cov(G_{p},S_{p})$, with strict inequality
if and only if $0 < k < 1$. 

(ii) If $corr(G_{p}, S_{p}) \ge 0$, then $corr(G_{c},S_{c}) \ge corr(G_{p},S_{p})$,
with strict inequality if and only if $0 < k < 1$ \uline{or} 
$corr(G_{p}, S_{p}) > 0$.
\end{prop}
```
```{r pic-intuition, fig.subcap = c("Caste society ($k = 1$): parents", "Caste society: children", "Egalitarian society ($k = 0$): parents", "Egalitarian society: children", "Intermediate society ($0 < k < 1$): parents", "Intermediate society: children"), fig.cap = "Theory: shaded area is the population distribution. Dotted lines are attractiveness isoquants. Solid dots are example parents, transparent dots are example children. The right hand side shows the children's generation.", fig.ncol = 2, fig.width = 2.75, fig.height = 2.4, fig.align='center'}


setup_plot <- function () {
  par(mar = rep(2.1, 4))
  plot.new()
  plot.window(c(-1.2, 1.2), c(-1.2, 1.2))
  arrows(x0 = -1.1, x1 = 1.1, y0 = -1.1, length = .1)
  arrows(y0 = -1.1, y1 = 1.1, x0 = -1.1, length = .1)
  text(-1.1, 1.2, "S", cex = 0.7)
  text(1.2, -1.1, "G", cex = 0.7)
}

draw_dist <- function (w) {
  if (! is.matrix(w)) w <- matrix(c(w, 0, 0, 1), nrow = 2)
  mygrey <- grey(.85, .03)
  for (rad in seq(0.2, 1, 0.1)) {
    v <- car::ellipse(c(0, 0), shape = w, radius = rad, add = TRUE, center.pch = NULL,
               fill = TRUE, col = mygrey, lwd = 0, segments = 241)
  }
  
  v
}

draw_dist_with_a <- function (width, shift, curved_lines = FALSE) {
  setup_plot()  
  v <- draw_dist(width)
  nr_v <- nrow(v)
  
  rotation <- (-shift + seq_len(nr_v)) %% nr_v
  rotation[rotation == 0] <- nr_v
  
  v <- v[rotation, ]
  v_rev <- v[nr_v:1, ]
  
  ivls <- quantile(seq_len(nr_v/2), seq(0.2, 0.8, .2), type = 1)
  
  start_x <- v[ivls, 1]
  end_x   <- v_rev[ivls, 1]
  start_y <- v[ivls, 2]
  end_y <- v_rev[ivls, 2]
  
  if (curved_lines) {
    for (l in seq_along(start_x)) {
      sx <- start_x[l] + c(0, 0.15, 0.85, 1) * (end_x[l] - start_x[l])
      sy <- start_y[l] + c(0, 0.05, 0.65, 1) * (end_y[l] - start_y[l])
      spl <- stats::spline(sx, sy)
      if (l ==2) mid_spl <- spl
      lines(spl, lty = 2)
    }
    
    start_x <- mid_spl$x[2]
    start_y <- mid_spl$y[2]
    end_x <- mid_spl$x[8]
    end_y <- mid_spl$y[8]
    inset <- 0
  } else {
    segments(x0 = start_x, x1 = end_x, y0 = start_y, y1 = end_y, lty = 2)  
    
    ivls <- ivls[2:3]
  
    start_x <- mean(v[ivls, 1])
    end_x   <- mean(v_rev[ivls, 1])
    start_y <- mean(v[ivls, 2])
    end_y   <- mean(v_rev[ivls, 2])
    inset <- 0.15
  }
  
  
  along <- function (tt) list(
    x = start_x + tt * (end_x - start_x),
    y = start_y + tt * (end_y - start_y)
  )
  
  end_inset <- 0.15
  starts <- along(c(inset, 1 - inset))
  ends   <- along(c(0.5 - end_inset, 0.5 + end_inset))
  pts <- along(c(inset/2, .5, 1 - inset/2))
  arrows(x0 = starts$x, x1 = ends$x, y0 = starts$y, y1 = ends$y,
         length = 0.05)
  
  points(x = pts$x, y = pts$y, pch = c(19, 21, 19), cex = 0.8)
}

draw_dist_with_a(.8, 0)
setup_plot()
invisible(draw_dist(matrix(c(.4, 0, 0, 1), nrow = 2)))

draw_dist_with_a(.8, 60)
setup_plot()
invisible(draw_dist(matrix(c(.8, 0, 0, .4), nrow = 2)))

draw_dist_with_a(.8, -30, curved_lines = TRUE)
setup_plot()
invisible(draw_dist(matrix(c(1, .4, .4, .8), nrow = 2)))



```

We view $k = 0$ and $k = 1$ as theoretical "ideal types". Proposition
\ref{prop:prop1} therefore shows that in almost any realistic society,
social status will become correlated with genetic traits which are
considered attractive in marriage markets.

We now relax the condition that children are exactly at the mean of
their parents' values for $G$ and $S$. Let

```{=tex}
\begin{align*}
g_{c(i)} & = \bar{g}_{i}+\varepsilon_{i}^{G} \\
s_{c(i)} & = \bar{s}_{i}+\varepsilon_{i}^{S}
\end{align*}
```
where

$$
\bar{g}_{i} = \frac{g_{i} + g_{p(i)}}{2}; \bar{s}_{i} = \frac{s_{i} + s_{p(i)}}{2};
$$

$\varepsilon^{G}$ has mean 0 and variance $\sigma_{G}^{2}$; and
$\varepsilon^{S}$ has mean 0 and variance $\sigma_{S}^{2}$.

```{=tex}
\begin{prop}
\label{prop:robustness}

\begin{enumerate}

\item If $\sigma_{G}^{2}$ and $\sigma_{S}^{2}$ are small enough and $corr(G_{p},S_{p}) \ge 0$,
then $corr(G_{c},S_{c}) > corr(G_{p},S_{p})$ for $k \in (0,1)$.

\item If $\varepsilon^{G}$ and $\varepsilon^{S}$ are uncorrelated with
each other and with $\bar{G}$ and $\bar{S}$; and if $G_{p}$ and
$S_{p}$ are uncorrelated, then $corr(G_{c},S_{c}) \ge 0$, with strict
inequality if and only if $0<k<1$.

\end{enumerate}

\end{prop}
```
The conditions in Proposition \ref{prop:robustness} are quite plausible.
For $G$, they require that either variance in siblings' scores on some
summary statistic is not too large, or that it is uncorrelated with the
parents' scores. Both of these hold for most polygenic scores, which are
additive sums of many small effects of alleles derived randomly from one
or other parent. For $S$, the conditions would hold, for example, if $S$
measures wealth, which is inherited not too unequally between siblings;
or if wealth is inherited unequally but not in a way that correlates
with $S$ or $G$.

It is worth considering what kind of social arrangements would *violate*
these conditions. For example, suppose that parents' combined wealth is
inherited by the child with the lowest value of $g_{c(i)}$. This creates
a negative correlation between $s_{c(i)}$ and $g_{c(i)}$.

In the model, intermediate values of $k$ drive increased covariance
between genetics and social status. That naturally raises the question
whether the change in covariance and/or correlation is increasing in $k$
towards some maximum value in $(0,1)$, then decreasing, i.e., whether it
is quasiconcave in $k$. In general the answer is no, even on the
assumption that

```{=tex}
\begin{equation}
A(g_{i},s_{i}) = (1-k)g_{i} + ks_{i}
\label{eq:linearity}
\end{equation}
```
i.e. that indifference curves are straight lines in $(G, S)$ space.
However, quasiconcavity in $k$ does hold when indifference curves are
straight and $(G, S)$ are jointly normally distributed. We conjecture
that it also holds whenever the distribution $(G, S)$ is log-concave or
even quasiconcave.

## Discussion

The "marriage market" here is a reduced form mechanism, encompassing
everything that makes a difference to partner choice. For example, if
earned income affects attractiveness in the marriage market, then
society's level of meritocracy in the labour market will correlate with
the value of $k$: a more meritocratic labour market will allow people
with low social status but high human capital (genetically determined in
part) to earn more, and therefore to match with more attractive
partners.

The contents of both $S$, social status, and $G$ -- "good genes" in the
marriage market -- are likely to vary across societies. $S$ could
encompass variables like social class or caste; ethnic identity in
"ranked" ethnic systems; or in modern societies, SES. Regarding $G$,
standards of physical attractiveness, and other characteristics which
make someone a "good match", vary both across societies and within a
society over time.

Recent empirical work shows high persistence of SES over time, in
particular at the top. One suggested reason for this is that unmeasured
family characteristics persist along with measured wealth
[@clark2015intergenerational]. Our model captures this idea. For
simplicity, assume the original model with no noise in children's values
of $g$ or $s$ (condition (\@ref(eq:children-average))), and let
condition (\@ref(eq:linearity)) hold. Consider a regression of
children's social status on parent's social status:

$$
s_{c(i)} = \alpha + \beta_S s_i + \varepsilon_{c(i)}.
$$

The value of $\beta_S$ will be less than one, since parents match on
downward sloping isoattractiveness curves: within each curve, relatively
wealthy individuals match with less wealthy individuals on average. Now
consider a regression of child on parent "attractiveness" $A$:

$$
A_{c(i)} = \alpha + \beta_A A_i + \varepsilon_{c(i)}.
$$

Since children are on the same attractiveness curve as both their
parents, $\beta_A = 1$. Thus, regressions on measured components of
social status will underestimate true persistence over time, embodied in
genetic variation. Indeed, grandparents' social status will
independently predict grandchildren's social status, even after
controlling for parents' social status, because of the unmeasured
pathway via parents' genetics.

The model predicts variation in the strength of SGAM. In particular, in
"caste societies" where there is complete endogamy within social status
groups, there is no scope for SGAM, because marriage partners do not
trade off genetics for social status. The model also assumes that social
status is inherited randomly from one parent, in the same way a genetic
allele is inherited. This assumption can be weakened. For example, if
social status is inherited deterministically from the father, then the
results remain unchanged (for each pair of parents, just assume that one
randomly chosen parent is the father).

In modern societies, both SGAM and meritocratic mobility are likely to
be at play. Genetic variants that cause (e.g.) higher income and wealth
will be inherited along with components of social status such as
inherited wealth. At the same time, higher social status and "good
genes" will assort in the marriage market, even if that higher social
status is caused by purely environmental variation. Our empirical
analysis shows this latter process at work.

# Data and methods

To test the theory, we use data from the UK Biobank, a study of about
500,000 individuals born between 1935 and 1970. The Biobank contains
information on respondents' genetics, derived from DNA microarrays,
along with questionnaire data on health and social outcomes.

```{r calc-validate-pairs}

prop_shared <- calc_prop_shared_children(mf_pairs)
n_one_has_kid <- prop_shared$one
n_both_same_kid <- prop_shared$both

```

The Biobank does not contain explicit information on spouse pairs. We
categorize respondents as pairs if they:

-   had the same home postcode on at least one occasion;[^4]
-   both reported the same homeownership/renting status, length of time
    at the address, and number of children;
-   attended the same UK Biobank assessment center on the same day;
-   both reported living with their spouse ("husband, wife or partner");
-   consisted of one male and one female.

[^4]: A typical UK postcode contains about 15 properties.

```{r estimate-false-positives, eval = FALSE}
# We can use the number of duplicates to estimate our "false positive"
#   ratio. Suppose that a proportion x of our pairs are false positives who 
#   just live in the
#   same postcode. Suppose UKBB samples a proportion y of people in each postcode. 
#   Then the number of duplicates will be xy. 
#
# Note to calculate xy we need to count *whenever* they 
# appear in the mf_pairs file. I.e. if they are false positives, we want
# to catch them matching with someone else.


# code to estimate y:
# in the alto_file, count how many people are observed at the same coordinate
alto <-readr::read_csv(alto_file)

# count how many people are *not* in the alto file (i.e. for them,
# people_at_same_coordinate should be 1. Indeed in alto, the min is 2.)
# 500K is about right :-)
not_in_alto <- 500000 - length(unique(alto$ID))

all_per_coord <- c(alto$people_at_same_coordinate[! duplicated(alto$coordinate_ID)]
                   , rep(1, not_in_alto))
n_at_same <- mean(all_per_coord)

# this is "y" above:
sample_prop_per_coord <- n_at_same / 15 # 15 people in a postcode on average

# to calculate xy:
# start with our "purified" sample mf_pairs, but without deleting any duplicates
# count how many are duplicates in the sense that they are ever at the 
# same coordinate with someone else

mf_pairs_tmp <- readr::read_csv(mf_pairs_file)

mf_pairs_tmp$eid.x <- mf_pairs_tmp$ID.m
mf_pairs_tmp$eid.y <- mf_pairs_tmp$ID.f
mf_pairs_tmp <- add_data_to_pairs(mf_pairs_tmp, famhist, resid_scores, 
                                suffix = c(".m", ".f"))

mf_pairs_tmp %<>% filter(
                f.680.0.0.f  == f.680.0.0.m,  # same rent/own status
                f.699.0.0.f  == f.699.0.0.m,  # same length of time in hh
                f.709.0.0.f  == f.709.0.0.m,  # same n occupants of hh
                n_children.f == n_children.m, # same n kids
                f.54.0.0.f   == f.54.0.0.m,   # same assessment center 
                f.53.0.0.f   == f.53.0.0.m,   # attended assessment same day
                f.6141.0.0.f == 1,            # both living with spouse
                f.6141.0.0.m == 1,            #   "     "     "    "
                female.m != female.f,         # heterosexual couples only
              )

mf_pairs_ids <- unique(c(mf_pairs_tmp$ID.m, mf_pairs_tmp$ID.f))

# observed duplicate proportion, using the alto ID. This also counts
# people of the same sex, is that right?

dupe_ids <- intersect(mf_pairs_ids, alto$ID[duplicated(alto$ID)])
# this is "xy" above
observed_dupe_prop <- length(dupe_ids)/length(mf_pairs_ids)

# so "x" above is: hmm, not quite right yet, this is > 1!!
estimated_false_pos <- observed_dupe_prop/sample_prop_per_coord

```

We also eliminate all pairs where either spouse appeared more than once
in the data. This leaves a total of `r pretty(nrow(mf_pairs))` pairs. Some of
these could be false positives, i.e. people who are not each others'
spouse but simply live in the same postcode. To validate the accuracy of
our measures, we use genetic relationships. Some respondents in the
Biobank sample have a child who is also in the sample, as inferred from
genetic data. Among our spouse pairs, `r n_one_has_kid` have a genetic
child of at least one partner in the sample. For `r n_both_same_kid` of
these, the child is the genetic child of both partners. If this
subsample is representative, then at least
`r percent(n_both_same_kid/n_one_has_kid)` of the pairs who have had a
child, have had a child together. This is a lower bound, because those
who had a child with someone else may also have had a child with the
presumed partner in our data. As a point of comparison, 11% of families
with dependent children included a stepchild in England and Wales in
2011 [@ons2011stepfamilies].

It is still possible that some pairs in our data may not be actual spouses. In
the appendix, to sign any possible bias in our estimates resulting from this, we
use a dataset of "known fake" pairs. We show that estimated coefficients of
interest are closer to zero among these fake pairs than among our candidate
"real pairs". Because of this, any fake pairs remaining in our data are likely
to bias our coefficients towards zero.

Our key dependent variable is spouse's *Polygenic Score for Educational
Attainment* (PSEA). A polygenic score is a DNA-derived summary measure
of genetic risk or propensity for a particular outcome, created from
summing small effects of many common genetic variants, known as Single
Nucleotide Polymorphisms (SNPs). We focus on PSEA rather than other
polygenic scores for two reasons. First, educational attainment plays a key role in
human mate search. People are attracted to educated potential partners
[@buss1986preferences; @belot2013dating]; spouse pairs often have
similar levels of educational attainment, as well as similar PSEA
[@vandenberg1972assortative; @schwartz2005trends; @greenwood2014marry;
@hugh2016assortative]. Second, PSEA predicts a set of important socioeconomic
variables, including not only education but also social and geographic mobility, 
IQ, future income and wealth [@belsky2016genetics; @barth2020genetic; @papageorge2020genes].[^discuss]

[^discuss]: See @papageorge2020genes for a detailed discussion of polygenic 
scores, aimed at economists.

```{r calc-psea-check}

mod_own_psea <- feols(university ~ EA3, data = famhist, note = FALSE)
mod_own_psea_within <- feols(university ~ EA3 | sib_group, data = famhist,
                               note = FALSE)

n_own_psea <- glance(mod_own_psea)[["nobs"]]
n_own_psea_within <- glance(mod_own_psea_within)[["nobs"]]

mod_own_psea <- tidy(mod_own_psea)
mod_own_psea_within <- tidy(mod_own_psea_within)

eff_own_psea  <- mod_own_psea[[2, "estimate"]]
pval_own_psea <- mod_own_psea[[2, "p.value"]]

eff_own_psea_within  <- mod_own_psea_within[[1, "estimate"]]
pval_own_psea_within <- mod_own_psea_within[[1, "p.value"]]

stopifnot(pval_own_psea < 2e-16)
stopifnot(pval_own_psea_within < 2e-16)
```

We calculate PSEA using per-SNP summary statistics from [@lee2018gene],
re-estimated excluding UK Biobank participants.[^5] We normalize the score to
have mean 0 and variance 1. Because polygenic scores are created from estimates
of many presumably tiny effects, they contain a large amount of noise relative
to the true best estimator that could be derived from genetic data. For
instance, PSEA explains only 11–13% of variance in educational attainment [out
of sample; @lee2018gene], whereas the true proportion explained by genetic
variation - the heritability - is estimated from twin studies to be about 40%
[@branigan2013variation]. In addition, polygenic scores are no more guaranteed
to be causal than any other independent variable. For example, social
stratification by descent may lead genes to be associated with educational
attainment even while playing no causal role [@selzam2019comparing].

[^5]: PSEA was computed by summing the alleles across \~1.3 million
    genetic variants weighted by their effect sizes as estimated in
    genome-wide association studies (GWASs) that excluded UK Biobank.
    PSEA was then residualized on the first 100 principal components of
    the SNP array data. Further details can be found in
    @abdellaoui2019genetic.

Despite these points, PSEA has non-trivial estimated effects on educational
attainment. PSEA correlates with measures of education, including university
attendance and years of full-time education; within-siblings regressions, where
PSEA is randomly assigned by the "lottery of meiosis", confirm this correlation
is at least partly causal [@lee2018gene]. We recheck these facts within the UK
Biobank sample. In a simple linear regression (N = `r pretty(n_own_psea)`) of
university attendance on PSEA, a one-standard-deviation increase in PSEA was
associated with a `r pretty(eff_own_psea*100)` percentage point increase in the
probability of university attendance ($p < 2 \times 10^{-16}$). In a
within-siblings regression among genetic full siblings (N = 
`r pretty(n_own_psea_within)`), the increase was 
`r pretty(eff_own_psea_within * 100)` ($p < 2 \times 10^{-16}$). This suggests 
that about half of the raw correlation of PSEA with university attendance is
down to confounds like good environments or parental nurture, while the
remainder is causal. Nevertheless, the causal effect remains substantial: for a
rough comparison, the (ITT) effect on college attendance of the Moving To
Opportunity experiment in the US was 2.5 percentage points [@chetty2016effects].

Figures \@ref(fig:pic-basic-corr-uni) and
\@ref(fig:pic-basic-corr-income) illustrate the core idea of
SGAM. The X axis shows a measure of one partner's socio-economic status:
university attendance (Figure \@ref(fig:pic-basic-corr-uni)) or income
(Figure \@ref(fig:pic-basic-corr-income)). The Y axis plots the other
partner's mean PSEA. Both males and females who went to university had
spouses with higher PSEA. So did males and females with higher income.
Since DNA is inherited, these people's children will also have higher
PSEA.

```{r pic-basic-corr-uni, fig.subcap = c("Female PSEA by male educational attainment", "Male PSEA by female educational attainment"), fig.cap = "Spouse PSEA against own university attendance. Lines show 95\\% confidence intervals", fig.width = 3, fig.height = 3}

pic_ss <- stat_summary(fun.data = mean_cl_normal, na.rm = TRUE)
pic_cc <- coord_cartesian(ylim = c(-0.15, 0.25))

mf_pairs %>% 
      filter(! is.na(university.m)) %>% 
      mutate(University = ifelse(university.m, "Yes", "No")) %>% 
      ggplot(aes(University, EA3.f)) + 
        pic_ss + 
        pic_cc +
        labs(x = "Male university attendance", y = "Female PSEA")

mf_pairs %>% 
      filter(! is.na(university.f)) %>% 
      mutate(University = ifelse(university.f, "Yes", "No")) %>% 

      ggplot(aes(University, EA3.m)) + 
        pic_ss + 
        pic_cc +
        labs(x = "Female university attendance", y = "Male PSEA")
        
```

```{r pic-basic-corr-income, fig.subcap = c("Female EA3 by male income", "Male EA3 by female income"), fig.cap = "Spouse PSEA against own estimated income in first job. Lines show 95\\% confidence intervals", fig.width = 3, fig.height = 3}

pic_ss <- stat_summary(
            fun.data = mean_se, 
            fun.args = list(mult = 1.96), 
            na.rm = TRUE
          )
pic_cc <- coord_cartesian(ylim = c(-0.15, 0.25))

mf_pairs %>% 
      filter(! is.na(first_job_pay.m)) %>% 
      mutate(Income = santoku::chop_deciles(first_job_pay.m, labels = 1:10)) %>% 
      ggplot(aes(Income, EA3.f)) + 
        pic_ss + 
        # pic_cc + 
        labs(x = "Male income decile", y = "Female PSEA")

mf_pairs %>% 
      filter(! is.na(first_job_pay.f)) %>% 
      mutate(Income = santoku::chop_deciles(first_job_pay.f, labels = 1:10)) %>% 
      ggplot(aes(Income, EA3.m)) + 
        pic_ss + 
        # pic_cc +
        labs(x = "Female income decile", y = "Male PSEA")
        
```

These figures do not prove that SGAM is taking place. Since an
individual's own PSEA correlates with both their educational attainment,
and their income, both figures could be a result of genetic assortative
mating (GAM) alone [@hugh2016assortative]. To demonstrate SGAM, we need
a source of social status which is exogenous to genetics. Also, the link
between social status and spouse genetics is likely to be noisy, for
three reasons: first, polygenic scores contain a large amount of error, as
discussed above; second, causal mechanisms behind variation in social status 
are likely to be noisy; third, to paraphrase @shakespeare1595midsummer, the 
spouse matching process is notoriously unpredictable. So, we need a large N to
give us sufficient power. This rules out time-limited shocks such as
changes to the school leaving age [@Davies_2018].

We use *birth order*. It is known that earlier-born children receive more
parental care and have better life outcomes, including measures of SES such as
educational attainment and occupational status [@Lindahl_2008; @booth2009birth;
@black2011older]. On the other hand, all full siblings have the same *ex ante*
expected genetic endowment from their parents, irrespective of their birth
order. For example, siblings' expected polygenic score is equal to the mean of
their parents' polygenic scores.[^6] We can therefore use birth order as a
"shock" to social status. Despite this language, we do not claim that birth
order is exogenous to all other variables. For example, it naturally correlates
with parental age, and it may also be related to the family's economic position at
the time of birth. We only claim that birth order is exogenous to genetic
variation.

[^6]: Although genetic variation is randomly assigned to children at
    birth, genetics and birth order could be dependent if parents'
    choice of whether to have more children is endogenous to the genetic
    endowment of their earlier children. We check for this below.

Our main independent variable is respondents' birth order, i.e. their
number of elder siblings plus one. For controls we use family size, i.e.
respondents' total number of siblings including themselves; month of birth; age
at interview; respondents' own PSEA; and their father's and/or mother's age
at their birth (calculated from parent's current age, only available if
the parent was still alive). For most regressions, we use only
respondents with between 1 and 5 siblings, i.e. with a family size of
2-6. 

To test whether birth order effects are mediated by SES, we use two measures:
income, and university attendance. Income is a direct measure of SES, while
university attendance is a predictor of income over the entire life course. The
UK Biobank data only contains a direct measure of current household income,
which is inappropriate for our purposes because it includes income from both
spouses. Instead, we estimate income in the respondent's first job, from the
job's Standard Occupational Classification (SOC) code. We match this with
information on average earnings by SOC from the Annual Survey of Hours and
Earnings 2007 [@ONS2007ASHE]. This data is only available for a subset of
respondents.

## Decomposing the birth order effect on spouse genetics

Ideally, we might prefer to use birth order as an instrument for SES. However,
our measures of social status are noisy and incomplete. For example, we know
whether subjects attended university, but not which university. Birth order
likely affects both measured and unmeasured aspects of SES. So, an instrumental
variables approach would be likely to fall foul of the exclusion restriction.

Instead, we conduct a mediation analysis, following the strategy of
@heckman2013understanding. We first confirm statistically that birth
order affects our measures of respondents' SES (income and education).
Then, we regress spouse's PSEA on birth order, with and without
controlling for SES. Under the assumption that birth order is exogenous
to own genetics, these regressions identify the effect of birth order,
plus other environmental variables that correlate with it, on own social
status and spouse's genetics. Also, if the estimated effect of birth
order on spouse's PSEA changes when SES is controlled for, that is
evidence that SES mediates the effect of birth order.


Linearizing our model so that $A(g, s) = (1-k)g + ks$ and applying
\@ref(eq:same-A) shows that:

```{=tex}
\[
\frac{d A(g_{p(i)}, s_{p(i)})}{d s_i} = k
\]
```

We wish to test whether $k \in (0, 1)$, i.e. whether SGAM is taking
place. If $k > 0$ then an increase in $i$'s social status $s_i$ will
increase $i$'s attractiveness $A$; if $k < 1$ then an increase in $A$
will be associated (in expectation) with an increase in $i$'s partner's
genetic endowment $g_{p(i)}$. We therefore wish to estimate the effect
of $i$'s status on their partner's genetics, while controlling for $i$'s
own genetics $g_i$. Since measures of genetic endowment, such as PSEA,
are noisy and incomplete, it is not enough to include them as controls in the
regression. Instead, we use birth order as a source of variation in
$s_i$ which is orthogonal to $g_i$.

We follow @heckman2013understanding to decompose the aggregate treatment
effect into components due to observed and unobserved proximate channels
affected by the treatment. Our aim is to estimate the effect of SES (as
an effect of birth order) on spouse PSEA.

Assume $B$ is a variable indicating birth order. The observed outcome
(spouse PSEA) is

```{=tex}
\begin{align}
\label{eq:Y}
Y = \sum_{b \in B}Y_b\delta_b
\end{align}
```
where $Y_1, Y_2$ etc. are counterfactual outcomes for the first-born,
second-born etc. and $\delta_b$ is a dummy variable which is 1 if birth
order equals $b$. Given $b$, spouse PSEA is assumed to be independent
across observations conditional on some predetermined controls which are
assumed to be unaffected by $b$.

Let $\theta_b$ be a set of mediators, i.e. proximate outcomes determined
by $b$, which account (at least in part) for the $b$ treatment effect on
spouse PSEA. We can think of $\theta_b$ as all the effects on
attractiveness, such as increments to SES, health, cognitive and
non-cognitive skills, that individuals receive due to their birth rank.

Our linear model is

```{=tex}
\begin{align}
\label{eq:Yb}
Y_b = \kappa_b + \alpha_b\theta_b + \beta_b \cdot \mathbf{X} + \tilde{\varepsilon}_b
\end{align}
```
where $\tilde{\varepsilon}_b$ is a mean-zero residual assumed
independent of $\theta_b$ and $\mathbf{X}$. We simplify by assuming that
$\beta_b = \beta$ and $\alpha_b = \alpha$ for all $b$, i.e. that the
effects of $\mathbf{X}$ and $\theta_b$ don't differ by birth order.

We can split the mediators in $\theta_b$ into a set $J_m$ of measured
mediators, including measures of SES, and a set $J_u$ of mediators that
we cannot measure. Write

```{=tex}
\begin{align}
\label{eq:Yb2}
Y_b = \tau_b + \sum_{j \in J_m} \alpha^j \theta^j_b + \beta \cdot \mathbf{X} + \varepsilon_b
\end{align}
```
where $\tau_b = \kappa_b + \sum_{j \in J_u} \alpha^j E(\theta^j_b)$ and
$\varepsilon_b = \tilde{\varepsilon}_b + \sum_{j \in J_u} (\theta^j_b - E(\theta^j_b))$.
We assume differences in unmeasured investments due to $b$ are
independent of $\mathbf{X}$.

With the assumptions above and substituting \@ref(eq:Yb2) into
\@ref(eq:Y) we obtain:

```{=tex}
\begin{align}
\label{eq:Ynew}
Y = \tau_1+\sum_{b \in B\backslash\{1\}} \delta_b(\tau_b-\tau_1) + \sum_{j \in J_m} \alpha^j \theta^j  +   \beta \cdot \mathbf{X} + \varepsilon
\end{align}
```
where $\theta^j = \sum_{b \in B}\theta^j_b\delta_b$ and
$\varepsilon = \sum_{b \in B}\varepsilon_b\delta_b$; $\tau_b-\tau_1$
measures the contribution of unmeasured variables to the average
treatment effect of birth order $b$, relative to birth order 1; and
$\theta^j, j \in J_m$ denote the observed mediators that we can measure.

Estimating the above by OLS will generate unbiased estimates of
$\alpha^j$ if $\theta^j$ is measured without error and is uncorrelated
with the error term $\varepsilon$. Since $\varepsilon$ contains both
individual disturbances and differences in unmeasured investments due to
birth order, there are two identifying assumptions that need to hold for unbiased
OLS estimates. First, the measured investments (specifically SES) should
be independent of unmeasured investments generated by birth order.
Failing this, the estimates of $\alpha^j$ will be conflated with the
effects of unmeasured investments. Second, the measured investments should
be uncorrelated with other shocks $\tilde{\varepsilon}_b$. The overall
treatment effect (relative to birth order $b = 1$) can then be
decomposed as follows:

```{=tex}
\begin{align}
\label{eq:YbY1}
E(Y_b - Y_1) = \tau_b - \tau_1 + \sum_{j \in J_m} \alpha^j E(\theta^j_b - \theta^j_1)
\end{align}
```
where $\tau_b - \tau_1$ is the unmeasured component of the treatment
effect, and $\sum_{j \in J_m} \alpha^j E(\theta^j_b - \theta^j_1)$ is
the treatment effect due to measured investments.

By running a least square regression of \@ref(eq:YbY1), we can estimate
$\tau_b - \tau_1$ . If assumption (a) above holds, the part of the birth
order treatment effect on spouse PSEA that is due to measured mediators,
including SES, can be constructed using the estimated $\alpha^j$ and the
effects of birth order on measured investments. Specifically, the part
of the birth order effect that is due to university education (or
income) on spouse PSEA will be the coefficient of university (income) in
the regression of spouse PSEA in equation \@ref(eq:Ynew), multiplied by
the coefficient of birth order from a model where the dependent variable
is an indicator for university (income).

# Results


```{r tbl-bo-first-stage}
fml_bo_1st <- list()
fml_bo_1st[["University"]] <- update(fml_bo_psea[[1]], university.x ~ . )
fml_bo_1st[["Income"]] <- update(fml_bo_psea[[1]], first_job_pay.x ~ .)
fml_bo_1st[["Fluid IQ"]] <- update(fml_bo_psea[[1]], fluid_iq.x ~ . )
fml_bo_1st[["Height"]] <- update(fml_bo_psea[[1]], height.x ~ .)
mod_bo_1st <- purrr::map(fml_bo_1st, fixest::feols, data = mf_pairs_reg, 
                           notes = FALSE)

coef_bo <- purrr::map(mod_bo_1st, tidy) %>%
             purrr::map(filter, term == "birth_order.x") %>% 
             purrr::map_dbl("estimate")

huxreg(mod_bo_1st, 
         coefs = c(
           "Birth order" = "birth_order.x", 
           "PSEA" = "EA3.x",
           "Parents' age at birth" = "par_age_birth.x"
         ),
         statistics = c(
           "N"  = "nobs", 
           "R2" = "r.squared"
          ),
         note = my_note,
         stars = my_stars,
         tidy_args = tidy_args
       ) %>% 
       insert_row(after = 7, "Family size dummies", rep("Yes", 4)) %>% 
       insert_row(after = 8, "Birth month dummies", rep("Yes", 4)) %>% 
       insert_row(after = 9, "Birth year dummies", rep("Yes", 4)) %>% 
       set_bottom_border(8:9, everywhere, 0) %>% 
       set_align(8:10, -1, "center") %>% 
       set_number_format(2:7, -1, 4) %>% 
       set_width(0.8) %>% 
       set_caption("Regressions of variables on birth order")
       
```


We first regress our measures of socio-economic status, university attendance
and income from first job, on birth order in our sample. We also do the same for
two non-SES mediators that could be affected by birth order: fluid IQ and height.
We control for respondent's own PSEA and their parents' age at birth (see below).
Table \@ref(tab:tbl-bo-first-stage) shows that birth order significantly
predicts all these variables. Effect sizes are quite substantial: a single extra
elder sibling reduces the chance of attending university by about 
`r pretty(-coef_bo["University"] * 100)` percentage points, income by about 
£`r pretty(-coef_bo["Income"] * 1000, format = "d")`, fluid IQ by about 
`r pretty(-coef_bo["Fluid IQ"])` points on a 13 point test, and height by about 
`r pretty(-coef_bo["Height"])` centimeters.

Next we run regressions of spouse PSEA on birth order, within our
dataset of spouse pairs. Table \@ref(tab:tbl-bo-psea-basic) reports the
results. Column 1 controls only for family size (using dummies). As
expected, higher birth order is negatively associated with spouse's
PSEA, though the estimated effect size is small and insignificant. 
Column 2 includes the respondent's own PSEA, as well as dummies for birth year 
to control for cohort effects, and dummies for birth month to control for 
seasonality effects. The effect size of birth order is not much changed.

Column 3 includes parents' age at birth. Within a family, later children
have older parents by definition. Older parents have more life
experience and may have higher income, which would presumably help later
children.[^7] Including parents' age means we can separate the effect
of parental age from birth order. This reduces the N by a lot, since
only respondents with a live parent reported the necessary data. However,
the effect of birth order jumps in size and becomes significant at the 5 per
cent level. Meanwhile, parents' age has a
positive effect. This suggests that estimates in columns 1-2 mixed two
opposite-signed effects: having older parents versus being later in
birth order.

[^7]: We often only have data only for one parent. We use this, or take the mean
  if we have both. There are also potential 
  genetic effects from parental age, though recent research has rejected these 
  in favour of "social" explanations [@kristensen2007explaining; @black2011older].
  @cochran2013paternal report that mutational load is approximately
  linear in father's age, while it is constant in mother's age. We
  observe very similar results if we control only for father's age at
  respondent's birth.


```{r tbl-bo-psea-basic}

fml_bo_psea_base <- list()

fml_bo_psea_base[[1]] <- EA3.y ~ birth_order.x | factor(n_sibs.x)
fml_bo_psea_base[[2]] <- EA3.y ~ birth_order.x + EA3.x + factor(birth_mon.x) |
                            factor(n_sibs.x) + factor(YOB.x)
  
fml_bo_psea_base[[3]] <- EA3.y ~ birth_order.x + EA3.x + factor(birth_mon.x) +
                            par_age_birth.x | factor(n_sibs.x) + factor(YOB.x)

fml_bo_psea_base <- lapply(fml_bo_psea_base, Formula::as.Formula)


mod_bo_psea_base <- lapply(fml_bo_psea_base, fixest::feols, 
                 data = mf_pairs_reg,
                 notes = FALSE
               )

huxreg(mod_bo_psea_base, 
         coefs = c(
           "Birth order"           = "birth_order.x", 
           "Own PSEA"              = "EA3.x",
           "Parents' age at birth" = "par_age_birth.x"
         ),
         statistics = c(
           "N"  = "nobs", 
           "R2" = "r.squared"
          ),
         note = my_note,
         stars = my_stars,
         tidy_args = tidy_args
       ) %>% 
       insert_row(after = 7, "Family size dummies", rep("Yes", 3)) %>% 
       insert_row(after = 8, "Birth month dummies", "No", "Yes", "Yes") %>% 
       insert_row(after = 9, "Birth year dummies", "No", "Yes", "Yes") %>% 
       set_bottom_border(8:9, everywhere, 0) %>% 
       set_align(8:10, -1, "center") %>% 
       set_number_format(2:7, -1, 4) %>% 
       set_width(0.8) %>% 
       set_caption("Regressions of spouse PSEA on birth order")


```

Having tested that birth order affects spouse's PSEA, we now look for
potential mediators of this effect. Despite the lower N, we continue to
control for respondents' parents' age, since this removes a confound
which would bias our results towards zero.[^9]

[^9]: The appendix reports results without controlling for parents' age.

```{r tbl-bo-psea}


mod_bo_psea <- lapply(fml_bo_psea, fixest::feols, 
                 data = mf_pairs_reg,
                 notes = FALSE
               )

huxreg(mod_bo_psea, 
        coefs = c(
          "Birth order"           = "birth_order.x", 
          "University"            = "university.xTRUE", 
          "Income"                = "first_job_pay.x",
          "Own PSEA"              = "EA3.x",
          "Parents' age at birth" = "par_age_birth.x",
          "Fluid IQ"              = "fluid_iq.x",
          "Height"                = "height.x"
        ),
         note = my_note,
         stars = my_stars,
         tidy_args = tidy_args
       ) %>% 
       insert_row(after = 15, "Family size dummies", rep("Yes", 4)) %>% 
       insert_row(after = 16, "Birth month dummies", rep("Yes", 4)) %>% 
       insert_row(after = 17, "Birth year dummies", rep("Yes", 4)) %>% 
       set_bottom_border(16:17, everywhere, 0) %>% 
       set_align(16:18, -1, "center") %>% 
       set_number_format(2:15, -1, 4) %>% 
       set_tb_padding(3) %>% 
       set_caption("Regressions of spouse PSEA on birth order and potential mediators")
```

Table \@ref(tab:tbl-bo-psea) shows the results. Column 1 shows the
effect of birth order, using the same specification as column 3 of the
previous table. The remaining columns add potential mediators of
birth order effects. Column 2 includes our first measure of
socio-economic status: university attendance. As controls, we also
include fluid IQ and height, both of which could be affected by birth
order and affect spouse matching. Column 3 adds our second measure of
socio-economic status, income in first job. Column 4 includes both.

When we add university attendance (column 2), the effect of birth order
drops and becomes insignificant, while the coefficient for university is
positive and highly significant. Fluid IQ and height are also positive
and significant. Controlling for income alone (column 3), again the effect of
birth order shrinks and becomes insignificant, while income has a positive and
highly significant effect. Lastly, the same pattern holds when we 
control for both university and income (column 4).

```{r tbl-mediation-prop, cache = TRUE}

# percent of main effect that is mediated by uni:
# uni effect of BO (first stage) * PSEA effect of uni (column 2)
# divided by PSEA effect of BO (column 1)

mediators <- c("university.xTRUE", "first_job_pay.x", "height.x",
                 "fluid_iq.x")

compute_mediation_effects <- function (resample = FALSE) {
  mf_pairs_tmp <- mf_pairs_reg
  if (resample) {
    mf_pairs_tmp <- mf_pairs_tmp[sample(nrow(mf_pairs_tmp), replace = TRUE),]
  }
  
  mod_bo_1st_boot <- purrr::map(fml_bo_1st, fixest::feols, 
                                  data = mf_pairs_tmp, notes = FALSE)
  mod_bo_psea_boot <- purrr::map(fml_bo_psea, fixest::feols, 
                                   data = mf_pairs_tmp, notes = FALSE)
  mod_bo_overall_boot <- fixest::feols(fml_bo_psea[[1]], 
                                         data = mf_pairs_tmp, 
                                         notes = FALSE
                                       )
                                  

  # grab out the birth_order.x coefficient:
  mediator_effect_bo <- purrr::map_dbl(mod_bo_1st_boot, 
                                         list(coef, "birth_order.x"))

  psea_effect_mediators <- purrr::map(mod_bo_psea_boot[-1],
                                         ~coef(.x)[mediators])
  psea_effect_mediators <- matrix(unlist(psea_effect_mediators), ncol = 4,
                                  byrow = TRUE)
  colnames(psea_effect_mediators) <- mediators
  
  psea_effect_bo <- coef(mod_bo_overall_boot)["birth_order.x"]

  effects_uni <- mediator_effect_bo["University"] * 
    psea_effect_mediators[, "university.xTRUE"] 
  effects_income <- mediator_effect_bo["Income"] * 
    psea_effect_mediators[, "first_job_pay.x"]
  effects_height <- mediator_effect_bo["Height"] * 
    psea_effect_mediators[, "height.x"] 
  effects_fluid_iq <- mediator_effect_bo["Fluid IQ"] * 
    psea_effect_mediators[, "fluid_iq.x"]
  
  effects_pct <- cbind(
          University = effects_uni, 
          Income     = effects_income, 
          Height     = effects_height, 
          `Fluid IQ` = effects_fluid_iq
        )
  # divide by overall effect re-estimated using data in each of models 2-4
  effects_pct <- effects_pct / psea_effect_bo
  
  effects_pct
}


# for each 1st stage model, run 199 bootstraps and gather coefs
n_reps <- 199

# bootstraps is now 3 x 4 x n_reps
# cis <- apply(bootstraps, 1:2, quantile, c(0.05, 0.95), na.rm = TRUE)
estimates <- compute_mediation_effects(resample = FALSE)

format_est_ci <- function(coef, model) {
  # char <- glue::glue("{percent(estimates[model, coef], 0.1)} ({percent(cis[1, model, coef], 0.1)}, {percent(cis[2,model, coef], 0.1)})")
  char <- percent(estimates[model, coef], 0.1)
  char[is.na(estimates[model, coef])] <- "-"
  char
}

med_matrix <- format_est_ci(c("University", "Income", "Fluid IQ", "Height"), 
                              1:3)
med_matrix <- matrix(med_matrix, ncol = 3, byrow = TRUE)
rownames(med_matrix) <- c("University", "Income", "Fluid IQ", "Height")
colnames(med_matrix) <- paste("Model", 2:4)
as_huxtable(med_matrix, add_colnames = TRUE, add_rownames = "") %>% 
      theme_article() %>% 
      set_caption("Percent of birth order effects accounted for by mediators, models 2-4")  
      # add_footnote("Bootstrap 90% confidence intervals in brackets") 

```

Under the assumptions discussed above, we can estimate the proportion of
the birth order effect that is mediated by these variables. Table
\@ref(tab:tbl-mediation-prop) reports this for each model in columns
2-4. Each estimate is the coefficient of birth order on the mediator,
times the coefficient of the mediator on spouse PSEA, divided by the
coefficient of birth order on spouse PSEA estimated from column 1, i.e.
without mediators. 

Our next regressions split up the data into subsets. Cultural
stereotypes often assume that the link between status and genes is not
symmetric across the genders, for example, that males with high SES are
particularly likely to marry attractive spouses. To test this, we
separately regress male spouses' PSEA on female birth order, and female
spouses' PSEA on female birth order. We also rerun regressions among the
subset of individuals who had children. A significant result here will
confirm that the association between status and genetics is carried over
into the next generation.

Table \@ref(tab:tbl-bo-subsets) shows the results. Columns 1 and 2 use birth
order of male respondents to predict female spouses' PSEA. Column 1 runs the
regression of birth order; column 2 adds university attendance as a mediator
(here, we focus on university attendance alone so as to keep our N large).
Columns 3 and 4 repeat the exercise for female respondents, using their birth
order to predict male spouses' PSEA. Significance drops below 5% in these subsets.
However, the pattern of coefficient sizes is the same as in the main regression:
university attendance is highly significant, and adding it reduces the absolute
size of the birth order effect.  Columns 5 and 6 use only couples with children. 
Here, birth order is significant in the base specification, and again, 
university attendance still seems to mediate the birth order effect.

Spouse matching is strictly a single process in which neither side is an
"independent variable". In Appendix Table \@ref(tab:tbl-sur), we estimate a
matching model in the style of @chiappori2012fatter, estimating the effects of 
birth order (and PSEA) on both spouse's PSEA and spouse's birth order, and
testing for proportionality of coefficients across the dependent variables. 
For women only we reject proportionality of coefficients, suggesting that there
may be some assortative matching separately on each dimension.

```{r tbl-bo-subsets}

mod_bo_males <- lapply(fml_bo_psea[1:2], fixest::feols, 
                 data = mf_pairs_reg %>% filter(x == "Male"),
                 notes = FALSE
               )
mod_bo_females <- lapply(fml_bo_psea[1:2], fixest::feols, 
                 data = mf_pairs_reg %>% filter(x == "Female"),
                 notes = FALSE
               )

mf_pairs_children <- mf_pairs_reg %>% 
                   filter(n_children.x > 0, n_children.y > 0)
mod_bo_children <- lapply(fml_bo_psea[1:2], fixest::feols, 
                 data = mf_pairs_children,
                 notes = FALSE
               )
# 
# ta_m <-  list(conf.int = FALSE, cluster = 
#                 list(mf_pairs_reg$f.54.0.0.x[mf_pairs_reg$x =="Male"]))
# ta_f <-  list(conf.int = FALSE, cluster = 
#                 list(mf_pairs_reg$f.54.0.0.x[mf_pairs_reg$x == "Female"]))
# ta_ch <-  list(conf.int = FALSE, cluster = list(mf_pairs_children$f.54.0.0.x))

reg_list <- list(
        "Male respondents"     = mod_bo_males[[1]],
        "Male respondents"     = mod_bo_males[[2]],
        "Female respondents"   = mod_bo_females[[1]],
        "Female respondents"   = mod_bo_females[[2]],
        "With children"        = mod_bo_children[[1]],
        "With children"        = mod_bo_children[[2]]
      )

huxreg(reg_list,
         coefs = c(
            "Birth order"           = "birth_order.x", 
            "University"            = "university.xTRUE",
            "Own PSEA"              = "EA3.x",
            "Parents' age at birth" = "par_age_birth.x",
            "Fluid IQ"              = "fluid_iq.x",
            "Height"                = "height.x"
          ),
         statistics = c("N" = "nobs", "R2" = "r.squared"),
         note = my_note,
         stars = my_stars,
         tidy_args = tidy_args
       ) %>% 
       insert_row(after = 13, "Family size dummies", rep("Yes", 6)) %>% 
       insert_row(after = 14, "Birth month dummies", rep("Yes", 6)) %>% 
       insert_row(after = 15, "Birth year dummies", rep("Yes", 6)) %>% 
       set_bottom_border(14:15, everywhere, 0) %>% 
       set_align(14:16, -1, "center") %>% 
       set_font_size(10) %>% 
       set_width(1) %>% 
       set_tb_padding(3) %>% 
       set_lr_padding(3) %>% 
       set_caption("Regressions of spouse PSEA on birth order: subsets")

```

\FloatBarrier

## Robustness

Although all children of the same parents have the same polygenic scores
in expectation, it could still be possible that genetics correlates with
birth order within the sample. This could happen in three ways. First,
siblings with high birth order will typically come from larger families
than those with low birth order, and parents of different-sized families
are likely to differ systematically on many dimensions, including
genetics. We controlled for this by including a full set of family size
dummies in the regression. Second, there could be selection bias. For
example, if later siblings with high PSEA, and earlier siblings with low
PSEA, are more likely to enter the sample, then this would bias our
results. Thirdly, parents might select family size on the basis of
genetics. For example, suppose that if the first child has a phenotype
reflecting a high PSEA, parents are more likely to have a second child.
Then within the subset of two-child families, first children would have
higher-than-average PSEA, while second children would not.

```{r tbl-pgs-check}

check_cor <- function (score_name) {
  fml_check <- as.formula(paste0(score_name, ".x ~ factor(n_sibs.x) + birth_order.x"))
  mod_check <- lm(fml_check, mf_pairs_reg)
  mod_check %>% 
        tidy() %>% 
        filter(term == "birth_order.x") %>% 
        mutate(score = score_name)
}

check_cor_within <- function (score_name) {
  fml_check_within <- as.formula(paste0(score_name, 
                        ".x ~ 0 + factor(n_sibs.x) + factor(n_sibs.x):factor(birth_order.x)"))
  mod_check <- lm(fml_check_within, mf_pairs_reg)
  # NB: the model creates NA estimates for when n_sibs <= birth_order
  # since these have 0 rows in the data. I think this is not a problem.
  mod_check %>% 
        tidy() %>% 
        filter(grepl("birth_order", term)) %>% 
        filter(! is.na(estimate)) %>% 
        mutate(score = score_name)
}

drake::loadd(score_names)
score_names <- paste0(score_names, "_resid")
balance_check <- purrr::map_dfr(score_names, check_cor)
balance_check_within <- purrr::map_dfr(score_names, check_cor_within)

if (any(balance_check$p.value < 0.1/33)) stop("Some check p values < 0.1/33; rewrite text!")

n_below_10 <- sum(balance_check$p.value < 0.1)
stopifnot(n_below_10 == 5)
if(any(
        balance_check %>% filter(p.value < 0.10) %>% pull(estimate) > 0.02
      )) stop("Some coefs > 0.02; rewrite text!")

n_within_coefs <- nrow(balance_check_within)
if (any(balance_check_within$p.value < 0.001)) stop("Some within-siblings p values < 0.001; rewrite text!")


max_coef <- max(abs(balance_check$estimate))

mod_bo_own_psea <- lm(EA3.x ~ birth_order.x, 
                      mf_pairs_reg,
                      subset = n_sibs.x==3)
coef_own_psea_3 <- tidy(mod_bo_own_psea)[[2, "estimate"]]
pval_own_psea_3 <- tidy(mod_bo_own_psea)[[2, "p.value"]]
stopifnot(pval_own_psea_3 < 0.05)

```

To check for the latter two problems, we run balance tests on 33 different
polygenic scores.[^10] We regress each score on own birth order, controlling for
family size. No scores were significant at $p < 0.10/33$. Five scores were
significant at $p < 0.10$, all with effect sizes of less than 0.02 per standard
deviation. Table \@ref(tab:tbl-bo-psea-pgs) in the appendix reruns our
regressions controlling for these scores. Results are almost unchanged. To test
whether polygenic scores might vary across birth orders within a particular
family size, we also regress each score on a full set of birth order dummies,
interacted with a full set of family size dummies. None of the 
`r n_within_coefs` birth order coefficients were significant at $p < 0.001$.
However, among families of size 3, there is a significant positive correlation
of birth order with own PSEA (effect size `r pretty(coef_own_psea_3, 3)`, 
$p = `r pretty(pval_own_psea_3, 2)`$). Table \@ref(tab:tbl-bo-psea-no3) in the
appendix therefore reruns our regressions with families of size 3 excluded.
Results are substantially unchanged. Of course, there could still be 
unmeasured genetic variants which correlate with birth order in our sample.
Nevertheless, a wide set of polygenic scores shows no large or significant
correlation. This makes us more confident that birth order is indeed exogenous
to genetics.

[^10]: Polygenic scores were residualized on the first 100 principal
    components of the genetic data. Scores were for: ADHD, age at
    menarche, age at menopause, agreeableness, age at smoking
    initiation, alcohol use, Alzheimer's, autism, bipolarity, BMI, body
    fat, caffeine consumption, cannabis (ever vs. never), cognitive
    ability, conscientiousness, coronary artery disease, smoking
    (cigarettes per day), type II diabetes, drinks per week, educational
    attainment (EA2 and EA3), anorexia, extraversion, height, hip
    circumference, major depressive disorder, neuroticism, openness,
    smoking cessation, schizophrenia, smoking initiation, waist
    circumference, and waist-to-hip ratio.

The appendix reports other robustness checks, including a specification with
separate dummies for each value of birth order, and replacing university attendance
with age of leaving full-time education.

# Conclusion

Our empirical analysis shows that in a contemporary developed society,
earlier-born children had spouses with higher PSEA. We also provide
evidence that these effects are mediated by socio-economic status, specifically
income and education.

There are multiple mechanisms by which advantage is transmitted across
generations. High-income parents may invest more in their children's human
capital, transfer wealth via gifts and bequests, model valuable skills, or
provide them with advantageous social networks. They may also pass on causally
relevant genetic variants. This channel has been proposed as a reason for the
surprising persistence of inequality over generations
[@clark2015intergenerational; @clark2021bell]. One problem with this theory is
that in the absence of assortative mating, genetic variation regresses swiftly
to the mean, with coefficient $r = 0.5$ per generation. Thus to explain long-run
persistence, the genetic theory seems to require very high levels of genetic
assortative mating. SGAM may help to solve this puzzle. Persistence will be
increased if, in addition to genetic assortative mating, high SES itself
attracts "good genes". At the same time, SGAM changes the interpretation of 
genetics. Genetic variation is not an exogenous input into the social system, 
but an endogenous outcome of the structure of the marriage market, as our 
theoretical model shows. 

SGAM also provides a new explanation for the observed association of genes with
SES. Unlike meritocratic social mobility, the leading alternative explanation,
SGAM may apply to a historically wide range of societies. Whilst a
degree of meritocracy exists in modern capitalist economies, opportunities to
earn status have been far more limited in most societies throughout history
[@smelser1966social]. On the other hand, assortative mating is likely to be a
cultural universal [@buss1989sex]. Thus, SGAM predicts that genetic differences
across social status should exist in all stratified societies. In fact, people
in many societies have believed that innate traits do vary by social
status. The ancient Greeks described the social elite as *καλοί κἀγαθοί* ("fine
and good"), while the Roman nobility were the *optimates* ("best").[^11] This
belief has been explained by the tendency to believe in a just world
[@furnham1993just], or as an ideology promoted by the dominant class [e.g.
@gramsci1971selections]. However, it may also simply have been a recognition of
(social) reality. In other words, the belief that elites are taller, stronger,
better-looking, etc. is not much different from the belief that elites are
richer and more powerful, and may be held for similar reasons. In future, it
may be possible to directly test for genetic differences across social status
in ancient DNA samples.

[^11]: The appendix contains a selection of relevant historical
    quotations.


Our empirical analysis has focused on significance rather than effect size. This
is by necessity: even the latest estimates of PSEA contain substantial noise
compared to the "true" polygenic score. As polygenic scores become more accurate,
we will be better able to gauge the true effect size of SGAM on both the
inheritance of inequality and the genes-SES gradient. We expect this to be
an increasingly important research topic.

In our model, the association between social status and genetic variation
depends on the society's marriage market. The association is weaker when
marriage markets are very socially egalitarian, with spouse matching driven only
by genetics, or very inegalitarian, with matching driven only by status. Thus,
there is a non-monotonic relationship between "meritocracy" in marriage markets,
and the gene-environment correlation ("rGE") between genes and SES. This logic
is different from the standard mechanism, in which more meritocratic societies
give genetic variants a greater role in causing labour market success, so that
the relationship between social meritocracy and rGE is monotonic.  Comparing rGE
across societies or over time is beyond our scope here, but we see it as a good
area for future theory and empirics. Marriage markets may play an important role
in explaining how inequality and intergenerational mobility vary across
different societies.

Marriage markets are likely to respond to other features of the economy. For
instance, a society's level of redistribution could affect the relative weight
of social status in the attractiveness function. If so, then part of the effect
of large-scale social changes may take place over the very long run, by altering
marriage markets and thus affecting the genes-SES gradient and intergenerational
mobility. Intuitively one would expect that a more "egalitarian" society and
marriage market would lead to a weaker genes-SES gradient. But in our model, the
covariance between genes $G$ and status $S$ is non-monotonic in the weight of
status in marriage markets $k$, so this intuition may not always hold true. A
natural extension of our theory would be to endogenize the marriage market to
investigate further. In any case, the potential for economic changes to have
such long-run effects suggests the potential for hysteresis, with initial social
differences cumulating over time via their effect on genetic inequality.

The broadest message of this paper is that *genetics are a social
outcome*. Both popular and scientific discourse often parse genetics as
"nature", in opposition to "nurture" or "environment" [e.g.
@chakravarti2003nature; @plomin2019blueprint]. This idea expresses the
fact that our genetic endowment is fixed at birth and cannot be
influenced by our social environment (though genes *always* interact with the
environment to cause individual outcomes). But the idea that human
genetics are natural can be highly misleading. Humans inherit their
genes from their parents, along with other forms of inheritance such as
economic and cultural capital. Human parents, in turn, form spouse
pairs, bear children, and raise them, within social institutions. A
person's genetic inheritance is a social and historical fact about them,
and not a fact of nature, any more than their inherited wealth or social
status is natural. As [@marx1844economic] wrote, "History is the true
natural history of man". Genetic endowments can even be viewed as
another form of capital, alongside human, social and cultural capital: a
resource to be sought, accumulated and competed over. The analysis of
this kind of capital is an exciting area for further research, which
will require the contributions of both social scientists and
geneticists.

\FloatBarrier

\newpage

# Appendix

## Proof of Proposition \@ref(prop:prop1)

```{=tex}
\begin{proof}


Within each pair $i,p(i)$ write $d$ for the person with $s_{d}>s_{p(d)}$
and $m$ for $p(d)$.  If
$k < 1$, then $g_{d} < g_{m}$. (If $k = 0$, then define $d$ as the person
with $g_{d} < g_{p(d)}$.) We integrate over the pairs to calculate the 
covariance in the parents' generation:
\[
cov(G_{p},S_{p})=\int\frac{1}{2}(g_{d}s_{d}+g_{p(d)}s_{p(d)})\,\mathrm{d}d.
\]

For the children, the equivalent expression is
\[
cov(G_{c},S_{c})=\int g_{c(d)}s_{c(d)}\mathrm{d}d,
\]

observing that $EG_{c} = ES_{c}=0$ from (\ref{eq:children-average}).

Take an arbitrary pair $d,m$. Write
\begin{align*}
g_{d}s_{d} & = (g_{c}-\Delta g)(s_{c}+\Delta s);\\
g_{m}s_{m} & = (g_{c}+\Delta g)(s_{c}-\Delta s)
\end{align*}
where 
\begin{align*}
\text{\ensuremath{\Delta g=\frac{g_{m}-g_{d}}{2}\ge0}}, & \textrm{strictly so if and only if }k>0;\\
\Delta s=\frac{s_{d}-s_{m}}{2}\ge0 & \textrm{strictly so if and only if }k<1.
\end{align*}
Taking the average of the parents gives
\[
\frac{1}{2}(g_{d}s_{d}+g_{m}s_{m})= g _{c}s_{c}-\Delta g\Delta s.
\]
This is less than $g_{c}s_{c}$ if $0 < k < 1$, and equal to it if $k = 0$
or $k = 1$. Plugging this into the integral shows that 
\[
cov(G_{p},S_{p})\le cov(G_{c},S_{c})
\]
again with strict inequality if and only if $0 < k < 1$. This proves the first
part. 

To prove the second part, write
\begin{equation}
corr(G_{j},S_{j})=\frac{cov(G_{j},S_{j})}{\sqrt{var(G_{j})var(S_{j})}}\textrm{ for both generations }j\in\{p,c\}.\label{eq:corr-cov-var}
\end{equation}
where 
\begin{align*}
var(G_{p}) & =\frac{1}{2}\int g_{d}^{2}+g_{p(d)}^{2}\textrm{d}d;\\
var(G_{c}) & =\int g_{c(d)}^{2}\textrm{d}d.
\end{align*}

Much as before,
\begin{align*}
g_{d}^{2}+g_{m}^{2} & =(g_{c}-\Delta g)^{2}+(g_{c}+\Delta g)^{2}\\
 & =2g_{c}^{2}+2(\Delta g)^{2}\\
 & \ge2g_{c}^{2}.
\end{align*}

This shows that $var(G_{c})\le var(G_{p})$ and a similar argument
shows $var(S_{c})\le var(S_{p})$. Thus the covariance is higher (and
positive) in the children's generation, while the variances are lower.
Combining these ensures that 
\[
corr(G_{c},S_{c}) \ge corr(G_{p},S_{p}).
\]

Since for any $k$, either $var(G_{c}) < var(G_{p})$ or $var(S_{c}) < var(S_{p})$,
the only way to get strict equality for the above is if $k \in \{0,1\}$
and $cov(G_{c}, S_{c}) = cov(G_{p},S_{p}) = 0$.

\end{proof}
```
To show that the condition in the second part cannot be relaxed further,
consider the distribution in Figure
\ref{fig:Correlation-counterexample}. There is negative correlation in
the parents' generation (the shaded area). If $k = 1$ or is close enough
to 1, then assortative mating along the dotted lines will reduce the
variance of $S$ along those lines, pushing the distribution towards the
darker central area, without affecting the covariance. This will make
the correlation more negative. After repeated generations the horizontal
variance within values of $G$ will almost disappear and the correlation
will approach -1.

```{=tex}
\begin{figure}
\begin{centering}
\includegraphics[width=0.5\columnwidth]{correlation-counterexample}
\par\end{centering}
\caption{Correlation counterexample\label{fig:Correlation-counterexample}}

\end{figure}
```
## Proof of Proposition \ref{prop:robustness}

```{=tex}
\begin{proof}

Note that in proposition \ref{prop:prop1}, we took $g_{c(i)} = \bar{g}_{i}$ and $s_{c(i)} = \bar{s}_i$. Write 
\begin{align}
cov(G_{c},S_{c}) & =cov(\bar{G}+\varepsilon^{G},\bar{S}+\varepsilon^{S})\nonumber \\
 & = cov(\bar{G},\bar{S}) + cov(\varepsilon^{G},\bar{S}) + 
     cov(\bar{G},\varepsilon^{S}) + cov(\varepsilon^{G},\varepsilon^{S}).
 \label{eq:cov-with-error}
\end{align}
For any $X$ and $Y$, $cov(X, Y)$ is bounded by $\sqrt{var(X) var(Y)}$.
Plugging $\sigma_{G}^{2}$ and $\sigma_{S}^{2}$ into this formula
shows that under condition 1, $cov(G_{c},S_{c})$ will be arbitrarily
close to $cov(\bar{G},\bar{S})$. Similarly, writing
\[
var(G_{c}) = var(\bar{G}) + var(\varepsilon^{G}) + 2cov(\bar{G},\varepsilon^{G})
\]
shows that $var(G_{c})$ will approach $var(\bar{G})$ as $\sigma_{G}^{2}$
grows small, and similarly for $var(S_{c})$. Plugging these facts
into (\ref{eq:corr-cov-var}) shows that $corr(G_{c},S_{c})$ approaches
$corr(\bar{G},\bar{S})$ as $\sigma_{G}^{2}$ and $\sigma_{S}^{2}$
grow small. Proposition 1 then shows $corr(\bar{G},\bar{S}) < corr(G_{p},S_{p})$
for $k\in(0,1)$.

Under condition 2, $cov(G_{c}, S_{c}) = cov(\bar{G},\bar{S})$ since
the last three terms of the sum in (\ref{eq:cov-with-error}) are
zero. Then since 
\[
cov(\bar{G},\bar{S})\ge cov(G_{p},S_{p}) = 0
\]
with strict inequality iff $k\in(0,1)$, the covariance signs the
correlation. 

\end{proof}
```
\FloatBarrier

\newpage

## Robustness checks

Table \@ref(tab:tbl-bo-psea-no-par-age) reruns our central regressions,
dropping the control for parents' age at birth. Results show the same
pattern as in the main text: the coefficient for birth order is
negative, but changes sign when university attendance is added as a
potential mediator. However, the birth order effect is smaller overall,
and is never significant. We also ran regressions using father's age
only; results are similar to those in the main text.

Table \@ref(tab:tbl-bo-psea-dummies) reruns our central regressions but
includes a separate coefficient for each position in the birth order
(with firstborn as the baseline). The basic pattern of our main result
holds: birth order coefficients are generally negative; adding mediators
causes them to increase towards zero or to change sign. Birth order
effects appear largest for birth order 2-3. However, effects for later
birth orders are also imprecisely estimated (since fewer respondents
come from large families).

Notably when we add income, dummies for birth order 5 and 6 become large
and positive. This could be (for instance) because being the last born
has advantages after effects on SES have been netted out. Table
\@ref(tab:tbl-bo-subset-dummies) runs the same exercise for different
subsets: male respondents, female respondents, and couples with
children. The basic pattern that birth order coefficients shrink after
adding mediators is quite robust. Note however that here, the estimates
of effects for birth order 2-3 are larger for females.

We also ran a specification with separate birth order dummies within
each family size. Figure \@ref(fig:pic-bo-psea-interactions) shows 95%
confidence intervals for the birth order coefficients, from the column 2
specification including height and IQ controls but no mediators. Not
surprisingly, coefficients are imprecisely estimated. But most birth
order coefficients are negative compared to the baseline for firstborns.

Table \@ref(tab:tbl-bo-psea-pgs) reruns our regressions controlling for
several polygenic scores. Results are very close to those in the main
text.

Table \@ref(tab:tbl-bo-psea-age-fte) reruns relevant columns of Table
\@ref(tab:tbl-bo-psea) using age of leaving full-time education as a
measure of educational SES, instead of the university attendance dummy.
Results are similar to those in the main text: controlling for age of
leaving full-time education shrinks the effect of birth order and makes
it insignificant.

Table \@ref(tab:tbl-bo-psea-no3) reruns Table \@ref(tab:tbl-bo-psea)
excluding families of size 3. Results are very similar to those in the
main text.


```{r calc-sur-ratios}

# Plan
# - for men and women separately
# - estimate table 2, last column with DVs of (a) spouse PSEA and (b)
#   spouse birth order
# - check that the coefficients have the same ratios, as in "Fatter Attrn"
#   paper

# we avoid YOB fixed effects as it makes the estimation fail
fml_sur_psea <- EA3.y ~ birth_order.x + EA3.x + par_age_birth.x + 
                  factor(n_sibs.x) + factor(birth_mon.x)
fml_sur_bo <- update(fml_sur_psea, birth_order.y ~ .)

list_fml <- list(PSEA = fml_sur_psea, BO = fml_sur_bo)

# systemfit doesn't like missing data
mf_pairs_sf <- mf_pairs_reg %>% 
                 filter(
                   ! is.na(birth_order.y),
                   ! is.na(EA3.y),
                   ! is.na(birth_mon.x),
                   ! is.na(EA3.x),
                   ! is.na(par_age_birth.x)
                 )

mod_sur <- list()
mod_sur$females <- systemfit::systemfit(list_fml, 
                                          data = mf_pairs_sf %>% 
                                                   filter(female.x),
                                          method = "SUR"
                                        )
mod_sur$males <- systemfit::systemfit(list_fml, 
                                        data = mf_pairs_sf %>% 
                                                 filter(! female.x),
                                        method = "SUR"
                                      )

wald_eq_ratio <- purrr::map(mod_sur, 
                              nlWaldTest::nlWaldtest,
                              texts = "b[2] * b[22] = b[3] * b[21]",
                              df2 = TRUE
                            )


```


@chiappori2012fatter write down a matching model in which a person's
attractiveness is summarized by a single index. The linear version of the
model can be tested by regressing one partner's characteristics on each of the
other partner's characteristics in turn (using Seemingly Unrelated
Regressions), and checking that the coefficients have the same proportions
across each regression. We do this for birth order and PSEA. Table
\@ref(tab:tbl-sur) shows the results. We exclude year of birth dummies,
since they cause the estimation procedure to fail. 

We also check the proportionality of coefficients. We run a Wald test that

\[
\frac{\beta^{PSEA}_{PSEA}}{\beta^{PSEA}_{BO}} =
\frac{\beta^{BO}_{PSEA}}{\beta^{BO}_{BO}}
\]

where e.g. $\beta^{PSEA}_{BO}$ is the coefficient of birth order in the
regression targeting PSEA. P values are $p$ = 
`r pretty(wald_eq_ratio$males$p.value)` for males and $p$ = 
`r pretty(wald_eq_ratio$females$p.value)` for females. So for females there
is some evidence against this linear model.

```{r tbl-bo-psea-no-par-age}


age_table <- famhist %>% 
      filter(n_sibs >=2, n_sibs <= 6, ! is.na(birth_order)) %>% 
      group_by(n_sibs, birth_order) %>% 
      dplyr::summarize(
        N   = n(), 
        mfa = mean(fath_age_birth, na.rm = TRUE),
        mma = mean(moth_age_birth, na.rm = TRUE),
        .groups = "drop"
      )

fml_bo_no_par_age <- lapply(fml_bo_psea, update, 
                     . ~ . - par_age_birth.x, 
                     rhs = 1
                   )
mod_bo_no_par_age <- lapply(fml_bo_no_par_age, fixest::feols,
                     data  = mf_pairs_reg, 
                     notes = FALSE
                   )

fml_bo_fath_age <- lapply(fml_bo_no_par_age, update, 
                     . ~ . + fath_age_birth.x, 
                     rhs = 1
                   )
mod_bo_fath_age <- lapply(fml_bo_fath_age, fixest::feols,
                     data  = mf_pairs_reg, 
                     notes = FALSE
                   )

huxreg(mod_bo_no_par_age, 
        coefs = c(
          "Birth order"           = "birth_order.x", 
          "University"            = "university.xTRUE", 
          "Income"                = "first_job_pay.x", 
          "Own PSEA"              = "EA3.x",
          "Fluid IQ"              = "fluid_iq.x",
          "Height"                = "height.x"
        ),
         note = my_note,
         stars = my_stars,
         tidy_args = tidy_args
       ) %>% 
       insert_row(after = 13, "Family size dummies", rep("Yes", 4)) %>% 
       insert_row(after = 14, "Birth month dummies", rep("Yes", 4)) %>% 
       insert_row(after = 15, "Birth year dummies", rep("Yes", 4)) %>% 
       set_bottom_border(14:15, everywhere, 0) %>% 
       set_align(14:16, -1, "center") %>% 
       set_number_format(2:13, -1, 4) %>% 
       set_caption("Regressions of spouse PSEA, without controls for parents' age at respondent's birth")

```

```{r tbl-bo-psea-dummies}

# we use standard lms because fixest::wald breaks sometimes

fml_bo_psea_dummies <- purrr::map(fml_bo_psea, formula, collapse = TRUE)

fml_bo_psea_dummies <- purrr::map(fml_bo_psea_dummies, update, 
                                    . ~ . - birth_order.x + 
                                    factor(birth_order.x))

# use a formula here to preserve the `data` argument
mod_bo_psea_dummies <- purrr::map(fml_bo_psea_dummies, ~lm(.x, 
                         data = mf_pairs_reg))

mod_restricted <- purrr::map(mod_bo_psea_dummies, 
                               ~update(.x, . ~ . - factor(birth_order.x)))

my_vcov <- function (x) sandwich::vcovHC(x, type ="HC1")
wald_tests <- purrr::map2(mod_bo_psea_dummies, mod_restricted,
                          lmtest::waldtest, vcov = my_vcov)
wald_pvals <- purrr::map_dbl(wald_tests, ~ pull(tidy(.x), p.value)[2])


huxreg_coefs <- grep("factor\\(birth_order", names(coef(mod_bo_psea_dummies[[1]])),
                       value = TRUE)
names(huxreg_coefs) <- gsub("^.*(\\d).*$", "Birth order \\1",
                            huxreg_coefs)
huxreg_coefs <- c(
                   huxreg_coefs,
                   "University"  = "university.xTRUE", 
                   "Income"      = "first_job_pay.x",
                   "Own PSEA"     = "EA3.x",
                   "Parents' age at birth" = "par_age_birth.x"
                 )

huxtable::huxreg(mod_bo_psea_dummies, 
                   coefs     = huxreg_coefs,
                   note      = my_note,
                   stars     = my_stars,
                   tidy_args = tidy_args
                 ) %>% 
       insert_row(after = 19, "Wald p-value, birth order", wald_pvals) %>%
       insert_row(after = 20, "Family size dummies", rep("Yes", 4)) %>% 
       insert_row(after = 21, "Birth month dummies", rep("Yes", 4)) %>% 
       insert_row(after = 22, "Birth year dummies", rep("Yes", 4)) %>% 
       insert_row(after = 23, "Controls (IQ, height)", 
                    c("No", rep("Yes", 3))) %>% 
       set_bottom_border(21:23, everywhere, 0) %>% 
       set_align(21:24, -1, "center") %>% 
       set_number_format(2:20, -1, 4) %>% 
       set_caption("Regressions of spouse PSEA, separate birth order coefficient dummies") %>% 
       set_tb_padding(3) %>% 
       set_font_size(10)

```

```{r tbl-bo-subset-dummies}

fml_bo_psea_dummies <- fml_bo_psea %>% 
                         purrr::map(update_with_birth_order_dummies)

mod_bo_males_dummies <- lapply(fml_bo_psea_dummies[1:2], 
                                 fixest::feols, 
                                 data = mf_pairs_reg %>% 
                                   filter(x == "Male"),
                                 notes = FALSE
                        )
mod_bo_females_dummies <- lapply(fml_bo_psea_dummies[1:2], 
                                  fixest::feols, 
                                  data = mf_pairs_reg %>% 
                                    filter(x == "Female"),
                                  notes = FALSE
                          )

mf_pairs_children <- mf_pairs_reg %>% 
                       filter(n_children.x > 0, n_children.y > 0)
mod_bo_children_dummies <- lapply(fml_bo_psea_dummies[1:2], 
                             fixest::feols, 
                             data = mf_pairs_children,
                             notes = FALSE
                           )
reg_list <- list(
        "Males"                = mod_bo_males_dummies[[1]],
        "Males"                = mod_bo_males_dummies[[2]],
        "Females"              = mod_bo_females_dummies[[1]],
        "Females"              = mod_bo_females_dummies[[2]],
        "With children"        = mod_bo_children_dummies[[1]],
        "With children"        = mod_bo_children_dummies[[2]]
      )

wald_tests <- purrr::map(reg_list, fixest::wald, 
                           keep    = "birth_order",
                           se      = tidy_args$se,
                           print   = FALSE
                         )
wald_pvals <- purrr::map_dbl(wald_tests, "p")

huxreg_coefs <- grep("factor\\(birth_order", 
                       names(coef(mod_bo_males_dummies[[1]])),
                       value = TRUE)
names(huxreg_coefs) <- gsub(".*(\\d).*$", "Birth order \\1",
                            huxreg_coefs)

huxreg_coefs <- c(
                   huxreg_coefs,
                   "University"  = "university.xTRUE", 
                   "Own PSEA"     = "EA3.x",
                   "Parents' age at birth" = "par_age_birth.x"
                 )

huxtable::huxreg(reg_list, 
                   coefs = huxreg_coefs,
                   note = my_note,
                   stars = my_stars,
                   tidy_args = tidy_args
                 ) %>% 
       insert_row(after = 17, c("Wald p-value, birth order", wald_pvals)) %>% 
       insert_row(after = 18, "Family size dummies", rep("Yes", 6))   %>% 
       insert_row(after = 19, "Birth month dummies", rep("Yes", 6))   %>% 
       insert_row(after = 20, "Birth year dummies", rep("Yes", 6))    %>% 
       insert_row(after = 21, "Controls (IQ, height)", rep("Yes", 6)) %>% 
       set_bottom_border(19:21, everywhere, 0) %>%
       set_align(19:22, -1, "center") %>% 
       set_number_format(2:18, -1, 4) %>% 
       set_caption("Regressions of spouse PSEA, separate birth order coefficient dummies: subsets") %>% 
       set_tb_padding(3) %>% 
       set_lr_padding(2) %>% 
       set_font_size(9)
```

```{r pic-bo-psea-interactions, fig.align = "center", fig.cap = "Regressions of spouse PSEA: birth order dummies within different family sizes. Labels show birth order. Lines are 95 per cent confidence intervals."}
 
mf_pairs_reg$BO2 <- 1 * (mf_pairs_reg$birth_order.x == 2)
mf_pairs_reg$BO3 <- 1 * (mf_pairs_reg$birth_order.x == 3)
mf_pairs_reg$BO4 <- 1 * (mf_pairs_reg$birth_order.x == 4)
mf_pairs_reg$BO5 <- 1 * (mf_pairs_reg$birth_order.x == 5)
mf_pairs_reg$BO6 <- 1 * (mf_pairs_reg$birth_order.x == 6)
mf_pairs_reg$nsf <- factor(mf_pairs_reg$n_sibs.x)

fml_bo_psea_full <- update(fml_bo_psea[[1]], . ~ . - birth_order.x + 
                             nsf:(BO2 + BO3 + BO4 + BO5 + BO6))

mod_bo_psea_full <- fixest::feols(fml_bo_psea_full, mf_pairs_reg, 
                                    notes = FALSE)
coefs <- tidy(mod_bo_psea_full, conf.int = TRUE, se = "hetero")

suppressWarnings({
  coefs[["Family size"]] <- as.numeric(gsub("^nsf(.).*", "\\1", coefs$term))
  coefs[["Birth order"]] <- as.numeric(gsub(".*BO(.)$", "\\1", coefs$term))
})

coefs %<>% filter(`Birth order` %in% 1:8)

ns_by_n_sibs <- as.data.frame(table(mf_pairs_reg$n_sibs.x))
ns_by_n_sibs$Var1 <- as.integer(as.character(ns_by_n_sibs$Var1))

suppressWarnings({
  gg <- ggplot(coefs, aes(estimate, `Family size`, colour = `Birth order`)) + 
          scale_color_continuous(type = "viridis", name = "Birth order") + 
          geom_vline(xintercept = 0, linetype = 2) +
          geom_pointrange(aes(xmin = conf.low, xmax = conf.high), 
                         alpha = 0.7, size = 1.15, fatten = 1,
                         position = position_dodge2(width = 0.15)) + 
          geom_text(aes(label = `Birth order`), nudge_y = 0.2, size = 5) + 
          guides(colour = "none") +
          xlab("Estimate")
  print(gg) # captures the warnings inside the suppressWarnings() call
})

```

```{r tbl-bo-psea-pgs}

fml_bo_psea_pgs <- fml_bo_psea %>% 
                        purrr::map(update, 
                           . ~ . + 
                           alzheimer_resid.x + 
                           caffeine_resid.x + 
                           cognitive_ability_resid.x +
                           neuroticism_resid.x +
                           sc_substance_use_resid.x
                        )

mod_bo_psea_pgs <-  purrr::map(fml_bo_psea_pgs, fixest::feols, 
                         data = mf_pairs_reg,
                         notes = FALSE
                       )

huxreg(mod_bo_psea_pgs, 
         coefs = c(
           "Birth order"           = "birth_order.x", 
           "University"            = "university.xTRUE", 
           "Income"                = "first_job_pay.x",
           "Own PSEA"              = "EA3.x",
           "Parents' age at birth" = "par_age_birth.x",
           "Fluid IQ"              = "fluid_iq.x",
           "Height"                = "height.x"
         ),
         statistics = c(
           "N" = "nobs", 
           "R2" = "r.squared"
         ),
         note = paste(my_note,
                 "\nPolygenic scores: alzheimer's, caffeine, cognitive ability, neuroticism, substance use."),
         stars = my_stars,
         tidy_args = list(
           conf.int = FALSE, 
           cluster = list(mf_pairs_reg$f.54.0.0.x)
         )
       ) %>% 
       insert_row(after = 15, "Family size dummies", rep("Yes", 4)) %>% 
       insert_row(after = 16, "Birth month dummies", rep("Yes", 4)) %>% 
       insert_row(after = 17, "Birth year dummies", rep("Yes", 4)) %>% 
       insert_row(after = 18, "Polygenic score controls", rep("Yes", 4)) %>% 
       set_bottom_border(16:18, everywhere, 0) %>% 
       set_align(16:19, -1, "center") %>% 
       set_number_format(2:15, -1, 4) %>% 
       set_width(1) %>% 
       set_wrap(final(1), everywhere, TRUE) %>% 
       set_caption("Regressions of spouse PSEA with controls for polygenic scores")
```

```{r tbl-bo-psea-age-fte}

fml_bo_psea_age_fte <- fml_bo_psea[c(1, 2, 4)]
fml_bo_psea_age_fte[2:3] <- lapply(fml_bo_psea_age_fte[2:3], 
                                update, 
                                . ~ . - university.x + age_fulltime_edu.x)

mod_bo_psea_age_fte <- lapply(fml_bo_psea_age_fte, fixest::feols, 
                 data = mf_pairs_reg,
                 notes = FALSE
               )

huxreg(mod_bo_psea_age_fte, 
        coefs = c(
          "Birth order"           = "birth_order.x", 
          "Age left full-time educ." = "age_fulltime_edu.x", 
          "Income"                = "first_job_pay.x",
          "Parents' age at birth" = "par_age_birth.x",
          "Own PSEA"              = "EA3.x",
          "Fluid IQ"              = "fluid_iq.x",
          "Height"                = "height.x"
        ),
         note = my_note,
         stars = my_stars,
         tidy_args = tidy_args
       ) %>% 
       insert_row(after = 15, "Family size dummies", rep("Yes", 3)) %>% 
       insert_row(after = 16, "Birth month dummies", rep("Yes", 3)) %>% 
       insert_row(after = 17, "Birth year dummies", rep("Yes", 3)) %>% 
       set_bottom_border(16:17, everywhere, 0) %>% 
       set_align(16:18, -1, "center") %>% 
       set_number_format(2:15, -1, 4) %>% 
       set_width(0.8) %>% 
       set_caption("Regressions of spouse PSEA using age of leaving full-time education")
```

```{r tbl-bo-psea-no3}

mod_bo_psea_no3 <- lapply(fml_bo_psea, 
                            fixest::feols, 
                            data = mf_pairs_reg %>% 
                                     filter(n_sibs.x != 3),
                            notes = FALSE
                          )
huxreg(mod_bo_psea_no3, 
        coefs = c(
          "Birth order"           = "birth_order.x", 
          "University"            = "university.xTRUE", 
          "Income"                = "first_job_pay.x",
          "Parents' age at birth" = "par_age_birth.x",
          "Own PSEA"              = "EA3.x",
          "Fluid IQ"              = "fluid_iq.x",
          "Height"                = "height.x"
        ),
         note = my_note,
         stars = my_stars,
         tidy_args = tidy_args
       ) %>% 
       insert_row(after = 15, "Family size dummies", rep("Yes", 4)) %>% 
       insert_row(after = 16, "Birth month dummies", rep("Yes", 4)) %>% 
       insert_row(after = 17, "Birth year dummies", rep("Yes", 4)) %>% 
       set_bottom_border(16:17, everywhere, 0) %>% 
       set_align(16:18, -1, "center") %>% 
       set_number_format(2:15, -1, 4) %>% 
       set_caption("Regressions of spouse PSEA, excluding family size 3")
```


```{r tbl-sur}

mod_sur_psea <- purrr::map(mod_sur, ~ huxtable::tidy_replace(.x,
                                        tidied = 
                                          broom::tidy(.x) %>% 
                                          filter(grepl("^PSEA_", term)) %>%
                                          mutate(
                                           term = gsub("^PSEA_", "", term)
                                          ),
                                         glance = list(
                                           nobs = sum(summary(.x)$eq[[1]]$df)
                                         )
                          ))
mod_sur_bo <- purrr::map(mod_sur, ~ huxtable::tidy_replace(.x,
                                        tidied = 
                                          broom::tidy(.x) %>% 
                                          filter(grepl("^BO_", term)) %>%
                                          mutate(
                                           term = gsub("^BO_", "", term)
                                          ),
                                         glance = list(
                                           nobs = sum(summary(.x)$eq[[2]]$df)
                                         )
                          ))

# avoid warnings from glance() due to huxreg bug
suppressWarnings(huxreg(
        "Spouse PSEA"          = mod_sur_psea$males, 
        "Spouse birth order"   = mod_sur_bo$males, 
        "Spouse PSEA"        = mod_sur_psea$females, 
        "Spouse birth order" = mod_sur_bo$females,
        coefs                  = c(
                                    "Birth order" = "birth_order.x", 
                                    "PSEA"        = "EA3.x"
                                 ),
        statistics             = c(N = "nobs"),
        stars                  = my_stars
      )) %>% 
        insert_row(c("", "Males", "Males", "Females", "Females"), 
                     after = 0) %>%         
        merge_cells(1, 2:3) %>% 
        merge_cells(1, 4:5) %>% 
        set_align(1, everywhere, "center") %>% 
        insert_row("Controls (parents' age)", rep("Yes", 4), after = 6) %>% 
        insert_row("Family size dummies", rep("Yes", 4), after = 7) %>% 
        insert_row("Birth month dummies", rep("Yes", 4), after = 8) %>% 
        insert_row("Birth year dummies", rep("No", 4), after = 9) %>% 
        set_bottom_border(7:9, everywhere, 0) %>% 
        set_align(7:10, -1, "center") %>% 
        set_number_format(11, everywhere, 0) %>% 
        # merge_cells(10, 2:3) %>% 
        # merge_cells(10, 4:5) %>% 
        # set_align(10, everywhere, "center") %>% 
        set_caption(
          "Seemingly Unrelated Regressions on spouse characteristics")

```
\FloatBarrier

\newpage

### Regressions with "fake pairs"

Our dataset of pairs could still contain pairs who live in the same
postcode but are not spouses. These pairs might still show a
relationship between one partner's phenotype and the other's genotype.
For example, maybe early-born children grow up to live in richer
postcodes, along with people who have higher PSEA scores
[@abdellaoui2019genetic]. This could then bias the results. If the
coefficient for "fake pairs" is absolutely larger (smaller) than for
real pairs, then our results will be biased away from zero (towards
zero).

To sign the bias, we create a dataset of "known fake pairs". These are
opposite-sexed pairs who live in the same postcode, but do not share all
the characteristics listed for the real pairs. Specifically, from the
list of characteristics used to create our real pairs (same
homeownership status, same length of time at address, same number of
children, attended same assessment center, attended on same day, husband
reported living with spouse, wife reported living with spouse) the fake
pairs ticked exactly 5 out of 7 boxes.

```{r calc-fake-pairs}

drake::loadd(mf_pairs_fake)
prop_shared_fake <- calc_prop_shared_children(mf_pairs_fake)

n_one <- prop_shared_fake$one
n_both <- prop_shared_fake$both
```

We again use genetic children to confirm that the fake pairs are "real
fakes". Out of `r n_one` genetic children of the fake pairs, only
`r n_both` were children of both parents. Thus, the vast majority of
fake pairs do not appear to be spouses. Table
\@ref(tab:tbl-bo-psea-fake) reruns the regressions of Table
\@ref(tab:tbl-bo-psea-basic) using the fake pairs. Although the
coefficients on birth order are always negative, and significant
when controlling for parent's age, they are always absolutely smaller than the
corresponding coefficient in the main text. This suggests that
any fake pairs remaining in our data will have the effect of biasing our
results towards zero.

```{r tbl-bo-psea-fake}

drake::loadd(mf_fake_twice)
fake_pairs_reg <- regression_subset(mf_fake_twice)
mod_bo_psea_fake <- lapply(fml_bo_psea_base, fixest::feols, 
                 data = fake_pairs_reg,
                 notes = FALSE
               )

huxreg(mod_bo_psea_fake, 
         coefs = c(
           "Birth order"           = "birth_order.x", 
           "Own PSEA"              = "EA3.x",
           "Parents' age at birth" = "par_age_birth.x"
         ),
         statistics = c(
           "N"  = "nobs", 
           "R2" = "r.squared"
          ),
         note = my_note,
         stars = my_stars,
         tidy_args = tidy_args
       ) %>% 
       insert_row(after = 7, "Family size dummies", rep("Yes", 3)) %>% 
       insert_row(after = 8, "Birth month dummies", "No", "Yes", "Yes") %>% 
       insert_row(after = 9, "Birth year dummies", "No", "Yes", "Yes") %>% 
       set_bottom_border(8:9, everywhere, 0) %>% 
       set_align(8:10, -1, "center") %>% 
       set_number_format(2:7, -1, 4) %>% 
       set_width(0.8) %>% 
       set_caption("Regressions of PSEA on birth order: fake pairs")

```


\FloatBarrier

\clearpage

## Quotations on natural inequality

...your face and figure have nothing of the slave about them, and
proclaim you of noble birth.

-- *Odyssey*, Odysseus to Laertes

Citizens, we shall say to them in our tale, you are brothers, yet God
has framed you differently. Some of you have the power of command, and
in the composition of these he has mingled gold, wherefore also they
have the greatest honour; others he has made of silver, to be
auxiliaries; others again who are to be husbandmen and craftsmen he has
composed of brass and iron; and the species will generally be preserved
in the children. But as all are of the same original stock, a golden
parent will sometimes have a silver son, or a silver parent a golden
son.

-- Plato *Republic*

Nature would like to distinguish between the bodies of freemen and
slaves, making the one strong for servile labor, the other upright, and
although useless for such services, useful for political life in the
arts both of war and peace. But the opposite often happens -- that some
have the souls and others have the bodies of freemen.

-- Aristotle *Politics*

Sons have no richer endowment than the quality

A noble and brave father gives in their begetting.

-- Euripides *Heracleidae*

His looks are full of peaceful majesty,

His head by nature fram'd to wear a crown,

His hands to wield a sceptre....

-- Shakespeare *Henry VI Part 3*

A daughter of a green Grocer, walks the Streets in London dayly with a
baskett of Cabbage Sprouts, Dandelions and Spinage on her head. She is
observed by the Painters to have a beautiful Face, an elegant figure, a
graceful Step and a debonair. They hire her to Sitt. She complies, and
is painted by forty Artists, in a Circle around her. The Scientific Sir
William Hamilton outbids the Painters, Sends her to Schools for a
genteel Education and Marries her. This Lady not only causes the
Tryumphs of the Nile of Copenhagen and Trafalgar, but Seperates Naples
from France and finally banishes the King and Queen from Sicilly. Such
is the Aristocracy of the natural Talent of Beauty.

-- John Adams to Thomas Jefferson, on Emma Hamilton

\newpage

# References
