---
title: 'Room at the top: biological lock-in of social advantage'
author: David Hugh-Jones & Abdel Abdellaoui
date: "`r Sys.Date()`"
output: 
  bookdown::pdf_document2:
    toc: false
    latex_engine: xelatex
editor_options: 
  chunk_output_type: console
header-includes:
  - \usepackage{subfig}
  - \usepackage{amsmath}
  - \usepackage{amsthm}
  - \usepackage{setspace}
  - \usepackage{placeins}
  - \newtheorem{prop}{\protect\propositionname}
  - \providecommand{\propositionname}{Proposition}
---

```{r setup, include = FALSE}
library(drake)
library(dplyr)
library(magrittr)
library(AER)
library(ggplot2)
library(Hmisc)
library(huxtable)
library(broom)
library(fixest)
library(santoku)

knitr::opts_chunk$set(echo = FALSE)

drake::loadd(mf_pairs)
drake::loadd(mf_pairs_twice)
drake::loadd(famhist)
drake::loadd(resid_scores)

famhist %<>% left_join(resid_scores, by = "f.eid")
rm(resid_scores)

famhist$EA3 <- famhist$EA3_excl_23andMe_UK_resid


theme_set(theme_minimal())
```

# Introduction

Charles Murray (1995) warned of "a merging of the cognitive elite with the
affluent". On the opposite side of the political spectrum, Karl Marx (1844)
wrote "I am ugly, but I can buy the most beautiful woman.... the effect of
ugliness, its repelling power, is destroyed by money." These quotations suggest
that social advantages, such as wealth, caste or status, may be transformed into
biological advantages in the next generation, via assortative mating between
socially and genetically advantaged people. We call this process 
*biological lock-in*.

Figures \@ref(fig:pic-basic-corr-uni) and \@ref(fig:pic-basic-corr-income) 
illustrate the idea using data for spouse
pairs from UK Biobank. Figures \@ref(fig:pic-basic-corr-uni) plots one partner's 
mean polygenic score for educational attainment (PSEA) against a measure of the 
other partner's actual educational
attainment: possession of a university degree. University graduates had spouses
with higher PSEA.^[To minimize concerns about genetic stratification, i.e.
correlations between genetics and non-genetic forms of inherited advantage, PSEA
is residualized by the first 100 principal components of UK Biobank array data.]
Figure \@ref(fig:pic-basic-corr-income) plots one partner's PSEA against another
measure of social status: income. 

These figures do not prove that biological lock-in is taking place: since an
individual's own PSEA correlates with both their educational attainment, and
their income, both figures could be a result of partner selection on a purely
genetic basis. In this paper, we test the theory more rigorously, using
environmental shocks to social status that are unlikely to be correlated with
own genetics. First, we develop a simple theory of biological lock-in, to
illustrate how its effects vary with social structure.


```{r pic-basic-corr-uni, fig.subcap = c("Female PSEA by male educational attainment", "Male PSEA by female educational attainment"), fig.cap = "Social and genetic advantage among spouse pairs in UK Biobank", fig.width = 3, fig.height = 3}

pic_ss <- stat_summary(fun.data = mean_cl_normal, na.rm = TRUE)
pic_cc <- coord_cartesian(ylim = c(-0.05, 0.2))

mf_pairs %>% 
      filter(! is.na(university.m)) %>% 
      ggplot(aes(university.m, EA3.f)) + 
        pic_ss + pic_cc +
        labs(x = "Male university attendance", y = "Female PSEA")

mf_pairs %>% 
      filter(! is.na(university.f)) %>% 
      ggplot(aes(university.f, EA3.m)) + 
        pic_ss + pic_cc +
        labs(x = "Female university attendance", y = "Male PSEA")
        
```


```{r pic-basic-corr-income, fig.subcap = c("Female EA3 by male income", "Male EA3 by female income"), fig.cap = "Social and genetic advantage among spouse pairs in UK Biobank", fig.width = 3, fig.height = 3}

pic_ss <- stat_summary(fun.data = mean_cl_normal, na.rm = TRUE)
pic_cc <- coord_cartesian(ylim = c(-0.15, 0.25))

mf_pairs %>% 
      filter(! is.na(income_cat.m)) %>% 
      ggplot(aes(income_cat.m, EA3.f)) + 
        pic_ss + pic_cc + 
        labs(x = "Male income category", y = "Female PSEA")

mf_pairs %>% 
      filter(! is.na(income_cat.f)) %>% 
      ggplot(aes(income_cat.f, EA3.m)) + 
        pic_ss + pic_cc +
        labs(x = "Female income category", y = "Male PSEA")
        
```


# Theory

The population is size 1. There is a single binary genetic trait and a single
social trait. Individuals $i$ thus have $G_{i}\in\{0,1\}$ and $S_{i}\in\{0,1\}$;
write $\tau_{i}=(G_{i},S_{i})$ to fully describe $i$'s type. We will call
individuals with $\tau_{i}=(1,1)$, $(1,0)$, $(0,1)$ and $(0,0)$, b-types,
g-types, s-types and n-types respectively. The proportion of the population with
$G_{i}=1$ is denoted by $\gamma$, and the population proportion with $S_{i}=1$
is denoted by $\sigma$. In generation 1, $G_i$ and $S_i$ are assumed independent,
so e.g. the proportion of b-types is $\gamma\sigma$.

Higher values are assumed to be more attractive in the marriage market.
Indeed, individuals are divided into two classes, $H$ and $L$, for
the purpose of marital matching. Within $H$ and $L$, all individuals
form spouse pairs at random. Class $H$ is of size $\theta$: we assume
$\gamma\sigma < \theta < \min\{\gamma,\sigma\}$. We also assume $\gamma$
and $\sigma$ are not too different:

\begin{equation}
\textrm{Either }\sigma < \gamma < (1+\sigma)/2
\textrm{ or }
\gamma < \sigma < (1+\gamma)/2.
\label{eq:sigma-gamma-close}
\end{equation}

Married pairs have two children and each randomly inherits one parent's
genetic trait and one parent's social trait. As a result, $\sigma$
and $\gamma$ stay constant over generations.

B-types with $\tau_{i} = (1,1)$ are always in $H$. N-types with $\tau_{i} = (0,0)$
are always in $L$. Where the other groups fit depends on the nature
of the society. Of the $\theta - \gamma\sigma$ non-b-types
in $H$, a proportion $k$ are s-types and the remaining $1-k$ are
g-types. 

The value of $k$ gives the level of social stratification in the
society. To build intuition, we start with two extreme examples. Suppose first
that $k=1$ and that $\theta = \sigma$. This is a \emph{caste society}
where marital matching is entirely driven by social status. All individuals
in $H$ ($L$) have $S_{i} = 1$ ($S_{i} = 0$), and so do their children.
On the other hand, the proportion of individuals with $G_{i} = 1$ stays
at $\gamma$ within both groups. Thus, over generations, the distributions
of $G_{i}$ and $S_{i}$ remain independent in the population.

Suppose next that $k = 0$ and $\theta = \gamma$. This society is \emph{socially
egalitarian}: only genetics matter for marital matching. All individuals
in $H$ ($L)$ have $G_{i} = 1$ ($G_{i} = 0$), and so do their children.
The proportion of individuals with $S_{i} = 1$ stays at $\sigma$ within
both groups; hence, again, $G_{i}$ and $S_{i}$ remain independent
over generations.

In between these cases, $G_{i}$ and $S_{i}$ will become positively
correlated. We examine the correlation in the second generation.

\begin{prop}
The second-generation correlation between $G_{i}$ and $S_{i}$
is concave in $k$. If $\theta$ is close enough to $\min\{\sigma,\gamma\}$,
then the correlation has a maximum at $k\in(0,1)$.
\label{prop:second-gen-cor}
\end{prop}

The logic behind Proposition \ref{prop:second-gen-cor} is simple. As we have
seen, when spouse matching is entirely on the basis of social status, or entirely on the
basis of genetics, no correlation arises between the two. However, when spouses
are selected on the basis of both genetics and social status, then some s-types,
who only have high social status, marry some g-types, who only have valuable
genetic capital. As a result, some of their children inherit both (or neither),
and this increases the correlation between genetics and social status.

Figure \@ref(fig:pic-theory) plots the second generation correlation coefficient
of $S_i$ and $G_i$ against $k$, for $\gamma = \sigma = \theta = 0.5$.


```{r pic-theory, fig.cap = "Social stratification and biological lock-in: theory", fig.width = 4, fig.height = 3.5}
rho <- function (g, s, theta, k) {
  phi <- theta - g * s
  
  rhoval <- 1/2 * phi^2 * k * (1-k) / theta +
            1/2 * (g*(1 - s) - phi*(1 - k)) * ((1 - g)*s - phi*k) / (1 - theta)
  
  rhoval
}

g <- 0.5
s <- 0.5
theta <- 0.5
op <- par(mar = c(4, 4, 1, 1), mgp = c(2, 1, 0))
curve(rho(g, s, theta, k)/sqrt(s*(1-s)*g*(1-g)), xname = "k", ylab = expression(Corr(G[i],S[i])), lwd = 2, 
      las = 1, xaxp = c(0, 1, 1), yaxp = c(0, 0.15, 1))
par(mar = op)
```


# Data

As mentioned above, simple correlations between one partner's social status
and the other partner's genetics do not prove that biological lock-in is
taking place, because one's social status correlates with one's own genetics.
To demonstrate biological lock-in, we therefore need a source of social
advantage which is exogenous to genetics. One possibility is *birth order*.
It is well known that earlier-born children receive more parental care and
have better life outcomes. (XXX is it? Go check.) On the other hand, early-
and late-born full siblings have the same *ex ante* expected genetic endowment.
^[This might not be the case, if parents' choice of whether to
have more children is endogenous to the genetic endowment of their earlier 
children. We will check for this below.] We can therefore use birth order
as an instrument for social advantage.

We use data from UK Biobank, a study of about 500,000 individuals. 

TODO: describe N for birth order, describe PSEA calculation.


# Results 


TODO: reduced form


```{r pic-bo-university}
famhist %>% 
      filter(n_sibs <= 4, n_sibs >= 2, ! is.na(n_older_sibs)) %>% 
      mutate(
        `Birth order` = n_older_sibs + 1,
        `Family size` = n_sibs
      ) %>% 
      group_by(`Birth order`, `Family size`) %>% 
      dplyr::summarize(University = mean(university, na.rm = TRUE)) %>% 
      ggplot(aes(`Birth order`, University)) + 
      geom_col() +
      facet_grid(cols = vars(`Family size`), labeller = label_both) + 
      ylim(0, 0.15) 

```


```{r pic-bo-income}
famhist %>% 
      filter(
        n_sibs <= 4, 
        n_sibs >= 2, 
        ! is.na(n_older_sibs), 
        ! is.na(income_cat)
      ) %>% 
      mutate(
        `Birth order` = n_older_sibs + 1,
        `Family size` = n_sibs,
        `Income category` = factor(income_cat)
      ) %>% 
      ggplot(aes(`Birth order`, fill = `Income category`)) + 
      geom_bar(position = "fill") +
      facet_grid(cols = vars(`Family size`), labeller = label_both) + 
      scale_fill_brewer() + labs(y = "Proportion")
```


```{r tab-sunshine-regs, eval = FALSE}

famhist$birth_sun_s <- c(scale(famhist$birth_sun))

famhist$birth_lat_10k <- santoku::chop_width(famhist$birth_lat.0.0, 10000)
famhist$birth_lon_10k <- santoku::chop_width(famhist$birth_lon.0.0, 10000)
famhist$birth_lat_50k <- santoku::chop_width(famhist$birth_lat.0.0, 50000)
famhist$birth_lon_50k <- santoku::chop_width(famhist$birth_lon.0.0, 50000)

 

mod_uni_sun <- mod_income_sun <- mod_ea3_sun <- list()

# subset arg to lm doesn't work for poly(), which can't deal with NA values:
mod_uni_sun$poly <- famhist %>% 
                      filter(! is.na(birth_lat.0.0), ! is.na(birth_lon.0.0)) %>% 
                      {fixest::feglm(university ~ 
                                    birth_sun_s + 
                                    poly(birth_lat.0.0, birth_lon.0.0, degree = 5) |
                                    factor(birth_year) + 
                                    factor(birth_mon), 
                                  data   = .,
                                  family = "binomial"
                                )}

mod_income_sun$poly <- famhist %>% 
                        filter(! is.na(birth_lat.0.0), ! is.na(birth_lon.0.0)) %>% 
                        {lm(income_cat ~ 
                                      birth_sun_s + 
                                      poly(birth_lat.0.0, birth_lon.0.0, degree = 5) +
                                      factor(birth_year) + 
                                      factor(birth_mon), 
                                    data = .
                        )}

mod_ea3_sun$poly <- famhist %>% 
                      filter(! is.na(birth_lat.0.0), ! is.na(birth_lon.0.0)) %>% 
                      {lm(EA3 ~ 
                                    birth_sun_s + 
                                    poly(birth_lat.0.0, birth_lon.0.0, degree = 5) +
                                    factor(birth_year) + 
                                    factor(birth_mon), 
                                  data = .
                      )}

mod_uni_sun$fe50 <- fixest::feglm(
                      university ~ 
                      birth_sun_s | 
                      factor(birth_year) + 
                      factor(birth_mon) +
                      birth_lat_50k^birth_lon_50k,
                      data   = famhist,
                      family = "binomial"
                   )

mod_income_sun$fe50 <- fixest::feols(
                          income_cat ~ 
                          birth_sun_s | 
                          factor(birth_year) + 
                          factor(birth_mon) +
                          birth_lat_50k^birth_lon_50k,
                          data = famhist
                       )

mod_ea3_sun$fe50 <- fixest::feols(
                          EA3 ~ 
                          birth_sun_s | 
                          factor(birth_year) + 
                          factor(birth_mon) +
                          birth_lat_50k^birth_lon_50k,
                          data = famhist
                       )

mod_uni_sun$fe10 <- fixest::feglm(
                      university ~ 
                      birth_sun_s | 
                      factor(birth_year) +
                      factor(birth_mon) +
                      birth_lat_10k^birth_lon_10k,
                      data   = famhist,
                      family = "binomial"
                   )

mod_income_sun$fe10 <- fixest::feols(
                          income_cat ~ 
                          birth_sun_s | 
                          factor(birth_year) + 
                          factor(birth_mon) +
                          birth_lat_10k^birth_lon_10k,
                          data = famhist
                       )

mod_ea3_sun$fe10 <- fixest::feols(
                          EA3 ~ 
                          birth_sun_s | 
                          factor(birth_year) + 
                          factor(birth_mon) +
                          birth_lat_10k^birth_lon_10k,
                          data = famhist
                       )

huxreg(mod_uni_sun, 
        coefs = c("Sunshine" = "birth_sun_s"),
        tidy_args = list(se = "white")
      ) %>% 
      insert_row(after = 3, "Geography", "Polynomial", "50k dummies", "10k dummies") %>% 
      set_number_format(4, everywhere, NA) %>% 
      set_latex_float("ht")

huxreg(mod_income_sun, 
        coefs = c("Sunshine" = "birth_sun_s"),
        tidy_args = list(se = "white")
      ) %>% 
      insert_row(after = 3, "Geography", "Polynomial", "50k dummies", "10k dummies") %>% 
      set_number_format(4, everywhere, NA) %>% 
      set_latex_float("ht")

huxreg(mod_ea3_sun, 
        coefs = c("Sunshine" = "birth_sun_s"),
        tidy_args = list(se = "white")
      ) %>% 
      insert_row(after = 3, "Geography", "Polynomial", "50k dummies", "10k dummies") %>% 
      set_number_format(4, everywhere, NA) %>% 
      set_latex_float("ht")
```

## Regressions

We run regressions instrumenting one spouse's educational attainment with their
birth order. We control for overall family (i.e. sibling group)
size. Without this, then birth order would correlate with family size,
which itself may correlate with genetics, since parents of different family sizes
might have different genetic endowments. Indeed, respondents with more siblings
overall have substantially lower PSEA. We also control for subject's own PSEA. Lastly,
we allow the effect of birth order to vary depending on total family size.

Table \@ref(tab:tbl-instruments) shows instrumental variables regressions
using 2 stage least squares. We use only subjects with between 2 and 8 siblings. 
Column 1 pools men and women. Columns 2 and 3 run separate regressions by sex.
The effect is positive for both male and female subjects, but twice as strong, 
and significant, for males. 


```{r tbl-instruments}

mod_ea3_iv <- AER::ivreg(
                 EA3.y ~ university.x + factor(n_sibs.x) + EA3.x | 
                 n_older_sibs.x:factor(n_sibs.x) + factor(n_sibs.x) + EA3.x, 
                 data = mf_pairs_twice, 
                 subset = n_sibs.x >= 2 & n_sibs.x <= 8
               )

mod_ea3f_iv <- AER::ivreg(
                 EA3.f ~ university.m + factor(n_sibs.m) + EA3.m | 
                 n_older_sibs.m:factor(n_sibs.m) +factor(n_sibs.m) + EA3.m, 
                 data = mf_pairs, 
                 subset = n_sibs.m >= 2 & n_sibs.m <= 8
               )


mod_ea3m_iv <- AER::ivreg(
                 EA3.m ~ university.f + factor(n_sibs.f) + EA3.f | 
                 n_older_sibs.f:factor(n_sibs.f) + factor(n_sibs.f)  + EA3.f, 
                 data = mf_pairs, 
                 subset = n_sibs.f >= 2 & n_sibs.f <= 8
               )

# TODO: work out how to use coeftest and clustered vcov with ivreg
# 
mod_ea3_iv_or <- huxtable::tidy_override(mod_ea3_iv, extend = TRUE,
                 glance = as.list(glance(mod_ea3_iv, 
                 diagnostics = TRUE)))
mod_ea3f_iv_or <- huxtable::tidy_override(mod_ea3f_iv, extend = TRUE,
                 glance = as.list(glance(mod_ea3f_iv, 
                 diagnostics = TRUE)))
mod_ea3m_iv_or <- huxtable::tidy_override(mod_ea3m_iv,  extend = TRUE,
                 glance = as.list(glance(mod_ea3m_iv, 
                 diagnostics = TRUE)))
                                 
huxreg(
        "All spouses' EA3"    = mod_ea3_iv_or,
        "Female spouses' EA3" = mod_ea3f_iv_or, 
        "Male spouses' EA3"   = mod_ea3m_iv_or,
        coefs = c(
          "University" = "university.mTRUE", 
          "University" = "university.fTRUE",
          "University" = "university.xTRUE",
          "EA3"        = "EA3.m",
          "EA3"        = "EA3.f",
          "EA3"        = "EA3.x"
        ),
        statistics = c(
          "R2"              = "r.squared", 
          "N"               = "nobs",
          "Sargan test (p)" = "p.value.Sargan"
        ),
        note = "{stars}. Educational attainment instrumented by birth order."
      ) %>% 
      insert_row(c("Family size dummies", "Y", "Y", "Y"), after = 5) %>% 
      set_caption("Effects of educational attainment on spouse's EA3") %>% 
      set_latex_float("ht")
```


```{r}
mod_ea3_check <- lm(EA3 ~ 0 + factor(n_sibs) + n_older_sibs:factor(n_sibs), 
                    famhist, n_sibs >= 2 & n_sibs <= 8)
av <- tidy(car::Anova(mod_ea3_check)) 
av_pval <- av %>% 
              filter(term == "factor(n_sibs):n_older_sibs") %>% 
              pull(p.value)
stopifnot(av_pval > 0.05) # claim in text would be false
```


Although all children of the same parents have identical expected values of
polygenic scores, it is conceivable that parents might select family size
on the basis of genetics. For example, if the first child had a phenotype
reflecting a high (or low) polygenic score, then that might
affect the parents' decision to have a second child. To check this we 
regress birth order directly on PSEA. As before, we control for family size
and allow the effect of birth order to vary depending on family size.
Birth order variables are not jointly significant (anova p 
`r round(av_pval, 3)`) and their effect sizes are of the order of 0.01 of a 
standard deviation of PSEA.


\FloatBarrier

# Appendix

## Proof of Proposition \@ref(prop:second-gen-cor)

Write $b_{t}, g_{t}, s_{t} ,n_{t}$ for the population
proportion of each type in generation $t = 1, 2, ...$. Sometimes we drop the
subscript. Since $G_{i}$ and $S_{i}$ are independent in generation 1,

\begin{align*}
b_{1} & = \gamma\sigma; \\
g_{1} & = \gamma(1-\sigma); \\
s_{1} & = (1-\gamma)\sigma; \\
n_{1} & = (1-\gamma)(1-\sigma).
\end{align*}

Within either group $J\in\{H,L\}$, write the proportion of types
in generation $t$ as $b_{J},g_{J},s_{J},n_{J}$ for the respective
types. 

Dividing children up by their parents' types, we can calculate the
proportion with each of the four types:
\begin{center}
\begin{table}[h!]
\begin{centering}
\begin{tabular}{ccccc}

 & $b_{J}$ & $g_{J}$ & $s_{J}$ & $n_{J}$\tabularnewline
\hline 
$b_{J}$ & $1;0;0;0$ & $\frac{1}{2};\frac{1}{2};0;0$ & $\frac{1}{2};0;\frac{1}{2};0$ & $\frac{1}{4};\frac{1}{4};\frac{1}{4};\frac{1}{4}$\tabularnewline

$g_{J}$ & $\frac{1}{2};\frac{1}{2};0;0$ & $0;1;0;0$ & $\frac{1}{4};\frac{1}{4};\frac{1}{4};\frac{1}{4}$ & $0;\frac{1}{2};0;\frac{1}{2}$\tabularnewline

$s_{J}$ & $\frac{1}{2};0;\frac{1}{2};0$ & $\frac{1}{4};\frac{1}{4};\frac{1}{4};\frac{1}{4}$ & $0;0;1;0$ & $0;0;\frac{1}{2};\frac{1}{2}$\tabularnewline

$n_{J}$ & $\frac{1}{4};\frac{1}{4};\frac{1}{4};\frac{1}{4}$ & $0;\frac{1}{2};0;\frac{1}{2}$ & $0;0;\frac{1}{2};\frac{1}{2}$ & $0;0;0;0$\tabularnewline
\hline 
\end{tabular}
\par\end{centering}
\caption{Proportions of children born from different types of parents (b-type;
g-type; s-type; n-type)}
\end{table}
\par\end{center}

Summing these in each group, and using that $n_{H}=0$ and $b_{L}=0$,
gives the proportion of children of each type in each group:
\begin{align*}
b_{J,t+1} & =b_{J}+\rho_{J}\\
g_{J,t+1} & =g_{J}-\rho_{J}\\
s_{J,t+1} & =s_{J}-\rho_{J}\\
n_{J,t+1} & =n_{J}+\rho_{J}
\end{align*}

where 

\[
\rho_{J} \equiv \frac{1}{2}g_{J}s_{J}.
\]

In words, when a group contains both g-types and s-types, some of
them marry each other and produce some b- and n-type offspring: this
increases the correlation between $G_{i}$ and $S_{i}$.

Summing over the groups, we can write the new generation's proportion
of types in the whole population as
\begin{align*}
b_{t+1} & =b_{t}+\rho\\
g_{t+1} & =g_{t}-\rho\\
s_{t+1} & =s_{t}-\rho\\
n_{t+1} & =n_{t}+\rho
\end{align*}
where

\[
\rho \equiv \theta\rho_{H} + (1-\theta)\rho_{L}.
\]

We can now calculate the covariance between $G$ and $S$ in the second
generation. The covariance is just $E(G_{i}S_{i}) - E(G_{i})E(S_{i})$;
in general, at generation $t$, this is $b_{t} - \gamma\sigma$. For
$t = 2$ this is just equal to $\rho$. 

In the first generation, we can calculate:

\begin{align*}
b_{H,2} & =\gamma\sigma/\theta;\\
g_{H,2} & =(1-k)\phi/\theta\\
s_{H,2} & =k\phi/\theta;\\
n_{H,2} & =0;
\end{align*}

where 

\[
\phi \equiv \theta - \gamma\sigma.
\]

Similarly:

\begin{align*}
b_{L,2} & =0;\\
g_{L,2} & =\frac{\gamma(1-\sigma)-\phi(1-k)}{1-\theta}\\
s_{L,2} & =\frac{(1-\gamma)\sigma-\phi k}{1-\theta}\\
n_{L,2} & =\frac{(1-\gamma)(1-\sigma)}{1-\theta};
\end{align*}

and from these

\begin{align*}
\rho_{H} & =\frac{1}{2}\frac{\phi^{2}k(1-k)}{\theta^{2}};\\
\rho_{L} & =\frac{1}{2}\frac{[\gamma(1-\sigma)-\phi(1-k)][(1-\gamma)\sigma-\phi k]}{(1-\theta)^{2}}.
\end{align*}

Hence:

\begin{align*}
\rho= & \frac{1}{2}\frac{\phi^{2}k(1-k)}{\theta}+\\
 & \frac{1}{2}\frac{[\gamma(1-\sigma)-\phi(1-k)][(1-\gamma)\sigma-\phi k]}{(1-\theta)}.
\end{align*}

Note that when $k=1$ and $\theta=\sigma$, $\rho$ evaluates to 0.
Similarly, when $k=0$ and $\theta=\gamma$, $\rho$ evaluates to
0. This confirms the claims in the main text about the two extreme
example societies.

$\rho$ is a quadratic in $k$. Differentiating $\rho$ by $k$ gives

\begin{align}
\frac{d\rho}{dk} & =\frac{1}{2\theta}\phi^{2}(1-2k)+\frac{1}{2(1-\theta)}[\phi^{2}(1-2k)+\phi(\sigma-\gamma)]\nonumber \\
 & =\frac{1}{2\theta(1-\theta)}\phi^{2}(1-2k)+\frac{1}{2(1-\theta)}\phi(\sigma-\gamma)\nonumber \\
 & =\frac{1}{2(1-\theta)}\left[\frac{\phi^{2}}{\theta}(1-2k)+\phi(\sigma-\gamma)\right].\label{eq:d_rho_d_k}
\end{align}

This is decreasing in $k$, proving that $\rho$ is concave.

We find the maximum for $\rho$ by solving the first order condition.
In (\ref{eq:d_rho_d_k}), setting the term in square brackets to 0
gives:
\begin{equation}
k=\frac{1}{2}+\frac{1}{2}\frac{\theta}{\theta-\gamma\sigma}(\sigma-\gamma).\label{eq:k_FOC}
\end{equation}

We need to show this is strictly between 0 and 1 as $\theta$ approaches
$\sigma$ or $\gamma$.

First assume $\sigma<\gamma$. As $\theta\rightarrow\sigma$, the
expression (\ref{eq:k_FOC}) approaches $k=\frac{1}{2}+\frac{1}{2}\frac{(\sigma-\gamma)}{(1-\gamma)}$.
The requirement for this to be interior is that $\left|\sigma-\gamma\right|<\left|1-\gamma\right|$.
For $\sigma<\gamma$ this is equivalent to $\gamma<(1+\sigma)/2$,
which holds by condition (\ref{eq:sigma-gamma-close}).

Next assume $\gamma<\sigma$. As $\theta\rightarrow\gamma$, (\ref{eq:k_FOC})
approaches $k=\frac{1}{2}+\frac{1}{2}\frac{(\sigma-\gamma)}{(1-\sigma)}$.
This is interior when $\left|\sigma-\gamma\right|<\left|1-\sigma\right|$,
equivalently when $\sigma<(1+\gamma)/2$, which again holds by (\ref{eq:sigma-gamma-close}).




