---
title: 'Genetic lock-in of social advantage: evidence from UK Biobank'
author: David Hugh-Jones & Abdel Abdellaoui
date: "`r Sys.Date()`"
output: 
  bookdown::pdf_document2:
    toc: false
    latex_engine: xelatex
editor_options: 
  chunk_output_type: console
header-includes:
  - \usepackage{subfig}
  - \usepackage{amsmath}
  - \usepackage{amsthm}
  - \usepackage{setspace}
  - \usepackage{placeins}
  - \newtheorem{prop}{\protect\propositionname}
  - \providecommand{\propositionname}{Proposition}
---

```{r setup, include = FALSE}
library(Formula)
library(drake)
library(dplyr)
library(magrittr)
library(AER)
library(ggplot2)
library(Hmisc)
library(huxtable)
library(broom)
library(fixest)
library(santoku)
library(purrr)

knitr::opts_chunk$set(echo = FALSE)

drake::loadd(mf_pairs)
drake::loadd(mf_pairs_twice)
drake::loadd(famhist)
drake::loadd(resid_scores)

famhist %<>% left_join(resid_scores, by = "f.eid")
rm(resid_scores)

famhist$EA3 <- famhist$EA3_excl_23andMe_UK_resid


theme_set(theme_minimal())
```

# Introduction

Charles Murray (1995) warned of "a merging of the cognitive elite with the
affluent". On the opposite side of the political spectrum, Karl Marx (1844)
wrote "I am ugly, but I can buy the most beautiful woman.... the effect of
ugliness, its repelling power, is destroyed by money." These quotations suggest
that social advantages, such as wealth, caste or status, may be transformed into
biological advantages in the next generation, via assortative mating between
socially and genetically advantaged people. We call this process 
*genetic lock-in*.

Figures \@ref(fig:pic-basic-corr-uni) and \@ref(fig:pic-basic-corr-income) 
illustrate the idea using data for spouse
pairs from UK Biobank. Figures \@ref(fig:pic-basic-corr-uni) plots one partner's 
mean polygenic score for educational attainment (PSEA) against a measure of the 
other partner's actual educational
attainment: possession of a university degree. University graduates had spouses
with higher PSEA.^[To minimize concerns about genetic stratification, i.e.
correlations between genetics and non-genetic forms of inherited advantage, PSEA
is residualized by the first 100 principal components of UK Biobank array data.]
Figure \@ref(fig:pic-basic-corr-income) plots one partner's PSEA against another
measure of social status: income. 

These figures do not prove that genetic lock-in is taking place: since an
individual's own PSEA correlates with both their educational attainment, and
their income, both figures could be a result of partner selection on a purely
genetic basis. In this paper, we test the theory more rigorously, using
environmental shocks to social status that are unlikely to be correlated with
own genetics. First, we develop a simple theory of genetic lock-in, to
illustrate how its effects vary with social structure.


```{r pic-basic-corr-uni, fig.subcap = c("Female PSEA by male educational attainment", "Male PSEA by female educational attainment"), fig.cap = "Social and genetic advantage among spouse pairs in UK Biobank", fig.width = 3, fig.height = 3}

pic_ss <- stat_summary(fun.data = mean_cl_normal, na.rm = TRUE)
pic_cc <- coord_cartesian(ylim = c(-0.05, 0.2))

mf_pairs %>% 
      filter(! is.na(university.m)) %>% 
      ggplot(aes(university.m, EA3.f)) + 
        pic_ss + pic_cc +
        labs(x = "Male university attendance", y = "Female PSEA")

mf_pairs %>% 
      filter(! is.na(university.f)) %>% 
      ggplot(aes(university.f, EA3.m)) + 
        pic_ss + pic_cc +
        labs(x = "Female university attendance", y = "Male PSEA")
        
```


```{r pic-basic-corr-income, fig.subcap = c("Female EA3 by male income", "Male EA3 by female income"), fig.cap = "Social and genetic advantage among spouse pairs in UK Biobank", fig.width = 3, fig.height = 3}

pic_ss <- stat_summary(fun.data = mean_cl_normal, na.rm = TRUE)
pic_cc <- coord_cartesian(ylim = c(-0.15, 0.25))

mf_pairs %>% 
      filter(! is.na(income_cat.m)) %>% 
      ggplot(aes(income_cat.m, EA3.f)) + 
        pic_ss + pic_cc + 
        labs(x = "Male income category", y = "Female PSEA")

mf_pairs %>% 
      filter(! is.na(income_cat.f)) %>% 
      ggplot(aes(income_cat.f, EA3.m)) + 
        pic_ss + pic_cc +
        labs(x = "Female income category", y = "Male PSEA")
        
```


# Theory
There is a large population, whose members have a single genetic trait
$G_{i}$ and a single social trait $S_{i}$. These are random variables
which may be discrete or mixed and may also be correlated. Couples
match and have children: we write $G_{p}$ and $S_{p}$ for the parents'
generation, $G_{c}$ and $S_{c}$ for the children's generation, and
drop the subscripts for statements which apply to both. 

Couples match randomly among individuals of the same ``attractiveness'',
defined as
\[
A=a(G,S)=(1-k)f(G)+kv(S)
\]
where $f$ and $v$ are continuous increasing functions. Without loss
of generality, we measure $G$ and $S$ in ``attractiles'', so that
\[
A=(1-k)G+kS.
\]
We also scale them so that $EG_{p}=ES_{p}=0$, $Var(G_{p})=Var(S_{p})=1$.

Define
\[
B=kG-(1-k)S.
\]
This is orthogonal to $A$. It can be thought of as the relative weight
of ``Beauty'' or ``Brains'' in making someone attractive. We assume
that 
\begin{equation}
Var(B_{p}|A_{p})>0\textrm{ for a set of values of \ensuremath{A_{p}} of positive measure.}\label{eq:pos-var-B-given-A}
\end{equation}

Note that $EA_{p}=EB_{p}=0$. Recall that if $EX=EY=0$, $Var(X)=EX^{2}$
and $Cov(X,Y)=EXY$.

Couples have two children each, and children's value of $G$ and $S$
is equal to the mean of their parents. As a result, children have
the same value of $A$ as both their parents. 

Our main result that the correlation between genetic endowment and social
status is higher in the children's generation, so long as $0<k<1$.

\begin{prop}
When $0<k<1$, $Corr(G_{c}, S_{c}) > Corr(G_{p}, S_{p})$. When $k=0$
or $k=1$, $Corr(G_{c},S_{c}) = Corr(G_{p}, S_{p})$.
\label{prop:prop-increased-corr}
\end{prop}

Some points are in order. First, the proof does not depend strongly on the precise details of how $S$
and $G$ are inherited. All that is required is that the distribution of 
$A$ does not change too much from parents to children, that along values of
$A$, $E(B_c | A_p) \approx E(B_p | A_p)$
, and that 

\[
Var(B_{c}|A)\le Var(B_{p}|A)
\]

with strict inequality on a set of positive measure. These conditions
would hold, for example, if $S$ represented inherited wealth that
was divided up between children not too unequally (and without taking
account of the children's $G$!); or if children's expected value
of $G$ were equal to the mean of the parents' $G$, but with some
variance.

Second, the "marriage market" here is a reduced form
mechanism, encompassing that makes a difference to partner choice. For example,
if earned income affects attractiveness in the marriage market, then society's
level of meritocracy in the labour market will correlate with the value of $k$:
a more meritocratic labour market will allow people with low social status but
high human capital (partly genetically determined) to earn more, and therefore
to enter the high group.

Also, the contents of $G$ -- what counts as "good genes" in the marriage
market -- are themselves likely to vary across societies. For instance,
standards of physical attractiveness vary historically. Similarly, it is
plausible that what counted as a "good match", in terms of personality, physical
and intellectual characteristics, differed between medieval European nobility
and contemporary society. 

The model predicts variation in the strength of genetic lock-in. In
particular, in "caste societies" where there is complete endogamy within 
social status groups, there is no scope for genetic lock-in, because
marriage partners do not trade off genetics for social status. The model also 
assumes that social status is inherited randomly from one parent, in 
the same way a genetic allele is inherited. This assumption can be weakened. 
For example, if social status is inherited deterministically from the father, then the
results remain unchanged (for each pair of parents, just assume that one randomly
chosen parent is the father). 

Behaviour geneticists often make the point that in meritocratic societies,
successful people may transmit relevant genes to their offspring. (TODO: cite
relevant papers.) Like genetic genetic lock-in, meritocracy may therefore lead
to a correlation between social status and genetics. However, genetic lock-in is
a distinct, though overlapping, mechanism. Under meritocracy, certain genetic variants *cause* higher social
status and are then transmitted along with it. This logic does not apply in
non-meritocratic societies where social status is ascribed rather than earned.
Conversely, genetic variations which cause social status will become associated
with it, even in the absence of assortative mating.

By contrast, genetic lock-in applies to genetic variants that are *associated*
with higher social status in the spouse matching process. They do not need to
exert any influence whatsoever on an individual's own social status. This process
requires assortative mating, but does not require meritocracy. The logic
of genetic lock-in therefore applies to a historically much wider range of
societies, including societies where social status is wholly ascribed or
inherited, such as aristocracies.

In modern societies, both assortative mating and meritocracy are likely to be 
at play. Genetic variants that cause (e.g.) higher income and wealth will be 
inherited along with components of social status such as inherited wealth, 
networks and cultural capital. At the same time, higher social status and
"good genes" will assort in the marriage market, even if that higher social
status is caused by purely environmental variation. Our empirical analysis shows
this latter process at work.



# Data

As mentioned above, simple correlations between one partner's social status
and the other partner's genetics do not prove that genetic lock-in is
taking place, because one's social status correlates with one's own genetics.
To demonstrate genetic lock-in, we therefore need a source of social
advantage which is exogenous to genetics. One possibility is *birth order*.
It is well known that earlier-born children receive more parental care and
have better life outcomes. (XXX is it? Go check.) On the other hand, early-
and late-born full siblings have the same *ex ante* expected genetic endowment.
^[This might not be the case, if parents' choice of whether to
have more children is endogenous to the genetic endowment of their earlier 
children. We will check for this below.] We can therefore use birth order
as an exogenous shock to social status.

We use data from UK Biobank, a study of about 500,000 individuals. 

* TODO: describe N for birth order, describe PSEA calculation.
* TODO: look at mechanisms by which birth order might affect university
* TODO: get IQ data, control for it
* TODO: subset to spouses with children
* TODO: overall index of social status?


# Results 

Ideally we would instrument social status with birth order. However, our measures of
social status are noisy and incomplete. For example, we know whether subjects went
to university, but not which university they went to, and we only have rough
categorical data on household income. Birth order likely affects both these and 
other measures of social advantage. So, an instrumental variables approach
would probably fall foul of the exclusion restriction. 

Instead, we conduct a mediation analysis, following the strategy of Heckman and
Pinto (2013). We first regress our measures of own social status (i.e. income
and education) on birth order. Then, we regress spouse's PSEA on birth order,
with and without controlling for social status. Under the assumption that birth
order is exogenous to own genetics, these regressions identify the effect of
birth order, plus other environmental variables that correlate with it, on own
social status and spouse's genetics. Also, if the estimated effect of birth
order on spouse's PSEA changes when social status is included, that is evidence
that social status mediates the effect of birth order.

* TODO: clarify the empirical model, you may need help....
* TODO: estimate individual income from job SIC codes? ASHE gives data


```{r calc-bo-first-stage}
mod_bo_1st <- list()
mod_bo_1st$uni    <- lm(university ~ n_older_sibs + factor(n_sibs), famhist, 
                    subset = n_sibs >= 2 & n_sibs <= 8)
mod_bo_1st$income <- lm(income_cat ~ n_older_sibs + factor(n_sibs), famhist, 
                    subset = n_sibs >= 2 & n_sibs <= 8)

pval_bo_1st <- map(mod_bo_1st, 
                 ~tidy(.x) %>% filter(term == "n_older_sibs") %>% pull(p.value)
               )
pval_bo_1st %<>% map(format, digits = 3)
```

Figures \@ref(fig:pic-bo-uni) and \@ref(fig:pic-bo-income) show the
relationship between birth order, university education and income, separately for
respondents with 1-3 siblings. We test this formally in a linear regression
controlling for family size, which may be correlated with parents'
characteristics including genetics. Birth order is negatively correlated with
both measures (among respondents with 1-7 siblings: university $p$ = 
`r pval_bo_1st$uni`, income $p$ = `r pval_bo_1st$income`).


```{r pic-bo-uni, fig.cap = "University attendance by birth order and family size", fig.height = 3}
famhist %>% 
      filter(n_sibs <= 4, n_sibs >= 2, ! is.na(n_older_sibs)) %>% 
      mutate(
        `Birth order` = n_older_sibs + 1,
        `Family size` = n_sibs
      ) %>% 
      group_by(`Birth order`, `Family size`) %>% 
      dplyr::summarize(University = mean(university, na.rm = TRUE)) %>% 
      ggplot(aes(`Birth order`, University)) + 
      geom_col() +
      facet_grid(cols = vars(`Family size`), labeller = label_both) + 
      ylim(0, 0.15) 

```


```{r pic-bo-income, fig.cap = "Income by birth order and family size", fig.height = 3}
famhist %>% 
      filter(
        n_sibs <= 4, 
        n_sibs >= 2, 
        ! is.na(n_older_sibs), 
        ! is.na(income_cat)
      ) %>% 
      mutate(
        `Birth order` = n_older_sibs + 1,
        `Family size` = n_sibs,
        `Income category` = factor(income_cat)
      ) %>% 
      ggplot(aes(`Birth order`, fill = `Income category`)) + 
      geom_bar(position = "fill") +
      facet_grid(cols = vars(`Family size`), labeller = label_both) + 
      scale_fill_brewer(palette = 2) + 
      labs(y = "Proportion")
```


Next we run regressions of spouse PSEA on birth order, university attendance
and income.

TODO: try to explain this better!


```{r tbl-bo-psea}

fml_bo_psea <- list()

fml_bo_psea[[1]] <- EA3.y ~ n_older_sibs.x + EA3.x | factor(n_sibs.x)
fml_bo_psea[[2]] <- EA3.y ~ university.x + n_older_sibs.x + EA3.x | factor(n_sibs.x)
fml_bo_psea[[3]] <- EA3.y ~ income_cat.x + n_older_sibs.x + EA3.x | factor(n_sibs.x)
fml_bo_psea[[4]] <- EA3.y ~ income_cat.x + university.x + n_older_sibs.x + EA3.x |
      factor(n_sibs.x)

mf_pairs_reg <- mf_pairs_twice %>% 
                  filter(
                    n_sibs.x >= 2, 
                    n_sibs.x <= 8, 
                    ! is.na(n_older_sibs.x),
                    ! is.na(university.x),
                    ! is.na(income_cat.x)
                  )

mod_bo_psea <- lapply(fml_bo_psea, fixest::feols, 
                 data = mf_pairs_reg,
                 notes = FALSE
               )

huxreg(mod_bo_psea, 
        coefs = c(
          "Birth order" = "n_older_sibs.x", 
          "University"  = "university.xTRUE", 
          "Income"      = "income_cat.x",
          "Own EA3"     = "EA3.x"
        ),
         note = "{stars}. Standard errors clustered by spouse pair.",
         tidy_args = list(conf.int = FALSE, cluster = list(mf_pairs_reg$couple_id))
       ) %>% 
       insert_row(after = 9, "Family size dummies", rep("Yes", 4)) %>% 
       set_number_format(2:9, -1, 4) %>% 
       set_caption("Regressions of spouse PSEA")
```


```{r calc-bootstrap-proportions, cache = TRUE}

n_boot <- 199

bo_prop_boot <- replicate(n_boot, {
  mf_pairs_resample <- mf_pairs_reg %>% sample_frac(1, replace = TRUE)
  tmp  <- fixest::feols(fml_bo_psea[[1]], data = mf_pairs_resample, notes = FALSE)
  tmp2 <- fixest::feols(fml_bo_psea[[2]], data = mf_pairs_resample, notes = FALSE)
  tmp3 <- fixest::feols(fml_bo_psea[[3]], data = mf_pairs_resample, notes = FALSE)
  tmp4 <- fixest::feols(fml_bo_psea[[4]], data = mf_pairs_resample, notes = FALSE)
  orig_coef <- coef(tmp)["n_older_sibs.x"]
  c(
    prop_uni  = coef(tmp2)["n_older_sibs.x"]/orig_coef,
    prop_inc  = coef(tmp3)["n_older_sibs.x"]/orig_coef,
    prop_both = coef(tmp4)["n_older_sibs.x"]/orig_coef
  )
})
```


```{r clean-bootstrap-proportions}
bo_prop_ci <- apply(bo_prop_boot, 1, quantile, probs = c(.975, .025))
# the percentage reduction:
bo_prop_ci[] <- sprintf("%.1f%%", (1 - bo_prop_ci) * 100)

# point estimates from main regressions:
orig_coef <- coef(mod_bo_psea[[1]])["n_older_sibs.x"]
bo_prop_est <- apply(bo_prop_boot, 1, mean)
bo_prop_est  <- sprintf("%.1f%%", (1 - bo_prop_est) * 100)

```


Table \@ref(tab:tbl-bo-psea) shows the results. Column 1 shows the effect of 
birth order controlling only for own PSEA and family size. It establishes
that earlier-born children have spouses with higher PSEA. The effect size
is small. This is to be expected, because (a) the effects of birth order on
university, income and (presumably) other variables are small, and (b) PSEA is
measured with a lot of error. We aim to test theory rather than estimating 
an effect size, so we focus more on statistical significance.

Column 2 includes university attendance. Column 3 includes income. Column 4
includes both. We estimate the percentage decrease in the effect of birth order
across the columns, along with 95% confidence intervals for this figure, by
running bootstraps (N = `r n_boot`).^[The sample percentage decrease calculated
from the figures in Table \@ref(tab:tbl-bo-psea) is not the correct estimate,
since $E(X/Y) \ne EX/EY$.] Including university attendance alone reduces 
the effect of birth order by  `r bo_prop_est[1]`
(CI `r bo_prop_ci[1, 1]` -- `r bo_prop_ci[2, 1]`). Including income alone 
reduces the effect of birth order by `r bo_prop_est[2]` (CI 
`r bo_prop_ci[1, 2]` -- `r bo_prop_ci[2, 2]`).  Including both decreases the 
effect by `r bo_prop_est[3]` (CI `r bo_prop_ci[1, 3]` --`r bo_prop_ci[2, 3]`).


## Robustness


```{r tbl-pgs-check}

check_cor <- function (score_name) {
  fml_check <- as.formula(paste(score_name, "~ factor(n_sibs) + n_older_sibs"))
  mod_check <- lm(fml_check, famhist, n_sibs >= 2 & n_sibs <= 8)
  mod_check %>% 
        tidy() %>% 
        filter(term == "n_older_sibs") %>% 
        mutate(score = score_name)
}

drake::loadd(score_names)
score_names <- paste0(score_names, "_resid")
check_result <- purrr::map_dfr(score_names, check_cor)

if (any(check_result$p.value < 0.1/33)) stop("Some check p values < 0.1/33; rewrite text!")

n_10 <- sum(check_result$p.value < 0.1)
max_coef <- max(abs(check_result$estimate))
```


Although all children of the same parents have the same polygenic scores in
expectation, it could still be possible that genetics correlates with birth
order within the sample. This could happen if parents select family size on the
basis of genetics. For example, if the first child had a phenotype reflecting a
high (or low) polygenic score, then that might affect the parents' decision to
have a second child. Alternatively, respondents might select into the sample on
the basis of a combination of birth order and genetics. We check this by
regressing 33 different polygenic scores on birth order, controlling for family
size.^[Polygenic scores were residualized on the first principal components of
the genetic data.] Table \@ref(tab:tbl-pgs-check) shows the results. No scores
were significant at $p < 0.10/33$. `r n_10` scores were significant at $p <
0.10$ (body mass index, conscientousness, and neuroticism). Coefficients were
never greater than 0.01 of a standard deviation. Table
\@ref(tab:tbl-bo-psea-robust) in the appendix reruns regressions controlling for
these scores. Results are almost unchanged.


# Conclusion

TODO: Write!

\FloatBarrier

# Appendix

## Proof of Proposition \@ref(prop:increased-corr)

Rearranging the equations gives
\begin{align*}
S & =cA-dB,\\
G & =dA+cB
\end{align*}
where 
\begin{align*}
c & \equiv\frac{1-k}{z},\\
d & \equiv\frac{k}{z},\textrm{ and}\\
z & \equiv k^{2}+(1-k)^{2}.
\end{align*}
 We calculate the covariance of $G$ and $S$ in terms of the new
variables:
\begin{align}
Cov(G,S) & =E(GS)\nonumber \\
 & =E(dA+cB)(cA-dB)\nonumber \\
 & =E(cdA^{2}+(c^{2}-d^{2})AB-cdB^{2})\nonumber \\
 & =cd\ [Var(A)-Var(B)]+(c^{2}-d^{2})Cov(A,B)\label{eq:cov-G-S}
\end{align}
since $EA=EB=EG=ES=0$.

Now we note that since children's $A$ is equal to parents' $A$,
\begin{equation}
Var(A_{p})=Var(A_{c}),\label{eq:var-A-same}
\end{equation}
and indeed the whole distribution is the same:
\begin{equation}
Pr(A_{p}=a)=Pr(A_{c}=a)\textrm{ for all }a.\label{eq:dist-A-same}
\end{equation}

Next, note that for every value of $A$, 
\begin{equation}
E(B_{c}|A)=E(B_{p}|A),\label{eq:E-b-given-a-same}
\end{equation}
since within each pair, the children's value of $B$ is equal to the
parents' mean.

Lastly, for every value of $A$, the children's $B$ is the mean of
two random draws from $B$ given $A$. Therefore:
\begin{equation}
Var(B_{c}|A)=\frac{1}{2}Var(B_{p}|A).\label{eq:cond-var-B-less}
\end{equation}
Applying the law of total variance to $B$, and using (\ref{eq:cond-var-B-less})
and (\ref{eq:E-b-given-a-same}), along with (\ref{eq:pos-var-B-given-A}):
\begin{align}
Var(B_{c}) & =E_{A}(Var(B_{c}|A))+Var(E(B_{c}|A))\label{eq:var-B-less}\\
 & <E_{A}(Var(B_{p}|A))+Var(E(B_{p}|A))=Var(B_{p}).\nonumber 
\end{align}

Rewrite
\begin{align*}
 & Cov(A,B)\\
= & \int\int ab\ Pr(a)Pr(b|a)\ \textrm{d}b\ \textrm{d}a\\
= & \int a\ Pr(a)\int bPr(b|a)\ \textrm{d}b\ \textrm{d}a\\
= & \int a\ Pr(a)E(B|A)\ \textrm{d}a.
\end{align*}

Using (\ref{eq:dist-A-same}) and (\ref{eq:E-b-given-a-same}) then
shows that
\begin{equation}
Cov(A_{p},B_{p})=Cov(A_{c},B_{c}).\label{eq:cov-A-B-same}
\end{equation}

Plugging facts(\ref{eq:cov-A-B-same}), (\ref{eq:var-B-less}) and
(\ref{eq:cov-A-B-same}) into the expression in (\ref{eq:cov-G-S}),
we have shown
\begin{align*}
Cov(G_{p},S_{p})= & cd[Var(A_{p})-Var(B_{p})]+(c^{2}-d^{2})Cov(A_{p},B_{p})\\
\le & cd[Var(A_{c})-Var(B_{c})]+(c^{2}-d^{2})Cov(A_{c},B_{c})\\
= & Cov(G_{c},S_{c}).
\end{align*}
The inequality is strict when $cd$ is positive, equivalently when
$k\in(0,1)$, and holds with equality otherwise. The covariance and
correlation share the same sign, so this completes the proof.


\FloatBarrier

## Regressions controlling for polygenic scores

```{r tbl-bo-psea-robust}

fml_bo_psea_robust <- fml_bo_psea %>% 
                        purrr::map(Formula::as.Formula) %>% 
                        purrr::map(update, 
                           . ~ . + 
                           bmi_combined_resid.x + 
                           conscientiousness_resid.x + 
                           neuroticism_resid.x
                        )

mod_bo_psea_robust <-  purrr::map(fml_bo_psea_robust, fixest::feols, 
                         data = mf_pairs_reg,
                         notes = FALSE
                       )

huxreg(mod_bo_psea_robust, 
        coefs = c(
          "Birth order" = "n_older_sibs.x", 
          "University"  = "university.xTRUE", 
          "Income"      = "income_cat.x",
          "Own EA3"     = "EA3.x"
        ),
         note = "{stars}. Standard errors clustered by spouse pair.",
         tidy_args = list(conf.int = FALSE, cluster = mf_pairs_reg$couple_id)
       ) %>% 
       insert_row(after = 9, "Family size dummies", rep("Yes", 4)) %>% 
       insert_row(after = 10, "Polygenic score controls", rep("Yes", 4)) %>% 
       set_top_border(11, everywhere, 0) %>% 
       set_number_format(2:9, -1, 4) %>% 
       set_caption("Regressions of spouse PSEA with controls for polygenic scores")
```


