---
title: 'Effects of socio-economic status on genetics: evidence from UK Biobank'
author: David Hugh-Jones & Abdel Abdellaoui
date: "`r Sys.Date()`"
output: 
  bookdown::pdf_document2:
    toc: false
    latex_engine: xelatex
editor_options: 
  chunk_output_type: console
bibliography: bibliography.bib
spacing: double
header-includes:
  - \usepackage{subfig}
  - \usepackage{amsmath}
  - \usepackage{amsthm}
  - \usepackage{setspace}
  - \usepackage{placeins}
  - \newtheorem{prop}{\protect\propositionname}
  - \providecommand{\propositionname}{Proposition}
---

```{r setup, include = FALSE}
library(Formula)
library(car)
library(drake)
library(dplyr)
library(magrittr)
library(AER)
library(ggplot2)
library(Hmisc)
library(huxtable)
library(broom)
library(fixest)
library(santoku)
library(purrr)

knitr::opts_chunk$set(echo = FALSE)

drake::loadd(mf_pairs)
drake::loadd(mf_pairs_twice)
drake::loadd(famhist)
drake::loadd(resid_scores)

famhist %<>% left_join(resid_scores, by = "f.eid")
rm(resid_scores)

famhist$EA3 <- famhist$EA3_excl_23andMe_UK_resid

theme_set(theme_minimal())

```

# Introduction

Charles Murray (1995) warned of "a merging of the cognitive elite with the
affluent". On the opposite side of the political spectrum, Karl Marx wrote
(1974) "History is the true natural history of man" and  (1844) "I am ugly, but
I can buy the most beautiful woman.... the effect of ugliness, its repelling
power, is destroyed by money." These quotations suggest that social advantages, 
such as wealth, income, caste
or class, may have genetic consequences. If there is 
assortative mating between socially and genetically advantaged people, then
socio-economic status may become encoded in the DNA of subsequent generations.

Figures \@ref(fig:pic-basic-corr-uni) and \@ref(fig:pic-basic-corr-income) 
illustrate this idea using data for spouse
pairs from UK Biobank. Figures \@ref(fig:pic-basic-corr-uni) plots means of one
partner's polygenic score for educational attainment (PSEA), a purely genetic
measure, against a measure of the other partner's actual educational
attainment: possession of a university degree. University graduates had spouses
with higher PSEA.^[To minimize concerns about genetic stratification, i.e.
correlations between genetics and non-genetic inherited characteristics such
as culture, PSEA is residualized on the first 100 principal components of UK 
Biobank array data.] Figure \@ref(fig:pic-basic-corr-income) plots partner's
PSEA against another measure of own socio-economic status: income. Again, people
with higher income had spouses with higher PSEA.

These figures do not prove that genetic encoding of socio-economic status is
taking place. Since an individual's own PSEA correlates with both their
educational attainment, and their income, both figures could be a result of
partner selection on a purely genetic basis. In this paper, we test the theory
more rigorously, using environmental shocks to socio-economic status that are
unlikely to be correlated with own genetics. First, we develop a simple theory
of genetic lock-in, to illustrate how its effects vary with social structure.


```{r pic-basic-corr-uni, fig.subcap = c("Female PSEA by male educational attainment", "Male PSEA by female educational attainment"), fig.cap = "Social and genetic advantage among spouse pairs in UK Biobank", fig.width = 3, fig.height = 3}

pic_ss <- stat_summary(fun.data = mean_cl_normal, na.rm = TRUE)
pic_cc <- coord_cartesian(ylim = c(-0.05, 0.2))

mf_pairs %>% 
      filter(! is.na(university.m)) %>% 
      ggplot(aes(university.m, EA3.f)) + 
        pic_ss + pic_cc +
        labs(x = "Male university attendance", y = "Female PSEA")

mf_pairs %>% 
      filter(! is.na(university.f)) %>% 
      ggplot(aes(university.f, EA3.m)) + 
        pic_ss + pic_cc +
        labs(x = "Female university attendance", y = "Male PSEA")
        
```


```{r pic-basic-corr-income, fig.subcap = c("Female EA3 by male income", "Male EA3 by female income"), fig.cap = "Social and genetic advantage among spouse pairs in UK Biobank", fig.width = 3, fig.height = 3}

pic_ss <- stat_summary(fun.data = mean_cl_normal, na.rm = TRUE)
pic_cc <- coord_cartesian(ylim = c(-0.15, 0.25))

mf_pairs %>% 
      filter(! is.na(income_cat.m)) %>% 
      ggplot(aes(income_cat.m, EA3.f)) + 
        pic_ss + pic_cc + 
        labs(x = "Male income category", y = "Female PSEA")

mf_pairs %>% 
      filter(! is.na(income_cat.f)) %>% 
      ggplot(aes(income_cat.f, EA3.m)) + 
        pic_ss + pic_cc +
        labs(x = "Female income category", y = "Male PSEA")
        
```


# Theory

There is a large population, whose members have a single genetic trait
$g_{i}$ and a single social trait $s_{i}$. 

$G$ and $S$ are continuously distributed. Without loss
of generality, $EG = ES = 0$ and $Var(G) = Var(S) = 1$.^[Continuous distribution
is not strictly required.
All that is needed is for a set of pairs of positive measure to have different values
of $G$ and $S$, along a set of attractiveness curves of positive measure.] People 
pair according to an attractiveness function

\[
A(g_{i},s_{i}) = f((1-k)g_{i},ks_{i})
\]

which is smooth and strictly increasing in both its arguments. 

The key parameter is $k$. This describes the working of the society's marriage
market. If $k = 0$, only genetics $G$ are relevant in marriage markets, and social status $S$ has no effect. That is, the marriage market is highly egalitarian.
Conversely, if $k = 1$, only social status matters, to the complete exclusion of
genetics. This is the equivalent of a marriage market driven only by "caste" or
"class". Realistic societies are between these extremes.

Figure \@ref(fig:pic-intuition) shows the intuition behind our theory. The top
row shows a *caste marriage market* with $k = 1$. A typical pair is shown: children have
intermediate values of $G$ and $S$ between their two parents (hollow circle). In
this society pairs match only by social status; genetics plays no role. As a
result, while the variance of $G$ shrinks within each status group, genetics
remain uncorrelated with social status in the children's population
distribution, shown on the right. The next row shows a purely *egalitarian marriage market*
with $k = 0$. Parents match only by genetics and ignore social status. Again, as
a result there is no correlation between genetics and social status in the
children's generation. The bottom row shows a (more realistic) intermediate society, with an intermediate value of $k$. Because
both genetics and social status contribute to attractiveness, matched spouses
typically trade them off against each other. As a result, the distribution is
squeezed along the gradient of $k$, and $G$ and $S$ are correlated in the children's 
generation. We next prove this formally.

If $k = 0$, "indifference curves" of attractiveness are vertical lines
in $(G, S)$ space. If $k = 1$, they are horizontal lines. If $k \in (0,1)$
they are arbitrary downward sloping curves. 

Write $p(i)$ for $i$'s partner. Pairs always have the
same attractiveness.

\[
A(g_{i},s_{i}) = A(g_{p(i)},s_{p(i)}).
\]

Each pair has two children. We assume that both children $c(d)$ of
parents $d$, $m$ have

\begin{align}
g_{c(d)} & =\frac{g_{d}+g_{m}}{2};\label{eq:children-average}\\
s_{c(d)} & =\frac{s_{d}+s_{m}}{2}.\nonumber 
\end{align}

This is a strong assumption; we relax it later. For real world examples
approximated by it, $S$ could be wealth which is equally divided
between the children; $G$ could be a highly polygenic trait with
many small effects. Write $G_{p}, S_{p}$ to denote the population
variables in the parents' generation; $G_{c}, S_{c}$ for the children's
generation. 

\begin{prop}
\label{prop:prop1}
(i) $Cov(G_{c}, S_{c}) \ge Cov(G_{p},S_{p})$, with strict inequality
if and only if $0 < k < 1$. 

(ii) If $corr(G_{p}, S_{p}) \ge 0$, then $corr(G_{c},S_{c}) \ge corr(G_{p},S_{p})$,
with strict inequality if and only if $0 < k < 1$ \uline{or} 
$corr(G_{p}, S_{p}) > 0$.
\end{prop}


\begin{proof}

Within each pair $i,p(i)$ write $d$ for the person with $s_{d}>s_{p(d)}$
and $m$ for $p(d)$. (Think of these as "dukes" and "milkmaids",
or if you prefer "duchesses" and "tennis instructors".) If
$k < 1$, then $g_{d} < g_{m}$. (If $k = 0$, then define $d$ as the person
with $g_{d} < g_{p(d)}$.)

We integrate over the "dukes" to calculate the covariance in the
parents' generation:
\[
cov(G_{p},S_{p})=\int\frac{1}{2}(g_{d}s_{d}+g_{p(d)}s_{p(d)})\,\mathrm{d}d.
\]

For the children, the equivalent expression is
\[
cov(G_{c},S_{c})=\int g_{c(d)}s_{c(d)}\mathrm{d}d,
\]

observing that $EG_{c} = ES_{c}=0$ from (\ref{eq:children-average}).

Take an arbitrary pair $d,m$. Write

\begin{align*}
g_{d}s_{d} & =(g_{c}-\Delta g)(s_{c}+\Delta s);\\
g_{m}s_{m} & =(g_{c}+\text{\ensuremath{\Delta g)(s_{c}-\Delta s)}}
\end{align*}

where 

\begin{align*}
\text{\ensuremath{\Delta g=\frac{g_{m}-g_{d}}{2}\ge0}}, & \textrm{strictly so if and only if }k>0;\\
\Delta s=\frac{s_{d}-s_{m}}{2}\ge0 & \textrm{strictly so if and only if }k<1.
\end{align*}

Taking the average of the parents gives

\[
\frac{1}{2}(g_{d}s_{d}+g_{m}s_{m})= g _{c}s_{c}-\Delta g\Delta s.
\]

This is less than $g_{c}s_{c}$ if $0 < k < 1$, and equal to it if $k = 0$
or $k = 1$. Plugging this into the integral shows that 

\[
cov(G_{p},S_{p})\le cov(G_{c},S_{c})
\]

again with strict inequality if and only if $0 < k < 1$. This proves the first
part. A similar argument, showing $var(G_c) \le var(G_p)$ and $var(S_c) \le
var(S_p)$, proves the second part (see the appendix). Figure
\ref{fig:Correlation-counterexample} in the appendix shows that the condition in
the second part cannot be relaxed further.

\end{proof}


```{r pic-intuition, fig.subcap = c("Caste society ($k = 1$): parents", "Caste society: children", "Egalitarian society ($k = 0$): parents", "Egalitarian society: children", "Intermediate society ($0 < k < 1$): parents", "Intermediate society: children"), fig.cap = "Theory: shaded area is the population distribution. Dotted lines are attractiveness isoquants. Solid dots are example parents, transparent dots are example children. The right hand side shows the children's generation.", fig.ncol = 2, fig.width = 2.75, fig.height = 2.4, fig.align='center'}


setup_plot <- function () {
  par(mar = rep(2.1, 4))
  plot.new()
  plot.window(c(-1.2, 1.2), c(-1.2, 1.2))
  arrows(x0 = -1.1, x1 = 1.1, y0 = -1.1, length = .1)
  arrows(y0 = -1.1, y1 = 1.1, x0 = -1.1, length = .1)
  text(-1.1, 1.2, "S", cex = 0.7)
  text(1.2, -1.1, "G", cex = 0.7)
}

draw_dist <- function (w) {
  if (! is.matrix(w)) w <- matrix(c(w, 0, 0, 1), nrow = 2)
  mygrey <- grey(.85, .03)
  for (rad in seq(0.2, 1, 0.1)) {
    v <- car::ellipse(c(0, 0), shape = w, radius = rad, add = TRUE, center.pch = NULL,
               fill = TRUE, col = mygrey, lwd = 0, segments = 241)
  }
  
  v
}

draw_dist_with_a <- function (width, shift, curved_lines = FALSE) {
  setup_plot()  
  v <- draw_dist(width)
  nr_v <- nrow(v)
  
  rotation <- (-shift + seq_len(nr_v)) %% nr_v
  rotation[rotation == 0] <- nr_v
  
  v <- v[rotation, ]
  v_rev <- v[nr_v:1, ]
  
  ivls <- quantile(seq_len(nr_v/2), seq(0.2, 0.8, .2), type = 1)
  
  start_x <- v[ivls, 1]
  end_x   <- v_rev[ivls, 1]
  start_y <- v[ivls, 2]
  end_y <- v_rev[ivls, 2]
  
  if (curved_lines) {
    for (l in seq_along(start_x)) {
      sx <- start_x[l] + c(0, 0.15, 0.85, 1) * (end_x[l] - start_x[l])
      sy <- start_y[l] + c(0, 0.05, 0.65, 1) * (end_y[l] - start_y[l])
      spl <- stats::spline(sx, sy)
      if (l ==2) mid_spl <- spl
      lines(spl, lty = 2)
    }
    
    start_x <- mid_spl$x[2]
    start_y <- mid_spl$y[2]
    end_x <- mid_spl$x[8]
    end_y <- mid_spl$y[8]
    inset <- 0
  } else {
    segments(x0 = start_x, x1 = end_x, y0 = start_y, y1 = end_y, lty = 2)  
    
    ivls <- ivls[2:3]
  
    start_x <- mean(v[ivls, 1])
    end_x   <- mean(v_rev[ivls, 1])
    start_y <- mean(v[ivls, 2])
    end_y   <- mean(v_rev[ivls, 2])
    inset <- 0.15
  }
  
  
  along <- function (tt) list(
    x = start_x + tt * (end_x - start_x),
    y = start_y + tt * (end_y - start_y)
  )
  
  end_inset <- 0.15
  starts <- along(c(inset, 1 - inset))
  ends   <- along(c(0.5 - end_inset, 0.5 + end_inset))
  pts <- along(c(inset/2, .5, 1 - inset/2))
  arrows(x0 = starts$x, x1 = ends$x, y0 = starts$y, y1 = ends$y,
         length = 0.05)
  
  points(x = pts$x, y = pts$y, pch = c(19, 21, 19), cex = 0.8)
}

draw_dist_with_a(.8, 0)
setup_plot()
invisible(draw_dist(matrix(c(.4, 0, 0, 1), nrow = 2)))

draw_dist_with_a(.8, 60)
setup_plot()
invisible(draw_dist(matrix(c(.8, 0, 0, .4), nrow = 2)))

draw_dist_with_a(.8, -30, curved_lines = TRUE)
setup_plot()
invisible(draw_dist(matrix(c(1, .4, .4, .8), nrow = 2)))



```




## Robustness

We now relax the condition that children are exactly at the mean
of their parents' values for $G$ and $S$. Let

\begin{align*}
g_{c(i)} & =\bar{g}_{i}+\varepsilon_{i}^{G}\\
s_{c(i)} & =\bar{s}_{i}+\varepsilon_{i}^{S}
\end{align*}

where 

\[
\bar{g}_{i} = \frac{g_{i} + g_{p(i)}}{2}; \bar{s}_{i} = \frac{s_{i} + s_{p(i)}}{2}
\]

, $\varepsilon^{G}$ has mean 0 and variance $\sigma_{G}^{2}$ and
$\varepsilon^{S}$ has mean 0 and variance $\sigma_{S}^{2}$. 

\begin{prop}
\label{prop:robustness}
\begin{enumerate}
\item if $\sigma_{G}^{2}$ and $\sigma_{S}^{2}$ are small enough and $corr(G_{p},S_{p}) \ge 0$,
then $corr(G_{c},S_{c}) > corr(G_{p},S_{p})$ for $k \in (0,1)$.
\item if $\varepsilon^{G}$ and $\varepsilon^{S}$ are uncorrelated with
each other and with $\bar{G}$ and $\bar{S}$; and if $G_{p}$ and
$S_{p}$ are uncorrelated, then $corr(G_{c},S_{c}) \ge 0$, with strict
inequality if and only if $0<k<1$.
\end{enumerate}
\end{prop}

The conditions in Proposition \label{prop:robustness} are quite plausible. For
$G$, they require that either variance in siblings' scores on some summary statistic
is not too large, or that it is uncorrelated with the parents' scores. Both of 
these hold for most polygenic scores, which are additive sums of many small effects
of alleles derived randomly from one or other parent. For $S$, the conditions
would hold, for example, if $S$ measures wealth, which is inherited not too
unequally between siblings; or if wealth is inherited unequally but not in a
way that correlates with $S$ or $G$.

It is worth considering what kind of social arrangements would *violate* these
conditions. For example, suppose that parents' combined wealth is inherited by
the child with the lowest value of $g_{c(i)}$. This creates a negative correlation
between $s_{c(i)}$ and $g_{c(i)}$.

## Comparative statics

In the model, intermediate values of $k$ drive increased covariance between
genetics and social status. That naturally raises the question whether the
change in covariance and/or correlation is increasing in $k$ towards some
maximum value in $(0,1)$, then decreasing, i.e., whether it is quasiconcave
in $k$. In general the answer is no, even on the assumption that

\[
A(g_{i},s_{i}) = (1-k)g_{i} + ks_{i}
\]

i.e. that indifference curves are straight lines in $(G, S)$ space. However,
quasiconcavity in $k$ does hold when indifference curves are straight and $(G, S)$ are jointly normally distributed. We conjecture that it also holds whenever the 
distribution $(G, S)$ is quasiconcave.

* TODO: I can prove covariance is quasiconcave in $k$ for normal $G, S$. Prove
  the same for correlation?

## Discussion

The "marriage market" here is a reduced form
mechanism, encompassing that makes a difference to partner choice. For example,
if earned income affects attractiveness in the marriage market, then society's
level of meritocracy in the labour market will correlate with the value of $k$:
a more meritocratic labour market will allow people with low social status but
high human capital (partly genetically determined) to earn more, and therefore
to enter the high group.

Also, the contents of $G$ -- what counts as "good genes" in the marriage
market -- are themselves likely to vary across societies. For instance,
standards of physical attractiveness vary historically. Similarly, it is
plausible that what counted as a "good match", in terms of personality, physical
and intellectual characteristics, differed between medieval European nobility
and contemporary society. 

The model predicts variation in the strength of genetic lock-in. In
particular, in "caste societies" where there is complete endogamy within 
social status groups, there is no scope for genetic lock-in, because
marriage partners do not trade off genetics for social status. The model also 
assumes that social status is inherited randomly from one parent, in 
the same way a genetic allele is inherited. This assumption can be weakened. 
For example, if social status is inherited deterministically from the father, then the
results remain unchanged (for each pair of parents, just assume that one randomly
chosen parent is the father). 

Behaviour geneticists often make the point that in meritocratic societies,
successful people may transmit relevant genes to their offspring. (TODO: cite
relevant papers.) Like genetic genetic lock-in, meritocracy may therefore lead
to a correlation between social status and genetics. However, genetic lock-in is
a distinct, though overlapping, mechanism. Under meritocracy, certain genetic variants *cause* higher social
status and are then transmitted along with it. This logic does not apply in
non-meritocratic societies where social status is ascribed rather than earned.
Conversely, genetic variations which cause social status will become associated
with it, even in the absence of assortative mating.

By contrast, genetic lock-in applies to genetic variants that are *associated*
with higher social status in the spouse matching process. They do not need to
exert any influence whatsoever on an individual's own social status. This process
requires assortative mating, but does not require meritocracy. The logic
of genetic lock-in therefore applies to a historically much wider range of
societies, including societies where social status is wholly ascribed or
inherited, such as aristocracies.

In modern societies, both assortative mating and meritocracy are likely to be 
at play. Genetic variants that cause (e.g.) higher income and wealth will be 
inherited along with components of social status such as inherited wealth, 
networks and cultural capital. At the same time, higher social status and
"good genes" will assort in the marriage market, even if that higher social
status is caused by purely environmental variation. Our empirical analysis shows
this latter process at work.



# Data

As mentioned above, simple correlations between one partner's social status
and the other partner's genetics do not prove that genetic lock-in is
taking place, because one's social status correlates with one's own genetics.
To demonstrate genetic lock-in, we therefore need a source of social
advantage which is exogenous to genetics. One possibility is *birth order*.
It is well known that earlier-born children receive more parental care and
have better life outcomes. (XXX is it? Go check.) On the other hand, early-
and late-born full siblings have the same *ex ante* expected genetic endowment.
^[This might not be the case, if parents' choice of whether to
have more children is endogenous to the genetic endowment of their earlier 
children. We will check for this below.] We can therefore use birth order
as an exogenous shock to social status.

We use data from UK Biobank, a study of about 500,000 individuals. 

* TODO: describe N for birth order, describe PSEA calculation.
* TODO: look at mechanisms by which birth order might affect university
* TODO: get IQ data, control for it
* TODO: subset to spouses with children
* TODO: overall index of social status?


# Results 

Ideally we would instrument social status with birth order. However, our measures of
social status are noisy and incomplete. For example, we know whether subjects went
to university, but not which university they went to, and we only have rough
categorical data on household income. Birth order likely affects both these and 
other measures of social advantage. So, an instrumental variables approach
would probably fall foul of the exclusion restriction. 

Instead, we conduct a mediation analysis, following the strategy of Heckman and
Pinto (2013). We first regress our measures of own social status (i.e. income
and education) on birth order. Then, we regress spouse's PSEA on birth order,
with and without controlling for social status. Under the assumption that birth
order is exogenous to own genetics, these regressions identify the effect of
birth order, plus other environmental variables that correlate with it, on own
social status and spouse's genetics. Also, if the estimated effect of birth
order on spouse's PSEA changes when social status is included, that is evidence
that social status mediates the effect of birth order.

* TODO: clarify the empirical model, you may need help....
* TODO: estimate individual income from job SIC codes? ASHE gives data


```{r calc-bo-first-stage}
mod_bo_1st <- list()
mod_bo_1st$uni    <- lm(university ~ n_older_sibs + factor(n_sibs), famhist, 
                    subset = n_sibs >= 2 & n_sibs <= 8)
mod_bo_1st$income <- lm(income_cat ~ n_older_sibs + factor(n_sibs), famhist, 
                    subset = n_sibs >= 2 & n_sibs <= 8)

pval_bo_1st <- map(mod_bo_1st, 
                 ~tidy(.x) %>% filter(term == "n_older_sibs") %>% pull(p.value)
               )
pval_bo_1st %<>% map(format, digits = 3)
```

Figures \@ref(fig:pic-bo-uni) and \@ref(fig:pic-bo-income) show the
relationship between birth order, university education and income, separately for
respondents with 1-3 siblings. We test this formally in a linear regression
controlling for family size, which may be correlated with parents'
characteristics including genetics. Birth order is negatively correlated with
both measures (among respondents with 1-7 siblings: university $p$ = 
`r pval_bo_1st$uni`, income $p$ = `r pval_bo_1st$income`).


```{r pic-bo-uni, fig.cap = "University attendance by birth order and family size", fig.height = 3}
famhist %>% 
      filter(n_sibs <= 4, n_sibs >= 2, ! is.na(n_older_sibs)) %>% 
      mutate(
        `Birth order` = n_older_sibs + 1,
        `Family size` = n_sibs
      ) %>% 
      group_by(`Birth order`, `Family size`) %>% 
      dplyr::summarize(University = mean(university, na.rm = TRUE), 
                       .groups = "drop_last") %>% 
      ggplot(aes(`Birth order`, University)) + 
        geom_col() +
        facet_grid(cols = vars(`Family size`), labeller = label_both) 

```


```{r pic-bo-income, fig.cap = "Income by birth order and family size", fig.height = 3}
famhist %>% 
      filter(
        n_sibs <= 4, 
        n_sibs >= 2, 
        ! is.na(n_older_sibs), 
        ! is.na(income_cat)
      ) %>% 
      mutate(
        `Birth order` = n_older_sibs + 1,
        `Family size` = n_sibs,
        `Income category` = factor(income_cat)
      ) %>% 
      ggplot(aes(`Birth order`, fill = `Income category`)) + 
      geom_bar(position = "fill") +
      facet_grid(cols = vars(`Family size`), labeller = label_both) + 
      scale_fill_brewer(palette = 2) + 
      labs(y = "Proportion")
```


Next we run regressions of spouse PSEA on birth order, university attendance
and income.

* TODO: include age FTE as well as university - to check university has an
    "extra effect" - this slightly suggests that uni is a "marriage
    market" and not just granting extra skills
* TODO: get profession data from Abdel for ASHE
* TODO: enquire about job history data - does it predict current profession well?
  - otherwise we risk the critique that current profession is endogenous to
    spouse PSEA
* TODO: consider alternative exogenous shocks to income. For example, some
    professions are more "cyclical" than others wrt recessions. If we could
    do predicted income at age 21-25 from business cycle X profession, that might
    count as exogenous. (Could use an independent source to estimate evolution
    of incomes, e.g. GHS or BHPS)
* TODO: get IQ data from Abdel - again this controls for skills
* TODO: get 2178 "overall health" and 2188 "longstanding illness/disability/infirmity"
   - control for effect on health
* Overall idea here is to unpack the different effects of birth order, separate
   them from university/income.
* TODO: control for age (polynomial) and birth month (seasonality)
* TODO: check whether sunshine is well established in the medical/health literature 
    as an environmental effect. If not, probably skip it.
* TODO: split up by gender.
* TODO: look up "birth order" and "early parental investment" for economists on
  the topic
* TODO: maybe split by number of brothers/sisters?
* TODO: compare early life conditions for 1st/2nd/3rd etc. children, within
  fixed family size, using e.g. UKHLS?

```{r tbl-bo-psea}


fml_bo_psea <- list()

fml_bo_psea[[1]] <- EA3.y ~ n_older_sibs.x | factor(n_sibs.x)
fml_bo_psea[[2]] <- EA3.y ~ university.x + n_older_sibs.x | factor(n_sibs.x)
fml_bo_psea[[3]] <- EA3.y ~ income_cat.x + n_older_sibs.x | factor(n_sibs.x)
fml_bo_psea[[4]] <- EA3.y ~ income_cat.x + university.x + n_older_sibs.x |
      factor(n_sibs.x)

fml_bo_psea <- lapply(fml_bo_psea, Formula::as.Formula)

# common controls
fml_bo_psea <- lapply(fml_bo_psea, update, 
                        . ~ . + EA3.x + age_at_recruitment.x +
                          I(age_at_recruitment.x^2), 
                        rhs = 1
                      )

mf_pairs_reg <- mf_pairs_twice %>% 
                  filter(
                    n_sibs.x >= 2, 
                    n_sibs.x <= 8, 
                    ! is.na(n_older_sibs.x),
                    ! is.na(university.x),
                    ! is.na(income_cat.x)
                  )

mod_bo_psea <- lapply(fml_bo_psea, fixest::feols, 
                 data = mf_pairs_reg,
                 notes = FALSE
               )

huxreg(mod_bo_psea, 
        coefs = c(
          "Birth order" = "n_older_sibs.x", 
          "University"  = "university.xTRUE", 
          "Income"      = "income_cat.x",
          "Own EA3"     = "EA3.x"
        ),
         note = "{stars}. Standard errors clustered by spouse pair.",
         tidy_args = list(conf.int = FALSE, cluster = list(mf_pairs_reg$couple_id))
       ) %>% 
       insert_row(after = 9, "Family size dummies", rep("Yes", 4)) %>% 
       set_number_format(2:9, -1, 4) %>% 
       set_caption("Regressions of spouse PSEA")
```



```{r calc-bootstrap-proportions, cache = TRUE}

n_boot <- 199

bo_prop_boot <- replicate(n_boot, {
  mf_pairs_resample <- mf_pairs_reg %>% sample_frac(1, replace = TRUE)
  tmp  <- fixest::feols(fml_bo_psea[[1]], data = mf_pairs_resample, notes = FALSE)
  tmp2 <- fixest::feols(fml_bo_psea[[2]], data = mf_pairs_resample, notes = FALSE)
  tmp3 <- fixest::feols(fml_bo_psea[[3]], data = mf_pairs_resample, notes = FALSE)
  tmp4 <- fixest::feols(fml_bo_psea[[4]], data = mf_pairs_resample, notes = FALSE)
  orig_coef <- coef(tmp)["n_older_sibs.x"]
  c(
    prop_uni  = coef(tmp2)["n_older_sibs.x"]/orig_coef,
    prop_inc  = coef(tmp3)["n_older_sibs.x"]/orig_coef,
    prop_both = coef(tmp4)["n_older_sibs.x"]/orig_coef
  )
})
```


```{r clean-bootstrap-proportions}
bo_prop_ci <- apply(bo_prop_boot, 1, quantile, probs = c(.975, .025))
# the percentage reduction:
bo_prop_ci[] <- sprintf("%.1f%%", (1 - bo_prop_ci) * 100)

# point estimates from main regressions:
orig_coef <- coef(mod_bo_psea[[1]])["n_older_sibs.x"]
bo_prop_est <- apply(bo_prop_boot, 1, mean)
bo_prop_est  <- sprintf("%.1f%%", (1 - bo_prop_est) * 100)

```


Table \@ref(tab:tbl-bo-psea) shows the results. Column 1 shows the effect of 
birth order controlling only for own PSEA and family size. It establishes
that earlier-born children have spouses with higher PSEA. The effect size
is small. This is to be expected, because (a) the effects of birth order on
university, income and (presumably) other variables are small, and (b) PSEA is
measured with a lot of error. We aim to test theory rather than estimating 
an effect size, so we focus more on statistical significance.

Column 2 includes university attendance. Column 3 includes income. Column 4
includes both. We estimate the percentage decrease in the effect of birth order
across the columns, along with 95% confidence intervals for this figure, by
running bootstraps (N = `r n_boot`).^[The sample percentage decrease calculated
from the figures in Table \@ref(tab:tbl-bo-psea) is not the correct estimate,
since $E(X/Y) \ne EX/EY$.] Including university attendance alone reduces 
the effect of birth order by  `r bo_prop_est[1]`
(CI `r bo_prop_ci[1, 1]` -- `r bo_prop_ci[2, 1]`). Including income alone 
reduces the effect of birth order by `r bo_prop_est[2]` (CI 
`r bo_prop_ci[1, 2]` -- `r bo_prop_ci[2, 2]`).  Including both decreases the 
effect by `r bo_prop_est[3]` (CI `r bo_prop_ci[1, 3]` --`r bo_prop_ci[2, 3]`).

Our next regressions split up the data into subsets. We regress male
birth order on female spouses' PSEA; female birth order on male spouses' 
PSEA; and the subset of individuals who had children. Table 
\@ref(tab:tbl-bo-subsets) shows the results.

* TODO: describe results
* TODO: very few couple pairs (< 50%) report the same number of children. Why?
  - equally likely to be males or females reporting more.

```{r tbl-bo-subsets}

mod_bo_males <- lapply(fml_bo_psea[1:2], fixest::feols, 
                 data = mf_pairs_reg %>% filter(x == "Male"),
                 notes = FALSE
               )
mod_bo_females <- lapply(fml_bo_psea[1:2], fixest::feols, 
                 data = mf_pairs_reg %>% filter(x == "Female"),
                 notes = FALSE
               )

mf_pairs_children <- mf_pairs_reg %>% 
                   filter(n_children.x > 0, n_children.y > 0)
mod_bo_children <- lapply(fml_bo_psea[1:2], fixest::feols, 
                 data = mf_pairs_children,
                 notes = FALSE
               )

ta <-  list(conf.int = FALSE, cluster = list(mf_pairs_children$couple_id))

huxreg(c(mod_bo_males, mod_bo_females, mod_bo_children),
       coefs = c(
          "Birth order" = "n_older_sibs.x", 
          "University"  = "university.xTRUE", 
          "Own EA3"     = "EA3.x"
        ),
         note = "{stars}. Standard errors for columns 5 and 6 clustered by spouse pair.",
          tidy_args = list(list(), list(), list(), list(), ta, ta)
       ) %>% 
       # insert_row(after = 9, "Family size dummies", rep("Yes", 4)) %>% 
       # set_number_format(2:9, -1, 4) %>% 
       set_caption("Regressions of spouse PSEA: subsets")
```


## Robustness


```{r tbl-pgs-check}

check_cor <- function (score_name) {
  fml_check <- as.formula(paste(score_name, "~ factor(n_sibs) + n_older_sibs"))
  mod_check <- lm(fml_check, famhist, n_sibs >= 2 & n_sibs <= 8)
  mod_check %>% 
        tidy() %>% 
        filter(term == "n_older_sibs") %>% 
        mutate(score = score_name)
}

drake::loadd(score_names)
score_names <- paste0(score_names, "_resid")
check_result <- purrr::map_dfr(score_names, check_cor)

if (any(check_result$p.value < 0.1/33)) stop("Some check p values < 0.1/33; rewrite text!")

n_10 <- sum(check_result$p.value < 0.1)
if (n_10 != 3) stop("Different number of PGS had p < 0.10; rewrite text!")

max_coef <- max(abs(check_result$estimate))
```


Although all children of the same parents have the same polygenic scores in
expectation, it could still be possible that genetics correlates with birth
order within the sample. This could happen if parents select family size on the
basis of genetics. For example, if the first child had a phenotype reflecting a
high (or low) polygenic score, then that might affect the parents' decision to
have a second child. Alternatively, respondents might select into the sample on
the basis of a combination of birth order and genetics. We check this by
regressing 33 different polygenic scores on birth order, controlling for family
size.^[Polygenic scores were residualized on the first principal components of
the genetic data.]  No scores
were significant at $p < 0.10/33$. `r n_10` scores were significant at $p <
0.10$ (body mass index, conscientousness, and neuroticism). Coefficients were
never greater than 0.01 of a standard deviation. Table
\@ref(tab:tbl-bo-psea-robust) in the appendix reruns regressions controlling for
these scores. Results are almost unchanged.

One early explanation for birth order effects was that these could be due to
genetic mutations in older parents. More recent research has mostly rejected
this explanation in favour of "social" explanations. (TODO: cite that paper.)
Nevertheless, we rerun regressions controlling for father's age at birth,
using the subset of respondents who reported this data. (Cochran and Harpending
(2013) report that mutational load is approximately linear in father's age,
while it is constant in mothers' age. This is because female eggs are created 
once, before the female's own birth, while male sperma cells continue dividing 
throughout the lifespan.)  Table \@ref(tab:tbl-bo-fath-age) in the appendix
shows that the effect of birth order remains significant and surprisingly 
increases in size. 

* TODO: given that father's age correlates with birth order and has
  an opposite signed effect, should we use it as a control in the main
  regressions?
  - Problem is selection: people who answer father's age question are 
    those whose father is still alive.
* TODO: add mother's age also (and both)


# Conclusion

* Other points: to study inequality,  study marriage markets: this is where
  genes and SES meet. See Philip Oreopoulos, "Hint: money isn't everything"
  - need to know the existing literature... 

Behaviour geneticists have pointed out that in meritocratic societies, genetics
and social status will be correlated, because some genetic variants will be
causally linked to success in the labour market. We argue that causality can
also go the other way. If social factors and genetics both matter in marriage
markets, then "good genes" will become associated with markers of social status,
such as class, education or income.

Our empirical analysis shows that in a modern Western democracy, earlier-born
children had spouses with higher PSEA. Thus, an environmental shock has effects
on the genetics of people's spouses and becomes encoded in the genetics of their
children. We also provided evidence that these effects are mediated by social
status: income and education.

Social scientists have tended to assume that social status is unrelated to
"natural" differences between individuals, including genetic differences. This 
view dates back at least to Hobbes (1648). By contrast, it is widely 
believed in many societies that innate traits do vary across social classes. The
ancient Greeks described upper class people as *kaloi kagathoi* ("fine and good"),
while the Roman nobility were the *optimates* ("best"). See the appendix for a selection
of relevant historical quotations. This belief has been explained by invoking
people's tendency to believe in a just world [@furnham1993just], or as an
ideology promoted by the dominant class [e.g. @gramsci1971selections]. However, it
may also simply have been a recognition of (social) reality. In other words,  
the belief that elites are taller, stronger, better-looking, etc. is not much 
different from the belief that elites are richer and more powerful, and is likely 
to be held for fairly similar reasons. The belief that human traits are
inherited across generations is probably quite widespread across different cultures
[@moya2015reasoning]: this, plus the observation that status matters in marriage
markets, provides a cognitive basis for reasoning to the conclusion of class 
differences.

Over history, the marriage market mechanism has probably operated in a much greater
variety of societies than the labour market mechanism, because societies in
which status can be earned are relatively rare compared to those in which status
is inherited. (TODO: cite WEIRD? Or "status vs contract"?) As a result, we 
would predict long-standing correlations between status and causal 
genetic variants. This prediction could be tested on ancient DNA data.

Our analysis also has implications for the practice of controlling polygenic
scores by partialling out principal components of genetic data. This is done so
as to avoid confounding the effects of genetic variation with social
stratification. However, insofar as the geneticists' concept of stratification
(mating which is non-random with respect to genetics) overlaps with the
sociological concept of stratification (a hierarchical ranking),
stratification will predictably associate with causally relevant genetic
variants. For this reason, we would expect principal components to contain real
information about causally relevant variants. So, it is an empirical question whether
controlling for principal components improves or weakens the predictive power of
polygenic scores. Within-family analyses could resolve this question.

In popular media, and sometimes by social scientists, genetics are 
often thought of as an external, "natural" constraint on social structure.
This is one reason why behaviour genetics remains a controversial topic. The
broadest message of our research is that individual genetics are a social
outcome like any other. They can even be viewed as another form of 
capital, alongside human, social and cultural capital: a resource to be
sought, accumulated and competed over. The analysis of this kind of capital
is an exciting area for further research, to which social scientists and
geneticists can contribute.


\FloatBarrier

# Appendix

## Second part of Proposition \@ref(prop:prop1)

\begin{proof}
Write

\begin{equation}
corr(G_{j},S_{j})=\frac{cov(G_{j},S_{j})}{\sqrt{var(G_{j})var(S_{j})}}\textrm{ for both generations }j\in\{p,c\}.\label{eq:corr-cov-var}
\end{equation}

where 

\begin{align*}
var(G_{p}) & =\frac{1}{2}\int g_{d}^{2}+g_{p(d)}^{2}\textrm{d}d;\\
var(G_{c}) & =\int g_{c(d)}^{2}\textrm{d}d.
\end{align*}

Much as before,
\begin{align*}
g_{d}^{2}+g_{m}^{2} & =(g_{c}-\Delta g)^{2}+(g_{c}+\Delta g)^{2}\\
 & =2g_{c}^{2}+2(\Delta g)^{2}\\
 & \ge2g_{c}^{2}.
\end{align*}

This shows that $var(G_{c})\le var(G_{p})$ and a similar argument
shows $var(S_{c})\le var(S_{p})$. Thus the covariance is higher (and
positive) in the children's generation, while the variances are lower.
Combining these ensures that 

\[
corr(G_{c},S_{c}) \ge corr(G_{p},S_{p}).
\]

Since for any $k$, either $var(G_{c}) < var(G_{p})$ or $var(S_{c}) < var(S_{p})$,
the only way to get strict equality for the above is if $k \in \{0,1\}$
and $cov(G_{c}, S_{c}) = cov(G_{p},S_{p}) = 0$.

\end{proof}

To show that the condition in the second part cannot be relaxed further,
consider the distribution in Figure \ref{fig:Correlation-counterexample}.
There is negative correlation in the parents' generation (the shaded
area). If $k = 1$ or is close enough to 1, then assortative mating along the dotted
lines will reduce the variance of $S$ along those lines, pushing the
distribution towards the darker central area, without
affecting the covariance. This will make the correlation more negative.
After repeated generations the horizontal variance within values of $G$ will almost disappear and the correlation will approach -1.

\begin{figure}
\begin{centering}
\includegraphics[width=0.5\columnwidth]{correlation-counterexample}
\par\end{centering}
\caption{Correlation counterexample\label{fig:Correlation-counterexample}}

\end{figure}

## Proposition \ref{prop:robustness}

\begin{proof}

Note that in proposition \ref{prop:prop1}, we took $g_{c(i)} = \bar{g}_{i}$ and $s_{c(i)} = \bar{s}_i$. Write 

\begin{align}
cov(G_{c},S_{c}) & =cov(\bar{G}+\varepsilon^{G},\bar{S}+\varepsilon^{S})\nonumber \\
 & = cov(\bar{G},\bar{S}) + cov(\varepsilon^{G},\bar{S}) + 
     cov(\bar{G},\varepsilon^{S}) + cov(\varepsilon^{G},\varepsilon^{S}).
 \label{eq:cov-with-error}
\end{align}

For any $X$ and $Y$, $cov(X, Y)$ is bounded by $\sqrt{var(X) var(Y)}$.
Plugging $\sigma_{G}^{2}$ and $\sigma_{S}^{2}$ into this formula
shows that under condition 1, $cov(G_{c},S_{c})$ will be arbitrarily
close to $cov(\bar{G},\bar{S})$. Similarly, writing

\[
var(G_{c}) = var(\bar{G}) + var(\varepsilon^{G}) + 2cov(\bar{G},\varepsilon^{G})
\]

shows that $var(G_{c})$ will approach $var(\bar{G})$ as $\sigma_{G}^{2}$
grows small, and similarly for $var(S_{c})$. Plugging these facts
into (\ref{eq:corr-cov-var}) shows that $corr(G_{c},S_{c})$ approaches
$corr(\bar{G},\bar{S})$ as $\sigma_{G}^{2}$ and $\sigma_{S}^{2}$
grow small. Proposition 1 then shows $corr(\bar{G},\bar{S}) < corr(G_{p},S_{p})$
for $k\in(0,1)$.

Under condition 2, $cov(G_{c}, S_{c}) = cov(\bar{G},\bar{S})$ since
the last three terms of the sum in (\ref{eq:cov-with-error}) are
zero. Then since 
\[
cov(\bar{G},\bar{S})\ge cov(G_{p},S_{p}) = 0
\]
with strict inequality iff $k\in(0,1)$, the covariance signs the
correlation. 

\end{proof}

\FloatBarrier

## Robustness checks

```{r tbl-bo-psea-robust}

fml_bo_psea_robust <- fml_bo_psea %>% 
                        purrr::map(update, 
                           . ~ . + 
                           bmi_combined_resid.x + 
                           conscientiousness_resid.x + 
                           neuroticism_resid.x
                        )

mod_bo_psea_robust <-  purrr::map(fml_bo_psea_robust, fixest::feols, 
                         data = mf_pairs_reg,
                         notes = FALSE
                       )

huxreg(mod_bo_psea_robust, 
        coefs = c(
          "Birth order" = "n_older_sibs.x", 
          "University"  = "university.xTRUE", 
          "Income"      = "income_cat.x",
          "Own EA3"     = "EA3.x"
        ),
         note = "{stars}. Standard errors clustered by spouse pair.",
         tidy_args = list(conf.int = FALSE, cluster = mf_pairs_reg$couple_id)
       ) %>% 
       insert_row(after = 9, "Family size dummies", rep("Yes", 4)) %>% 
       insert_row(after = 10, "Polygenic score controls", rep("Yes", 4)) %>% 
       set_top_border(11, everywhere, 0) %>% 
       set_number_format(2:9, -1, 4) %>% 
       set_caption("Regressions of spouse PSEA with controls for polygenic scores")
```



```{r tbl-bo-fath-age} 


age_table <- famhist %>% 
      filter(n_sibs >=2, n_sibs <= 8, ! is.na(n_older_sibs)) %>% 
      group_by(n_sibs, n_older_sibs) %>% 
      dplyr::summarize(
        N   = n(), 
        mfa = mean(fath_age_birth, na.rm = TRUE),
        mma = mean(moth_age_birth, na.rm = TRUE)
      )

# not so helpful if mutation is linear in father's age;
# still significant anyway:
# 
# mf_pairs_young <- mf_pairs_reg %>% 
#                     filter(
#                       moth_age_birth.x <= 30 | fath_age_birth.x <= 30
#                     )
# 
# mod_bo_young <- lapply(fml_bo_psea[1:2], fixest::feols, 
#                  data = mf_pairs_young,
#                  notes = FALSE
#                )

fml_bo_fath_age <- lapply(fml_bo_psea[1:2], update, 
                     . ~ . + fath_age_birth.x, 
                     rhs = 1
                   )
mod_bo_fath_age <- lapply(fml_bo_fath_age, fixest::feols,
                     data  = mf_pairs_reg, 
                     notes = FALSE
                   )

huxreg(mod_bo_fath_age, 
        coefs = c(
          "Birth order"           = "n_older_sibs.x", 
          "University"            = "university.xTRUE", 
          "Father's age at birth" = "fath_age_birth.x",
          "Own EA3"               = "EA3.x"
        ),
         note = "{stars}. Standard errors clustered by spouse pair.",
         tidy_args = list(conf.int = FALSE, cluster = list(mf_pairs_reg$couple_id))
       ) %>% 
       insert_row(after = 9, "Family size dummies", rep("Yes", 2)) %>% 
       set_number_format(2:9, -1, 4) %>% 
       set_caption("Regressions of spouse PSEA controlling for father's age at birth")
```


## Quotations on social inequality and human nature

...your face and figure have nothing of the slave about them, and proclaim you of noble birth.

-- *Odyssey*, Odysseus to Laertes

Citizens, we shall say to them in our tale, you are brothers, yet God has framed you differently. Some of you have the power of command, and in the composition of these he has mingled gold, wherefore also they have the greatest honour; others he has made of silver, to be auxiliaries; others again who are to be husbandmen and craftsmen he has composed of brass and iron; and the species will generally be preserved in the children. But as all are of the same original stock, a golden parent will sometimes have a silver son, or a silver parent a golden son. 

-- Plato *Republic*


Nature would like to distinguish between the
bodies of freemen and slaves, making the one strong for servile labor,
the other upright, and although useless for such services, useful
for political life in the arts both of war and peace. But the opposite
often happens -- that some have the souls and others have the bodies
of freemen.

-- Aristotle *Politics*

Sons have no richer endowment than the quality
A noble and brave father gives in their begetting.

-- Euripides *Heracleidae*


His looks are full of peaceful majesty,
His head by nature fram’d to wear a crown,
His hands to wield a sceptre

-- Shakespeare *Henry VI Part 3*

3. For, when these creatures, being without a king, through fear dispersed in
all directions, the Lord created a king for the protection of this whole (creation),
4. Taking (for that purpose) eternal particles of Indra, of the Wind, of Yama, 
of the Sun, of Fire, of Varuna, of the Moon, and of the Lord of wealth (Kubera).
5. Because a king has been formed of particles of those lords of the gods, he 
therefore surpasses all created beings in lustre...


-- *Laws of Manu*

A daughter of a green Grocer, walks the Streets in London dayly with a baskett
of Cabbage Sprouts, Dandelions and Spinage on her head. She is observed by the
Painters to have a beautiful Face, an elegant figure, a graceful Step and a
debonair. They hire her to Sitt. She complies, and is painted by forty Artists, 
in a Circle around her. The Scientific Sir William Hamilton outbids the Painters, 
Sends her to Schools for a genteel Education and Marries her. This Lady not only 
causes the Tryumphs of the Nile of Copenhagen and Trafalgar, but Seperates Naples
from France and finally banishes the King and Queen from Sicilly. Such is the 
Aristocracy of the natural Talent of Beauty.

-- John Adams to Thomas Jefferson (on Emma Hamilton)
