---
title: |
  Trading social status for genetics in marriage markets: 
  evidence from Great Britain and Norway
author: |
  Abdel Abdellaoui\thanks{Department of Psychiatry, Amsterdam UMC, University 
  of Amsterdam, Amsterdam, The Netherlands. Email: a.abdellaoui@amsterdamumc.nl},
  Oana Borcan\thanks{School of Economics, University of East Anglia, Norwich, 
  UK. Email: O.Borcan@uea.ac.uk}, 
  Pierre-André Chiappori\thanks{Department of Economics, University of Columbia, 
    New York. Email: pc2167@columbia.edu},\
  David Hugh-Jones\thanks{Corresponding author. Email: davidhughjones@gmail.com},
  Fartein Ask Torvik\thanks{Norwegian Institute of Public Health, Oslo. 
    Email: f.a.torvik@psykologi.uio.no} &
  Eivind Ystrøm\thanks{Norwegian Institute of Public Health, Oslo. 
    Email: eivind.ystrom@psykologi.uio.no}
abstract: |
  Under social-genetic assortative mating (SGAM), socio-economic status (SES) and 
  genetically inherited traits are both assets in marriage markets, become
  associated in spouse pairs, and are passed together to future generations.
  This gives a new explanation for persistent intergenerational inequality and
  observed genetic differences between rich and poor. We model SGAM and test
  for it in two large surveys from Great Britain and Norway. Spouses of
  earlier-born siblings have genetics predicting more education. This effect is
  mediated by individuals' own education and income. Thus, shocks to SES are
  reflected in the DNA of subsequent generations. Under SGAM, the distribution of 
  genetic variants in society is endogenous to economic institutions. 
  \par\textbf{Keywords:} Assortative mating, MoBa, UK Biobank.
date: "`r Sys.Date()`"
output: 
  bookdown::pdf_document2:
    toc: false
    latex_engine: xelatex
    number_sections: false
    keep_tex: true
editor_options: 
  chunk_output_type: console
  markdown: 
    wrap: 72
bibliography: bibliography.bib
mainfont: Times
fontsize: 12pt
linkcolor: blue
header-includes:
  - \usepackage{subfig}
  - \captionsetup[subfloat]{labelformat=empty}
  - \captionsetup[figure]{width=5in}
  - \usepackage{setspace}\onehalfspacing
  - \usepackage{amsmath}
  - \usepackage{amsthm}
  - \usepackage{placeins}
  - \usepackage{etoc}
  - \newtheorem{prop}{Proposition}
  - \newtheorem{claim}{Claim}
---


```{r setup, include = FALSE}

library(robomit)
library(Formula)
library(car)
library(drake)
library(Hmisc)
library(dplyr)
library(magrittr)
library(AER)
library(ggplot2)
library(huxtable)
library(broom)
library(fixest)
library(santoku)
library(purrr)
library(scales) # percent should override santoku
library(systemfit)
library(nlWaldTest)

set.seed(27101975)

regression_subset <- function (data) {
  data %>% 
         filter(
           n_sibs.x >= 2, 
           n_sibs.x <= 6, 
           ! is.na(birth_order.x),
           ! is.na(height.x),
           ! is.na(fluid_iq.x),
           ! is.na(university.x),
           ! is.na(bmi.x),
           ! is.na(sr_health.x)
           # we don't demand first_job_pay.x is not NA, that would 
           # shrink the N by a lot
         )
}


calc_prop_shared_children <- function (mf_pairs) {
  drake::loadd(parent_child)
  # all pairs where at least 1 parent has a genetic child in the sample
  # note that this creates multiple parent-child relationships
  mf_w_parent <- mf_pairs %>% 
        left_join(parent_child, by = c("ID.m" = "parent_id"), 
                  relationship = "many-to-many") %>% 
        left_join(parent_child, by = c("ID.f" = "parent_id"), 
                  relationship = "many-to-many",
                  suffix = c(".m", ".f")) %>% 
        filter(! is.na(child_id.m) | ! is.na(child_id.f))
  
  
  n_one_has_kid <- length(unique(mf_w_parent$ID.m)) # using ID.f gives same
  mf_w_both_parents <-  mf_w_parent %>% 
                        filter(child_id.m == child_id.f) # NAs excluded
  n_both_same_kid <- length(unique(mf_w_both_parents$ID.m))
  return(list(one = n_one_has_kid, both = n_both_same_kid))
}


pretty <- function (n, digits = 2, ...) {
  formatC(n, digits = digits, big.mark = ",", ...)
}


update_with_birth_order_dummies <- function (fml) {
  update(fml, . ~ . - birth_order.x + factor(birth_order.x))
}


convert_moba_for_huxreg <- function (tidy, glance) {
  huxtable::tidy_replace(tidy, tidied = tidy, glance = glance)
}


knitr::opts_chunk$set(echo = FALSE)
knitr::knit_hooks$set(
  inline = function (x) {
    if (is.numeric(x)) x <- as.character(round(x, getOption("digits")))
    x <- gsub("-", "\u2212", x)
    paste(as.character(x), collapse = ", ")
  }
)
options(huxtable.long_minus = TRUE)
theme_set(theme_minimal())

drake::loadd(mf_pairs)
drake::loadd(mf_pairs_twice)
drake::loadd(famhist)
drake::loadd(resid_scores)

famhist %<>% left_join(resid_scores, by = "f.eid")
rm(resid_scores)
famhist$EA3 <- famhist$EA3_excl_23andMe_UK_resid

mf_pairs_reg <- regression_subset(mf_pairs_twice)

my_note <- "{stars}. Standard errors clustered by spouse pair in parentheses."

# for p values
my_stars <- c(`***` = 0.001, `**` = 0.01, `*` = 0.05, `+` = 0.10)
# my_stars <- NULL

# basic formulae
fml_bo_psea <- list()

fml_bo_psea[[1]] <- EA3.y ~ birth_order.x | factor(n_sibs.x)
fml_bo_psea[[2]] <- EA3.y ~ university.x + birth_order.x | factor(n_sibs.x)
fml_bo_psea[[3]] <- EA3.y ~ first_job_pay.x + birth_order.x | factor(n_sibs.x)
fml_bo_psea[[4]] <- EA3.y ~ first_job_pay.x + university.x + birth_order.x |
      factor(n_sibs.x)

fml_bo_psea <- lapply(fml_bo_psea, Formula::as.Formula)

# common controls
fml_bo_psea <- lapply(fml_bo_psea, update, 
                        . ~ . + EA3.x + par_age_birth.x)
fml_bo_psea <- lapply(fml_bo_psea, update, 
                        . ~ . | . + factor(YOB.x) + factor(birth_mon.x))

# alternative mediators
fml_bo_psea[-1] <- lapply(fml_bo_psea[-1], update,
                            . ~ . + fluid_iq.x + height.x + bmi.x + sr_health.x)

# basic tidy args
tidy_args <- list(se = "cluster", cluster = "couple_id", conf.int = FALSE)
# tidy_args <- list(se = "hetero", conf.int = FALSE)

reg_coefs <- c(
          "Birth order"           = "birth_order.x", 
          "University"            = "university.xTRUE", 
          "Income"                = "first_job_pay.x",
          "Fluid IQ"              = "fluid_iq.x",
          "Height"                = "height.x",
          "BMI"                   = "bmi.x",
          "Self-reported health"  = "sr_health.x",
          "Own PSEA"              = "EA3.x",
          "Parents' age at birth" = "par_age_birth.x"
        )

moba_reg_coefs <- c(
          "Birth order"           = "parity", 
          "University"            = "university", 
          "Income"                = "incomez",
          "Height"                = "height",
          "BMI"                   = "bmi",
          "Own PSEA"              = "eapgsresid",
          "Parents' age at birth" = "parentalage"
        )
# load moba results
load("moba-results/tradinggenetics_moba_v04.Rdata")
```

```{r TODO-NOTES, eval = FALSE}

* TODO - April 2023

- work on integrating the model with the theory, esp. now we have two societies.
  Can we estimate $a$? And plug in estimates of $\theta$ so as to calculate
  the long-run genes-SES correlation? (And we could also estimate this
  within our data...)
- do something about Norway over time?
- see also narrow TODO list in the Norway data section

# TODO - Apr 2022

* Principal components, own correlation with birth order? Just to check.   Cf. Abdel's earlier critique of GAM papers.
* Versions of first stage with sibling FE? I did this, they are the right
  sign except for BMI, but never significant and with big SEs. Maybe doesn't
  add much.
* Extension with a ne b: does correlation increase in difference between a and b?
  DONE
* Oana, do analysis of missing mediator and how close it would have to correlate
  with univ. attendance to make our results insignificant
* What to do with the Chiappori et al style matching model? Ask Pierre
  DONE
* Comp statics of long run, using Imp Fun Thm;
  solve comp stats and do graphs
* Rework conclusion esp re $\theta$, $\gamma$
    DONE
* $E[x'_2 | x_2]$ and $E[x'_2 | x_1, x_2]$ in prop-gamma, 
  to show that $\sigma$ is a confound while $\gamma$ is a mediator?
* Connect theory to Oana's empirics
* G-E interactions? We mention in intro; could just mention it as a potential
  extension
  DONE


# * Why missing all the first job data? - because not everyone does the
#   online followup. We could use "current job" which might be endogenous
#   to your spouse; then do first job as a robustness check.
#   
# 
# * maybe - consider alternative exogenous shocks to income. For example, some # professions are more "cyclical" than others wrt recessions. If we could do 
# predicted income at age 21-25 from business cycle X profession, that might 
# count as exogenous. (Could use an independent source to estimate evolution
# of incomes, e.g. GHS or BHPS)
# 
# * Maybe: number of elder *brothers*? (We don't have this info but we have
#   total number of brothers, so we can interact this with the birth order
#   effect)
# 


# Check prop 2. (Bob question: if there's noise, then shouldn't
# correlation tend to a limit less than 1? And if so, then at that limit,
# parental correlation should equal child correlation).
# - I think the claim doesn't contradict this. But do double-check!


# TODO: We may have to control, rather than residualize, on PCs
# to get accurate p-values, as our N is not huge by the end (or is 2000 
# enough not to worry)

# Connect the model to existing theoretical models of matching 
# (e.g. Gale-Shapley, Becker 1974, econometrics of matching). DHJ.

# Robustness: look at all 33 polygenic scores for families of size 3. Rerun
# regressions excluding family size of 3. Appendix. DHJ.

# For future: consider gender asymmetries - e.g. does male SES matter more?
# Might need some other "good genes" measures e.g. waist-hip ratio.

# Maybe full set of birth order dummies within family size
# but maybe only for family size 2-4 or so?
# 

# * NB: why in mf_pairs_twice do we see differences in own EA3 with birth_order?
#   - just chance? it's only true for families of size 3...


# NOTES on economics literature
# Areas/topics
# * Econometrics of matching
#   - Choo & Siow 2006
#   - Chiappori et al 'fatter attraction'
#     - Quite simple regressions as a result of their basic framework
#     - Run regressions on each of the other person's characteristics;
#       in the linear version of their framework, coefficients should all be
#       proportional
#   - I'm not sure these guys are our key targets though, more likely some
#     people we might have to "appease"...!
# * Inheritance of inequality
# * Assortative mating
#   - a mechanism for inequality
#   - also related to family economics
#   - this might be a good way to set up the paper:
#   - "We bring together two explanations of persistent inequality. (1) Genetics
#     and (2) assortative mating."
#  * Getting comments from Greg Clark would be very useful.
#  * Maybe from Samuel Bowles too?
#  * They have a Science article on IG Wealth Transmission
#  * This special issue of Current Anthropology is important:
#    https://www.journals.uchicago.edu/toc/ca/2010/51/1
#    - they distinguish "material", "somatic" (inc genetic, but also e.g.
#      embodied knowledge), and "relational" wealth
#    - In their response to comments, Mulder et al. acknowledge that maybe 
#      one kind of wealth can help you acquire other kinds, and this is an
#      important area of future research.
#  * Fernandez et al. (Love and Money)
#    - inequality increases marital sorting because it increases the benefits
#      to matching with another skilled type;
#    - they confirm the correlation cross-country
#    - they also predict sorting lowers GDP (not sure how this works)
#    - there is a link with macroeconomics and the demographic transition in
#      this literature
#  * Eika et al. 2019
#    - Empirics of assortative mating in US from 1940s: it's increased at 
#      the bottom but gone down at the top
#    - Important for inequality but changes have led to little net increase
#      in inequality. (A simple accounting framework when you compare "what
#      if people had matched randomly into households")
#    - The motivation "assortative mating is increasing inequality" is 
#      a sufficient one, for this paper - it's a big issue.
#  * Doepke and Tertilt "families in macroeconomics"
#    - parents' fertility decisions are important for econ growth
#    - as is skill formation
#    - I think this is for another paper - e.g. what happens when the rich
#      have more children?
#  * Gould, Moav and Simhon 2008
#    - to read 
#    - relates end of polygyny to econ growth
#    - as above, maybe for another paper
#      
# PEOPLE you could talk to
# * Doepke
# * Clark
# * Fernandez
# * Bowles
```

\normalem

# Introduction

How families are formed, and transmit traits and assets to their offspring, is
crucial for understanding inequality. Assortative mating in marriage markets can
increase inequality between families [@breen2011educational; @greenwood2014marry]
and contribute to the persistence of inequality across generations, which
is surprisingly high [@clark2015intergenerational;@solon2018we]. Wealthy
families pass on advantages to their children through both genetic inheritance
and environmental influence [@Rimfeld_2018; @bjorklund2006origins; @sacerdote2011nature].

This paper examines a plausible aspect of marriage markets: both social
status and genetics contribute to a person’s attractiveness, and as a result,
genetics and inherited social status may become associated in subsequent
generations.[^status-SES] For example, suppose that wealth, intelligence and
health are advantages in a potential spouse. Then wealthy people are more
likely to marry intelligent or healthy people, and their children will inherit
both wealth, and genetic variants associated with intelligence or health. We
call this mechanism social-genetic assortative mating (SGAM). SGAM may be an
important channel for the transmission of inequality. It creates a genetic
advantage for privileged families, which may help to explain the surprising
long-run persistence of inequality. At the same time, this advantage is not a 
fact of biology, but is endogenous to the social structure. Indeed, under
SGAM, environmental shocks to a person’s social status may be reflected in
the genetics of his or her children.

[^status-SES]: *Social status* refers to characteristics that an individual
  possesses in virtue of their social position. For example, my wealth
  is a fact about me that holds in virtue of my relationship to
  certain social institutions (bank deposits, title deeds et cetera).
  Other examples include caste, class, income, and educational
  qualifications. *Socio-economic status* (SES) is a specific type of
  social status which exists in economically stratified societies, covering
  variables like educational attainment, occupational class, income and 
  wealth [e.g. @white1982relation].


Below, we first write down a theory where attractiveness in the
marriage market is a function of both socio-economic status (SES) and genetic
variants. We show that social-genetic assortative mating in one generation
increases the correlation between SES and genetic variants in the offspring
generation. This result provides a new explanation of the *genes-SES gradient* --
the systematic genetic differences between high- and low-SES people 
[@belsky2018genetic;@Rimfeld_2018;@bjorklund2006origins]. The 
dominant existing explanation for the gradient is meritocratic social mobility: 
if a genetic variant predicts success in the labour market, then it will 
become associated with high SES and will be inherited in high-SES families.
While under meritocracy, genes causes SES, under SGAM causality goes both ways, 
from genes to SES and vice versa. Also, the size of the genes-SES gradient 
depends on economic institutions. Under institutions which increase 
intergenerational mobility, like high inheritance tax rates, the genes-SES 
gradient becomes weaker. On the other hand, an increase in meritocracy can
make the gradient stronger.

Next, using data on spouse pairs from two large genetically-informed surveys in
Great Britain and Norway, we test the hypothesis that an individual’s higher
social status attracts spouses with genetic variants predicting greater educational
attainment. Our genetic measure, the polygenic score for educational attainment
(PSEA), derives from large-scale genome-wide association studies [@lee2018gene;
@okbay2022polygenic]. PSEA reflects a bundle of polygenic effects on underlying 
traits, including intelligence, personality, and physical and mental health
[@demange2021investigating]. PSEA predicts, and causes, educational attainment
itself, as well as intelligence and labour market outcomes. It is already known
that humans mate assortatively on PSEA [@hugh2016assortative;
@robinson2017genetic; @torvik2022modeling], which makes it a likely candidate 
for detecting SGAM.

The endogeneity of socio-economic status is the main challenge in identifying
the effect of SES on the spouse’s genetic endowment. For instance, people with
high educational qualifications tend to also have high PSEA, and as mentioned
above, they may take partners based on genetic similarity. Indeed, recent
studies show strong assortative mating on PSEA, much more than we would expect
if spouses matched only on observed measures of educational attainment
[@okbay2022polygenic]. To isolate the causal link from own SES to partner's
genes, we use a shock to SES which is independent of own genetics. Specifically,
we use a person's *birth order*. Earlier-born children receive higher parental
investment and have better life outcomes, including measures of SES such as
educational attainment and occupational status [@black2011older;
@booth2009birth; @Lindahl_2008]. At the same time, the facts of biology, in
particular the so-called "lottery of meiosis", guarantee that siblings' birth
order is independent of their genetic endowments.[^interaction] Because birth order 
could affect partner choice through both SES and non-SES mechanisms, we run a 
mediation analysis similar to @heckman2013understanding, decomposing the treatment 
effect into effects of measured and unmeasured mediating variables. Specifically, 
we estimate a reduced-form model with spouse polygenic scores for educational
attainment (PSEA) as the dependent variable, and own birth order as the main
independent variable. We then add in to the model measures of own socio-economic
status, including university attendance and income. Under certain assumptions, 
these variables can be interpreted as mediating the effect of birth order on 
spouse genetics.

[^interaction]: @muslimova2020dynamic find that PSEA and birth order
*interact* to produce human capital.

In both Great Britain and Norway, we find that later-born children have spouses
with significantly lower PSEA in the reduced-form regressions. When we add
mediators, including university attendance and/or income, the effect of birth
order shrinks substantially, becoming insignificant in Great Britain, while the
SES mediators significantly increase the spouse’s PSEA. The results are
robust to the inclusion of several controls, including non-SES mediators, and a
rich set of own genetic traits. Thus, SES appears to mediate the effect of birth
order on spouse genetics. The effects of individual mediators differ between the
two countries. While university attendance explains more than a third of the
effects of birth order in both Britain and Norway, income explains about 10% of
the effects in Britain but has little or no independent effect in Norway.
Although our main focus is on testing the basic mechanism of SGAM, this provides
suggestive evidence that in a more egalitarian society, some forms of SES are
less important to the marriage market.

Our paper contributes to several literatures. First, we study assortative mating
between genetic variants and SES. The economics literature on spouse matching
has typically focused on educational similarities [e.g.
@pencavel1998assortative;@chiappori2017partner] or social class or caste [e.g.
@abramitzky2011marrying;@banerjee2013marry], but also sorting based on age,
physical traits and ethnicity [@hitsch2010matching]. Matching decisions on the
marriage market follow multiple criteria, with some degree of substitutability
between them.[^others] For instance, @chiappori2012fatter showed that
individuals trade off BMI for partners’ income or education and that the
marginal rate of substitution between these characteristics is different for men
and women.

[^others]: @oreffice2010anthropometry show that height and BMI are associated 
with spouse earnings. @dupuy2014personality find spouse matching on multiple 
independent dimensions, including education, height, BMI and personality.
@chiappori2021analyzing analyse matching on multiple characteristics
and show that a three-dimensional matching model fits their data.

In genetics, @cloninger1979multifactorial present a model of genetic and cultural
transmission, where assortative mating is based directly on phenotype.
Sociocultural values are transmitted from parents. Assortative mating based on
phenotype homogamy, modeled simply as a correlation coefficient, leads to a
positive cultural-genetic association in offspring. The model estimates the
resemblances between relatives, which are identified through maximum
likelihood estimation on linear models, and therefore the analysis is
correlational. Our theory is similar in that we also assume the function on
which assortative mating occurs is additive in genetic values and
transmissible socioeconomic status (e.g. wealth). @heath1985resolving draw
inspiration from earlier papers [@rao1976resolution; @rao1979path] to introduce
assortative mating by social background (i.e., belonging to the same groups,
such as attending the same church or workplace).
@otto1995genetic extend assortative mating to include both phenotypic and social
homogamy. Early empirical work used structural models to estimate parameters from
correlations between twins and spouses of twins[@eaves1999comparing]. Since the
"genomics revolution", which has made individuals' DNA data directly available,
empirical research has focused on genetic assortative mating (GAM), the
phenomenon that people with similar genes marry each other. Recent research has
confirmed the long-standing conjecture that GAM takes place in contemporary human
populations [@hugh2016assortative; @robinson2017genetic; @howe2019genetic;
@torvik2022modeling]. Other work has analysed cross-trait assortative mating, 
when e.g. people with genes for height marry people with genes for intelligence
[@Beauchamp_2010; @Sundet_2005; @border2022cross]. As a result, the two types of 
genetic variation become associated. 
 
Here, we provide a microeconomic foundation for of assortative mating. The
matching process is based on maximising surplus from the match, which is
determined by both genetics and culturally heritable SES. Past papers using
additive models of phenotype determination have not studied the implications for
the variance of socioeconomic characteristics, i.e. economic inequality, in
future generations. In particular, our model shows how
the genes-SES gradient is affected by economic institutions (see below).
Our second key innovation is to use the "shock" of birth order to provide
evidence of substitutability between SES and genes. This gives a clean demonstration of an environmental input to assortative mating.
Note that our econometric identification remains agnostic on the precise
mechanism of assortative mating: individuals can trade SES for genes because they
meet in groups determined jointly by genes and SES [in the spirit of
@otto1995genetic], or they can match on phenotypes that are determined jointly by
genes and SES.


SGAM may affect economic inequality and intergenerational mobility.
Intergenerational mobility correlates with cross-sectional inequality
[@becker2018theory;@krueger2012rise], which has risen dramatically in
high-income countries, while intergenerational absolute mobility has declined
[@western2008inequality;@chetty2017fading].[^stats] Long-run estimates of 
intergenerational persistence of wealth are often surprisingly high, i.e.
higher than would be predicted by parent-child correlations in wealth
[@clark2015intergenerational; @barone2021intergenerational;
@solon2018we]. @clark2023inheritance argues that this could be explained by an
underlying process where unobserved genetic variation determines wealth. We 
show below that SGAM could also generate these patterns. The mechanism again is
unobserved genetic variation, but the interpretation is different, since genetic
endowments are not an exogenous source of variation, but an asset "traded" in
marriage markets in exchange for wealth and social status. Put another way, in
analysing the effect of ancestral wealth on descendant wealth, genetics need
not be a *confound*; under SGAM, it may be a *mediator*.

[^stats]: Though relative mobility has been stable [@chetty2014united]. In the 
United Kingdom the Gini coefficient has increased from 26% to 34.6% between 1977 
and 2020. The United States has seen a 10 percentage point rise to 43.3% 
from 1962 to 2013.

SGAM also affects cross-sectional inequality, like other forms of assortative
mating [@fernandez2001sorting;@fernandez2005love;@eika2019educational;
@chiappori2018marriage]. In particular, it increases cross-sectional genetic
inequality. Multiple lines of evidence show that there is a
"genes-SES gradient": many genetic variants are associated with high or low SES.
From twin studies, the heritability of occupational class and educational
attainment, i.e. the proportion of variance explained by genetic differences
between individuals, is around 50% [@Tambs_1989]. Genome-wide Complex Trait
Analysis (GCTA) shows that the family socio-economic status of 2-year-old
children can be predicted from their genes [@Trzaskowski_2014]. Children born
into higher-income families have more genetic variants predicting educational
attainment [@belsky2018genetic]. Studies comparing adoptees to non-adoptees show
that both post-birth environment and pre-birth conditions (genetics and to a
lesser extent prenatal environment) contribute to the transmission of wealth and
human capital [e.g.
@bjorklund2006origins]. Thus, the genes-SES gradient is an important source of
inequality. Genetic variation in human capital is key, since a likely cause of
the recent rise in inequality is the increase in market returns to human capital
[e.g. @kaplan2013s; @eika2019educational]. There is also a genes-SES gradient in
genetic predictors of health. DNA-derived scores predictive of several health
outcomes are associated with regional economic deprivation
[@abdellaoui2019genetic]. The correlation between education and health 
may be mediated by shared genetic causes [@amin2015schooling;
@boardman2015can]. Family SES correlates with several health-related polygenic
scores [@selzam2019comparing], and genetic variants associated with SES
significantly alter the genetic relationships between many mental health
outcomes [@marees2021genetic].

SGAM shows how marriage markets can lead high SES to be associated with
different genetic variants, i.e. it can explain the genes-SES gradient. The
standard explanation for the genes-SES gradient is returns to human capital 
in labour markets, also known as meritocracy. Parents with higher ability reap
higher market returns, and they may then pass both higher socio-economic status
and their genes to their children, leading to an association between the two
[@belsky2018genetic]. This mechanism depends on the level of meritocracy in
social institutions [@branigan2013variation; @Heath_1985]: in a society where
social status was ascribed rather than earned, it could not take effect. Indeed,
after the fall of communism in Estonia, the heritability of SES increased,
presumably because post-communist society allowed higher returns to talent
[@Rimfeld_2018]. By contrast, SGAM does not require meritocracy. Even when
social status is entirely ascribed, it can still become associated with certain
genetic variants, so long as their associated phenotypes are prized assets in
marriage markets. Since meritocracy is historically rare, while assortative
mating is universal, this suggests that genes-SES gradients are likely to be
historically widespread.

Both meritocracy and SGAM may increase social inequality overall, if there are
complementarities between genetic and environmental components of human capital, 
for example if higher-ability parents make more productive investments in
children’s human capital [@cunha2007technology; @cunha2010estimating;
@heckman2014economics; @kong2018nature], or if high-income parents are able to
invest more in transmitting their human capital [@becker2018theory]. Thus, by
bringing "good genes"[^language] and enriched environments together, SGAM may
increase inequality in the next generation.

[^language]: Throughout, "good genes" simply means genetic variants
associated with characteristics considered desirable in a society's marriage
markets. No normative valuation is implied.

Lastly, we contribute to a literature in economics that examines the
relationship between genetic and economic variables. @benjamin2011promises is an
early review. Several recent papers use polygenic scores, in particular
polygenic scores for educational attainment [@barth2020genetic;
@papageorge2020genes; @ronda2020family]. @barban2021effect use PSEA as an
instrument for education in a marital matching model. These papers --
like the vast majority of the behavior genetics literature [see e.g.
@plomin2008behavioral] -- take genetic endowments as exogenous and examine how
they affect individual outcomes, perhaps in interaction with the environment. We
take a different approach by putting genetics on the "left hand side". Thus, our
paper challenges the assumption, in economics and beyond, that genetic endowment
is exogenous to economic characteristics. While this may be tenable in
within-generation studies, it ceases to hold in intergenerational models.
Social-genetic assortative mating is a causal mechanism going from
socio-economic status to genes for heritable traits.

Our model also shows that the strength of this mechanism is endogenous to
social and economic institutions. When SES is highly transmissible across the
generations, this has the long-run effect of increasing the association between
SES and genetics. If so, institutional reforms that increase *intergenerational
mobility*, like mass education or inheritance taxation, may in the long run
affect not only economic but genetic inequality. Conversely, an increase in
*economic meritocracy* increases the association between SES and genetics in the
long run.[^prop] This poses the problem raised by @young1958rise, and more
recently by @markovits2019meritocracy: meritocracy may be self-limiting or even
self-undermining.

[^prop]: See Proposition \@ref(prop-gamma) below.

The observations behind SGAM are not new. That status and physical
attractiveness assort in marriage markets is a commonplace and a perennial theme
of literature. In the Iliad, powerful leaders fight over the beautiful
slave-girl Bryseis. In Jane Austen’s novels, wealth, attractiveness and “virtue”
all make a good match. @marx1844economic wrote “the effect of ugliness, its repelling
power, is destroyed by money.” The literature on mate preference from evolutionary
psychology [@buss1986preferences; @buss1989sex; @buss2019mate] confirms
that attractive mate characteristics include aspects of social status (“high
earning capacity,” “professional status”) as well as traits that are partly
under genetic influence (“intelligent,” “tall,” “kind,” “physically
attractive”). Despite this, to our knowledge, ours is the first work to examine
assortative mating between SES and genetics, and to model its social and
economic consequences.[^prev-work]

[^prev-work]: In addition to the work mentioned above on cultural and 
  genetic inheritance, @halsey1958genetics showed in a two-class model that 
  social mobility combined with assortative mating might increase the association 
  between genetics and social class. @belsky2018genetic offer three reasons for 
  the association between education-linked genetics and SES, but do not consider 
  SGAM.


```{r literature-notes, eval = FALSE}

# Notes 

# Fernandez et al 2005
# - skills are caused by investment in human capital
# - sorting is inefficient since it increases inequality of parents' capital
# and some people will be credit-constrained.
# 
# Marry Your Like (Greenwood 2014)
# - simple accounting methodology
# - ass mating has increased and has contributed to increase in inequality in the 
#   US, 1960 to 2005
#   
# Eika 2019
# - assortative mating has decreased among highly educated, increased among low
#   educated, 1960s-2010s. Overall, increase only until the 1980s
# - changes in ass mating barely move time trends in household income inequality
# - findings for US + 4 European countries

# Clark 201x - surname groups show higher intergen transmission of wealth 
# than 
# individual parent-child pairs. Also grandparents matter independently. 
# One explanation: underlying "social status" or "social competence" which
# is measured with error by wealth.

# Solon 2018 - reviews evidence on long-run transmissibility. Not
# much evidence for Clark's "0.7". Many alternative explanations.

# Becker-Tomes 1979?
# 

# Natural/cultural selection models

* Otto et al.
  - derive models with cultural and genetic inheritance, plus
    A.M. on the phenotype or on the underlying cultural type;
  - nb one thing we maybe add is a microfoundation in terms of indiv. choice..
  - metaanalysis of various parent/child correlations on IQ variation
    - so, no research design ish elements...

* Rao, Morton and Yee 1976
  - path analysis allowing for spousal correlations, both "E-E" and "G-E"
  - then fit parameters, again no research design;
  - I think there's room to argue that we add something simply by
    the use of the exog. variation

* Cloninger et al. 1979
  - path analysis; cultural transmission can vary btw mothers and fathers
  - no microfoundations

* Fisher 1918
  - mentions ass. mating though not the cultural kind
  - but does dist btw mating on phenotype and on underlying genetics
  
Others to look at:

* Beauchamp et al. 2011 
  - looks at ass. mating w.r.t. height/intelligence correlation
  - cross trait a.m. but both genetic

* Vinkhuyzen et al. 2010
  - A.M. of phenotypic/social types
  - effects on heritability of intelligence estimates
  - good for a sense of "state of the art"
  
* Heath and Eaves 1985
  - how to distinguish between phenotypic and social A.M.
  - a "classic"
```


# Model

```{r model-section, child="model-section.Rmd"}

```


# Data and methods


The central insight in our model is that higher SES and good genes assort in the 
marriage market. In particular, own SES and spouses' genes can become positively
correlated even if own genes and SES are not correlated (Proposition 1). In what
follows, we test for correlations among spouses between one partner's SES 
and the other partner's genetics. 

We use data from two sources: Great Britain and Norway. Doing so allows us to
check our basic result in two different societies, and also to make some 
(admittedly tentative) comparisons between them. Our Great Britain data
comes from the UK Biobank, a study of about 500,000 individuals born between
1935 and 1970 [@bycroft2018uk]. The Biobank contains information on respondents'
genetics, derived from DNA microarrays, along with questionnaire data on health
and social outcomes. The Biobank does not contain explicit information on spouse
pairs. We categorize respondents as pairs if they had the same home postcode on at least 
one occasion;[^4] both reported the same homeownership/renting status, length of time
at the address, and number of children; attended the same UK Biobank assessment 
center on the same day; both reported living with their spouse ("husband, wife or
partner"); and consisted of one male and one female. 
We also eliminate all pairs where either spouse appeared more than once
in the data. This leaves a total of `r pretty(nrow(mf_pairs))` pairs.[^matching]

[^4]: A typical UK postcode contains about 15 properties.

[^matching]: In the appendix, we test the validity of our matching process by 
counting the proportion of pairs who had a shared genetic child, in a subsample
of the data. We also check whether any misidentified pairs might have biased our 
results, by constructing a dataset of "known fake pairs".

Our Norway data comes from the Norway Mother, Father and Child Cohort Study
(MoBa), a population-based study of pregnant women and their partners and children
[@magnus2016cohort; @paltiel2014biobank]. Participants were recruited from all
over Norway from 1999-2008. 41% of women consented to participation. In this 
paper, we use about 100,000 genotyped individuals and about 45,000 genotyped 
spouse pairs. The Norway data has some advantages over UK Biobank, including higher
participation, larger sample size, and spouse pairs which are known rather than
inferred. On the other hand it is missing some variables, including IQ measures
and self-reported health. 

Our key dependent variable is spouse's *Polygenic Score for Educational
Attainment* (PSEA). A polygenic score is a DNA-derived summary measure
of genetic risk or propensity for a particular outcome, created from
summing small effects of many common genetic variants, known as Single
Nucleotide Polymorphisms (SNPs). We focus on PSEA rather than other
polygenic scores for two reasons. First, educational attainment plays a key role in
human mate search. People are attracted to educated potential partners
[@buss1986preferences; @belot2013dating]; spouse pairs often have
similar levels of educational attainment, as well as similar PSEA
[@vandenberg1972assortative; @schwartz2005trends; @greenwood2014marry;
@hugh2016assortative; @torvik2022modeling]. Second, PSEA predicts a set of 
important socioeconomic variables, including not only education but also social 
and geographic mobility, IQ, future income and wealth [@belsky2016genetics; @barth2020genetic; @papageorge2020genes].[^discuss]

[^discuss]: See @papageorge2020genes for a detailed discussion of polygenic 
scores aimed at economists.

```{r calc-psea-check}

mod_own_psea <- feols(university ~ EA3, data = famhist, note = FALSE)
mod_own_psea_within <- feols(university ~ EA3 | sib_group, data = famhist,
                               note = FALSE)

n_own_psea <- glance(mod_own_psea)[["nobs"]]
n_own_psea_within <- glance(mod_own_psea_within)[["nobs"]]

r2_own_psea <- glance(mod_own_psea)[["r.squared"]]
r2_own_psea_moba <- 0.08136 # calculated by Fartein

mod_own_psea <- tidy(mod_own_psea)
mod_own_psea_within <- tidy(mod_own_psea_within)

eff_own_psea  <- mod_own_psea[[2, "estimate"]]
pval_own_psea <- mod_own_psea[[2, "p.value"]]

eff_own_psea_within  <- mod_own_psea_within[[1, "estimate"]]
pval_own_psea_within <- mod_own_psea_within[[1, "p.value"]]

stopifnot(pval_own_psea < 2e-16)
stopifnot(pval_own_psea_within < 2e-16)

n_with_job_codes <- sum(! is.na(mf_pairs_reg$first_job_pay.x))
```


PSEA in the UK was calculated using per-SNP summary statistics from @lee2018gene,
re-estimated excluding UK Biobank participants; in Norway, using statistics 
from @okbay2022polygenic. The score was normalized to
have mean 0 and variance 1. Because polygenic scores are created from estimates
of many small effects, they contain a large amount of noise relative
to the true best estimator that could be derived from genetic data. For
instance, PSEA explains only 11–13% of variance in educational attainment 
[@lee2018gene], whereas the true proportion explained by genetic
variation -- the heritability -- is estimated from twin studies to be about 40%
[@branigan2013variation]. Also, polygenic scores are no more guaranteed
to be causal than any other independent variable. For example, social
stratification by ancestry may lead genes to be associated with educational
attainment even if they play no causal role [@selzam2019comparing].

Despite these points, PSEA has non-trivial estimated effects on educational
attainment. PSEA correlates with measures of education, including university
attendance and years of full-time education. Effect sizes are smaller but still
non-trivial in within-siblings regressions [@lee2018gene], where they can be
interpreted as causal, since genetic variation across siblings is guaranteed to
be random by the biological mechanism involved -- the "lottery of meiosis" (see
below). We recheck these facts within the UK Biobank sample. In a simple linear
regression (N = `r pretty(n_own_psea)`) of university attendance on PSEA, a
one-standard-deviation increase in PSEA was associated with a 
`r pretty(eff_own_psea*100)` percentage point increase in the probability of
university attendance ($p < 2 \times 10^{-16}$). In a within-siblings regression
among genetic full siblings (N = `r pretty(n_own_psea_within)`), the increase
was `r pretty(eff_own_psea_within * 100)` percentage points ($p < 2 \times
10^{-16}$). This suggests that about half of the raw
correlation of PSEA with university attendance is down to environmental
confounds like parental nurture, while the remainder is causal [cf. @lee2018gene]. Still, the
causal effect remains substantial: for a rough comparison, the (ITT) effect on
college attendance of the Moving To Opportunity experiment in the US was 2.5
percentage points [@chetty2016effects].

We use two measures of socio-economic status: income, and university attendance.
Income is a direct measure of SES. University attendance is a predictor of
income over the whole life course, and a form of SES in itself. The MoBa data
includes both university attendance and income. UK Biobank includes university
attendance, but only has a direct measure of current household income, which is
inappropriate for our purposes because it includes income from both spouses and
is measured after marriage. Instead, we estimate income in £000s in the
respondent's first job, by matching the job's Standard Occupational
Classification (SOC) code with average earnings by SOC from @ONS2007ASHE. Job
codes are only available for a subset of respondents among our pairs.

Figure \@ref(fig:pic-basic-corr) illustrates the core idea of SGAM within the UK
Biobank data. The X axis shows a measure of one partner's socio-economic status:
university attendance or income. The Y axis plots the other partner's mean PSEA.
Both males and females who went to university had spouses with higher PSEA. So
did males and females with higher income in their first job. Since DNA is
inherited, these people's children will also have higher PSEA.[^moba-pic]

[^moba-pic]: Figure \@ref(fig:pic-basic-corr-moba) in the appendix shows
the same plot for the MoBa sample.

```{r pic-basic-corr, fig.cap = "Spouse polygenic score for educational attainment (PSEA) against own university attendance and own income in first job (Great Britain). Lines show 95\\% confidence intervals. PSEA is normalized to have mean 0 and variance 1. Income is estimated from the respondent's first job, as the average income of the SOC job code.", fig.subcap = rep("", 4), fig.align = "center", fig.width = 3, fig.height = 3, fig.ncol = 2}

pic_ss <- stat_summary(fun.data = mean_cl_normal, na.rm = TRUE)
pic_cc <- coord_cartesian(ylim = c(-0.15, 0.25))
pic_theme <- theme(axis.title = element_text(size = 10), 
                   panel.grid.minor.x = element_blank(),
                   panel.grid.major.x = element_blank())
mf_pairs %>% 
      filter(! is.na(university.m)) %>% 
      mutate(University = ifelse(university.m, "Yes", "No")) %>% 
      ggplot(aes(University, EA3.f)) + 
        pic_ss + 
        pic_cc + 
        pic_theme +
        labs(x = "Male spouse university attendance", y = "Female spouse PSEA")
        

mf_pairs %>% 
      filter(! is.na(university.f)) %>% 
      mutate(University = ifelse(university.f, "Yes", "No")) %>% 
      ggplot(aes(University, EA3.m)) + 
        pic_ss + 
        pic_cc +
        pic_theme +
        labs(x = "Female spouse university attendance", y = "Male spouse PSEA")


pic_ss <- stat_summary(
            fun.data = mean_se, 
            fun.args = list(mult = 1.96), 
            na.rm = TRUE
          )
pic_cc <- coord_cartesian(ylim = c(-0.15, 0.25))

mf_pairs %>% 
      filter(! is.na(first_job_pay.m)) %>% 
      mutate(Income = santoku::chop_deciles(first_job_pay.m, labels = 1:10)) %>% 
      ggplot(aes(Income, EA3.f)) + 
        pic_ss + 
        pic_theme +
        labs(x = "Male spouse income decile", y = "Female spouse PSEA")

mf_pairs %>% 
      filter(! is.na(first_job_pay.f)) %>% 
      mutate(Income = santoku::chop_deciles(first_job_pay.f, labels = 1:10)) %>% 
      ggplot(aes(Income, EA3.m)) + 
        pic_ss + 
        pic_theme +
        labs(x = "Female spouse income decile", y = "Male spouse PSEA")
```


These plots do not prove that SGAM is taking place. Since an
individual's own PSEA correlates with both their educational attainment,
and their income, both figures could be a result of genetic assortative
mating (GAM) alone [@hugh2016assortative]. Indeed, recent studies show much
higher levels of GAM than could be explained by matching on the observed 
education phenotype alone [@okbay2022polygenic]. So, to demonstrate SGAM, we need
a source of social status which is exogenous to genetics. Also, the link
between social status and spouse genetics is likely to be noisy, for
three reasons: first, polygenic scores contain a large amount of error, as
discussed above; second, causal mechanisms behind variation in social status 
are likely to be noisy; third, to paraphrase @shakespeare1595midsummer, the 
spouse matching process is highly unpredictable. So, we need a large N to
give us sufficient power. This rules out time-limited shocks such as
changes to the school leaving age [@Davies_2018].

We use *birth order*. It is known that earlier-born children receive more
parental care and have better life outcomes, including measures of SES such as
educational attainment and occupational status [@Lindahl_2008; @booth2009birth;
@black2011older].[^literature] On the other hand, all full siblings have the
same *ex ante* expected genetic endowment from their parents, irrespective of
their birth order. This is guaranteed by the biological mechanism of meiosis,
which ensures that any gene is transmitted from either the mother or the father
to the child, with independent 50% probability
[@mendel1865experiments;@lawlor2008mendelian]. For example, siblings' expected
polygenic score is equal to the mean of their parents' polygenic scores.[^endog]
We can therefore use birth order as a "shock" to social status. "Shock" is in
quotes because we do not claim that birth order is exogenous to all other
variables. For example, it naturally correlates with parental age, and it may
also correlate with household SES at the time of birth. We only claim that birth
order is exogenous to genetic variation.

[^literature]: Earlier work was ambiguous on the effects of birth order 
  [e.g. @hauser1985birth;@hanushek1992trade]. However, this work often used
  unrepresentative samples and/or did not control for family size or parental age.
  More recent work improves on this and shows clear birth order effects.
  @kantarevic2006birth show that parental age is an important confound for birth
  order. @black2005more show substantial birth order effects in the whole Norwegian
  population, even in a family fixed-effects specification, and after controlling
  for mother's age. @booth2009birth examine UK families, controlling for family
  size and for parental age at birth, and show significant and substantial birth
  order effects on education. Some studies [e.g. @black2005more; @de2010birth] use
  twin births or the gender mix of children as instruments for family size. They
  too typically find that birth order has large negative effects. 

[^endog]: Although genetic variation is randomly assigned to children at
    birth, genetics and birth order could be dependent if parents'
    choice of whether to have more children is endogenous to the genetic
    endowment of their earlier children. We check for this below. 
    @isungset2021birth also find that birth order differences in education are
    not genetic.

Although birth order within a given family is
independent of genetics, birth order in the whole sample might be correlated with
other environmental and genetic factors. In particular, birth order naturally
correlates with family size, i.e. the total number of siblings born, because
e.g. third-born children must be in a family of at least three siblings. It
also correlates with parental age, which may affect the income and
maturity of the parents. To control for family size, we use dummies for
each possible value of family size.  In most regressions, we use only
respondents with between 1 and 5 siblings, i.e. with a family size of
2-6: beyond that, estimation would get noisy because of small cell sizes.
To control for parental age, we use father's and/or mother's age at the
respondent's birth. This is calculated from the relevant
parent's current age. In the UK Biobank, this data is only available if the
respondent's parent was still alive. For other controls we use respondent's
month of birth; age at interview; and own PSEA. Our claim is that controlling
for these variables, birth order is independent of genetic variation. 


## Decomposing the birth order effect on spouse genetics

Birth order offers a way to test the central assumption in our theoretical model -- that SES is traded for genes in the marriage markets -- by providing a shock to SES
which is exogenous to own genetics. Ideally, we might prefer to use birth order 
as an instrument for SES. However, our measures of social status are noisy and
incomplete. For example, we know whether subjects attended university, but not 
which university. Birth order likely affects both measured and unmeasured aspects 
of SES. So, an instrumental variables approach would fall foul of the exclusion
restriction.

Instead, we conduct a mediation analysis, following the strategy of
@heckman2013understanding. We first confirm that birth order affects our
measures of respondents' SES (income and education). Then, we regress spouse's
PSEA on birth order, with and without controlling for SES. Under the assumption
that birth order is exogenous to own genetics, these regressions identify the
effect of birth order, plus other environmental variables that correlate with
it, on own social status and spouse's genetics. Most importantly, if the 
estimated effect of birth order on spouse's PSEA changes when SES is controlled 
for, that is evidence that SES mediates the effect of birth order.

We follow @heckman2013understanding to decompose the aggregate treatment
effect into components due to observed and unobserved proximate channels
affected by the treatment. Our aim is to estimate the effect of SES (as
an effect of birth order) on spouse PSEA.

Assume $B$ is a multivalued variable indicating birth order. Let $Y_b$ be
the counterfactual outcome (spouse PSEA) for the first-born, second-born etc.
Given $b$, spouse PSEA is assumed to be independent across observations
conditional on some predetermined controls which are assumed not to be affected
by $B$.

Let $m_{b}$ be a set of mediators, i.e. proximate outcomes
determined by $b$, which account (at least in part) for the $b$
treatment effect on spouse PSEA. We can think of $m_{b}$ as all
the effects on attractiveness, such as increments to SES, health,
cognitive and non-cognitive skills, that individuals receive due to
their birth rank. We can split the mediators in $m_b$ into a set $J_m$ of
measured mediators, including university attendance and income in first job, 
and a set $J_u$ of mediators that we cannot measure. 

Our linear model is:

\begin{equation}
\label{eq:linear-model}
Y_b = \kappa_b + \sum_{j \in J_m} \alpha_b^j m^j_b+\sum_{j \in J_u} \alpha_b^j m^j_b + \mathbf{X^\prime} \symbf{\beta_b} + \tilde{\varepsilon}_b = \tau_b + \sum_{j \in J_m} \alpha_b^j m^j_b + \mathbf{X^\prime} \symbf{\beta_b} + \varepsilon_b 
\end{equation}


where $\tilde{\varepsilon}_b$ is a mean-zero residual assumed independent of
$m_b$ and $\mathbf{X}$; 
$\tau_b = \kappa_b + \sum_{j \in J_u} \alpha_b^j E(m^j_b)$; and 
$\varepsilon_b = \tilde{\varepsilon}_b + \sum_{j \in J_u} (m^j_b - E(m^j_b))$.
We simplify by assuming that $\beta_b = \beta$
and $\alpha_b = \alpha$ for all $b$, i.e. that the effects of $\mathbf{X}$ and
$m_B$ don't differ by birth order.[^assumptions] We assume differences in
unmeasured investments due to $b$ are independent of $\mathbf{X}$.

[^assumptions]:Under the assumption that measured and unmeasured mediators are
uncorrelated, we can test these assumptions by running an OLS regression of an
extended model \@ref(eq:model-to-estimate) where we interact the measured
mediators and controls with the treatment $B$, and test the significance of the
coefficients on the interaction terms ($\symbf{\alpha_b}=0$ and
$\symbf{\beta_b}=0$). See @heckman2013understanding and @fagereng2021wealthy 
for details and different applications. When we run the model with interactions,
only one interaction is significant after Bonferroni correction at $p < 0.05/34$:
the interaction of income in first job with the dummy for birth order 6. So 
overall, the uninteracted model seems a good enough approximation.

We use a linear model for each observed mediator variable:
\begin{equation}
\label{eq:mediator-model}
m^j_b=\mu_{0,j}+\mathbf{X^\prime}\symbf{\mu_{1,j}}+\mu_{2,j}\cdot b+\eta_j, j \in J_m  
\end{equation}
where \(\eta_j\) is a mean-zero residual. We also assume the treatment-specific 
intercepts are linear in $b$:
\begin{equation}
\label{eq:linear-intercept}
\tau_b=\tau_{0}+\tau b.  
\end{equation}

With the simplifying assumptions above and substituting \@ref(eq:mediator-model)
and \@ref(eq:linear-intercept) into \@ref(eq:linear-model) we obtain:
\begin{equation}
\label{eq:simplified-model}
Y_b = \tau_0+\tau b + \sum_{j \in J_m} \alpha^j (\mu_{0,j}+\mathbf{X^\prime}\symbf{\mu_{1,j}}+\mu_{2,j}\cdot b+\eta_j) + \mathbf{X^\prime} \symbf{\beta} + \varepsilon_b 
\end{equation}

Using equation \@ref(eq:simplified-model), we can decompose the average treatment 
effect of a change from birth order $b$ to $b'$ into the effect of measured mediators 
$m^j$ and unmeasured mediators on the outcome:

\begin{equation}
\label{eq:decomposition}
E(Y_b' - Y_b) = \tau(b' - b) + \sum_{j \in J_m} \alpha^j E(m^j_{b'} - m^j_b) \\
=\underbrace{\tau(b' - b)}_{\text{Direct effect + unmeasured mediators}} + \underbrace{\sum_{j \in J_m} \alpha^j \mu_{2,j} (b' - b)}_{\text{Effect of measured mediators}}
\end{equation}

We are primarily interested in estimating the effect of SES on spouse PSEA,
amongst the measured mediators, and furthermore we would like to measure the
relative importance of SES compared to other factors in predicting spouse PSEA.

We therefore estimate: 
\begin{equation}
\label{eq:model-to-estimate}
Y = \tau_0 + \tau B + \sum_{j \in J_m} \alpha^j m^j_b + \mathbf{X'} \symbf{\beta} + \varepsilon
\end{equation}


Estimating the above by OLS will generate unbiased estimates of
$\alpha^j$ if $m^j$ is measured without error and is
uncorrelated with the error term $\varepsilon$. Since
$\varepsilon$ contains both individual disturbances and differences
in unmeasured investments due to birth order, there are two identifying assumptions that need to hold for unbiased OLS estimates: (a) the measured investments
(specifically SES) should be independent of unmeasured investments
generated by birth order. Failing this, the estimates of $\alpha^j$ will be
conflated with the effects of unmeasured investments. Second, (b) the measured
investments should be uncorrelated with other shocks $\tilde{\varepsilon}_b$.
With respect to assumption (a), in our regressions we control for a set of
potential alternative mediators available in the data: height,
BMI, and (in UK Biobank only) fluid IQ and self-reported general health.  
With respect to assumption (b), we use further controls, such as parental age, 
year of birth, and own PSEA, to reduce unobserved variation in the error term. 

By running a least square regression of \@ref(eq:model-to-estimate), we can
estimate $\tau$ and $\alpha^j$. If assumption (a) holds, the part of the birth
order treatment effect on spouse PSEA that is due to measured mediators,
including SES, can be constructed using the estimated $\alpha^j$ and the effects
of birth order on measured mediators. We can estimate these effects in a second
step, from OLS regressions based on equation \@ref(eq:mediator-model) for each
measured mediator (in particular university attendance and income) on
$\mathbf{X}$ and $B$. The part of the birth order effect that is due to
university attendance (or income) on spouse PSEA will be the coefficient of
university/income in the regression of spouse PSEA in equation
\@ref(eq:model-to-estimate), multiplied by the coefficient of birth order on
university/income from equation \@ref(eq:mediator-model). We now apply this
framework to each of our two samples.

# Results: Great Britain


```{r tbl-bo-first-stage}
fml_bo_1st <- list()
fml_bo_1st[["University"]] <- update(fml_bo_psea[[1]], university.x ~ . )
fml_bo_1st[["Income"]] <- update(fml_bo_psea[[1]], first_job_pay.x ~ .)
fml_bo_1st[["Fluid IQ"]] <- update(fml_bo_psea[[1]], fluid_iq.x ~ . )
fml_bo_1st[["Height"]] <- update(fml_bo_psea[[1]], height.x ~ .)
fml_bo_1st[["BMI"]] <- update(fml_bo_psea[[1]], bmi.x ~ .)
fml_bo_1st[["Health"]] <- update(fml_bo_psea[[1]], sr_health.x ~ .)

mod_bo_1st <- purrr::map(fml_bo_1st, fixest::feols, data = mf_pairs_reg, 
                           notes = FALSE)

coef_bo <- purrr::map(mod_bo_1st, tidy) %>%
             purrr::map(filter, term == "birth_order.x") %>% 
             purrr::map_dbl("estimate")

tbl_note <- paste("Estimates from OLS regressions with the ",
"mediators (university attendance, income, fluid IQ, height, BMI, self-reported",
"health) as dependent",
"variables, and own birth order as the main independent variable. PSEA is the",
"polygenic score for educational attainment, which is normalized with mean 0 and",
"standard deviation 1. We include parents' age at birth (the mean of parents'", 
"ages) and further controls to ensure the balance of covariates across birth", 
"order. All data is from the UK Biobank for a sample of UK individuals born", 
"between 1935 and 1970.", my_note)

huxreg(mod_bo_1st, 
         coefs = c(
           "Birth order" = "birth_order.x", 
           "PSEA" = "EA3.x",
           "Parents' age at birth" = "par_age_birth.x"
         ),
         statistics = c(
           "N"  = "nobs", 
           "$R^2$" = "r.squared"
          ),
         note = tbl_note,
         stars = my_stars,
         tidy_args = tidy_args
       ) %>% 
       insert_row(after = 7, "Family size dummies", rep("Yes", 6)) %>% 
       insert_row(after = 8, "Birth month dummies", rep("Yes", 6)) %>% 
       insert_row(after = 9, "Birth year dummies", rep("Yes", 6)) %>% 
       set_bottom_border(8:9, everywhere, 0) %>% 
       set_align(8:10, -1, "center") %>% 
       set_number_format(2:7, -1, 4) %>% 
       set_width(1) %>% 
       set_font_size(10) %>%
       set_escape_contents(12, 1, FALSE) %>%
       set_caption("Regressions of mediators on birth order (Great Britain)")
       
```


We first regress our measures of socio-economic status, university attendance
and income from first job, on birth order in the UK Biobank spouse pairs. We also do
the same for four non-SES mediators that could be affected by birth order: fluid
IQ, height, body mass index (BMI) and a measure of self-reported health. We
control for respondent's own PSEA and their parents' age at birth (see below). Table
\@ref(tab:tbl-bo-first-stage) shows that birth order significantly predicts all
these variables. Effect sizes are quite substantial: on average, one extra elder
sibling reduces the chance of attending university by about 
`r pretty(-coef_bo["University"] * 100)` percentage points, income by about 
£`r pretty(-coef_bo["Income"] * 1000, format = "d")`, fluid IQ by about 
`r pretty(-coef_bo["Fluid IQ"])` points on a 13 point test, height by about 
`r pretty(-coef_bo["Height"])` centimeters, and self reported health by 
`r pretty(-coef_bo["Health"])` points on a 4-point scale; and increases BMI by 
`r pretty(coef_bo["BMI"])`.
 
Next we run regressions of spouse PSEA on birth order. 
Table \@ref(tab:tbl-bo-psea-basic) reports the results. Column 1
reports results controlling only for family size (using dummies). As expected,
higher birth order is negatively associated with spouse's PSEA, though the
estimated effect size is small and insignificant. Column 2 reports results
controlling for the respondent's own PSEA, as well as dummies for birth year to
control for cohort effects, and dummies for birth month to control for
seasonal effects. The effect size of birth order is not much changed.

Column 3 reports results controlling for parents' age at birth. Within a family,
later children have older parents by definition. Older parents have more life
experience and may have higher income, which may help later children.[^7]
@kantarevic2006birth show that mother's age at childbirth indeed mechanically
offsets the negative effect of birth order. Including parents' age means we can
separate the effect of parental age from birth order. This reduces the N by a
lot, since only respondents with a live parent reported the necessary data.
However, the effect of birth order jumps in size and becomes significant at the
5 per cent level. Meanwhile, parents' age has a positive effect. This suggests
that estimates in columns 1-2 mixed two opposite-signed effects: having older
parents versus being later in birth order.

[^7]: We often only have data only for one parent. We use this, or take the mean
if we have both. There are also potential genetic effects from parental age,
though recent research has rejected these in favour of "social" explanations
[@kristensen2007explaining; @black2011older]. @cochran2013paternal report that
mutational load is approximately linear in father's age, while it is constant in
mother's age. We observe very similar results if we control only for father's
age at respondent's birth.


```{r tbl-bo-psea-basic}

fml_bo_psea_base <- list()

fml_bo_psea_base[[1]] <- EA3.y ~ birth_order.x | factor(n_sibs.x)
fml_bo_psea_base[[2]] <- EA3.y ~ birth_order.x + EA3.x + factor(birth_mon.x) |
                            factor(n_sibs.x) + factor(YOB.x)
  
fml_bo_psea_base[[3]] <- EA3.y ~ birth_order.x + EA3.x + factor(birth_mon.x) +
                            par_age_birth.x | factor(n_sibs.x) + factor(YOB.x)

fml_bo_psea_base <- lapply(fml_bo_psea_base, Formula::as.Formula)


mod_bo_psea_base <- lapply(fml_bo_psea_base, fixest::feols, 
                 data = mf_pairs_reg,
                 notes = FALSE
               )

tbl_note <- paste("Estimates from OLS regressions with spouse PSEA as dependent",
"variable, and own birth order as the main independent variable. PSEA is the",
"polygenic score for educational attainment, which is normalized with mean 0 and",
"standard deviation 1. We include own PSEA, parents' age at birth (the mean of parents'",
"ages), and further controls (family size, birth year, and birth month dummies)", 
"in columns 2-3 to ensure the balance of covariates",
"across birth order. All data is from the UK Biobank for a sample of UK",
"individuals born between 1935 and 1970.", my_note)

huxreg(mod_bo_psea_base, 
         coefs = c(
           "Birth order"           = "birth_order.x", 
           "Own PSEA"              = "EA3.x",
           "Parents' age at birth" = "par_age_birth.x"
         ),
         statistics = c(
           "N"  = "nobs", 
           "$R^2$" = "r.squared"
          ),
         note = tbl_note,
         stars = my_stars,
         tidy_args = tidy_args
       ) %>% 
       insert_row(after = 7, "Family size dummies", rep("Yes", 3)) %>% 
       insert_row(after = 8, "Birth month dummies", "No", "Yes", "Yes") %>% 
       insert_row(after = 9, "Birth year dummies", "No", "Yes", "Yes") %>% 
       set_bottom_border(8:9, everywhere, 0) %>% 
       set_align(8:10, -1, "center") %>% 
       set_number_format(2:7, -1, 4) %>% 
       set_width(0.9) %>% 
       set_font_size(10) %>%
       set_escape_contents(12, 1, FALSE) %>%
       set_caption("Regressions of spouse PSEA on birth order (Great Britain)")


```

Having tested that birth order affects spouse's PSEA, we now look for
potential mediators of this effect. Despite the lower N, we continue to
control for respondents' parents' age, since this removes a confound
which would bias our results towards zero.[^9]

[^9]: Table \@ref(tab:tbl-bo-psea-no-par-age) in the appendix reports results 
  without controlling for parents' age.

```{r tbl-bo-psea}
 

mod_bo_psea <- lapply(fml_bo_psea, fixest::feols, 
                 data = mf_pairs_reg,
                 notes = FALSE
               )

tbl_note <- paste("Estimates from OLS regressions with spouse PSEA as dependent",
"variable, and own birth order and mediators (university attendance and income)",
"as the main independent variables. Columns 2-4 correspond to model (7). PSEA is", 
"the polygenic score for educational attainment, which is normalized with mean 0",
"and standard deviation 1. We include own PSEA, mean of parents’ ages at birth,",
"potential non-SES mediators (fluid IQ, height, BMI, self-reported health)",
"and further controls (family size, birth year, and birth month dummies)",
"to ensure the balance of",
"covariates across birth order. All data is from the UK Biobank for a sample of", 
"UK individuals born between 1935 and 1970.", my_note)

huxreg(mod_bo_psea, 
         coefs = reg_coefs,
         statistics = c(
             "N"  = "nobs", 
             "$R^2$" = "r.squared"
            ),
         note = tbl_note,
         stars = my_stars,
         tidy_args = tidy_args
       ) %>% 
       insert_row(after = 19, "Family size dummies", rep("Yes", 4)) %>% 
       insert_row(after = 20, "Birth month dummies", rep("Yes", 4)) %>% 
       insert_row(after = 21, "Birth year dummies", rep("Yes", 4)) %>% 
       set_bottom_border(20:21, everywhere, 0) %>% 
       set_align(20:22, -1, "center") %>% 
       set_number_format(2:19, -1, 4) %>% 
       set_tb_padding(2) %>% 
       set_width(1) %>% 
       set_font_size(10) %>%
       set_escape_contents(24, 1, FALSE) %>%
       #set_font_size(final(1), everywhere, 8) %>%
       set_caption("Regressions of spouse PSEA on birth order and mediators 
                   (Great Britain)")
```

Table \@ref(tab:tbl-bo-psea) shows the results. Column 1 shows the
effect of birth order, using the same specification as column 3 of the
previous table. The remaining columns add potential mediators of
birth order effects. Column 2 controls for our first measure of
socio-economic status: university attendance. We also include potential
non-SES mediators, which are affected by birth order and might affect spouse
matching: fluid IQ, height, BMI and self-reported health. Column 3 adds our 
second measure of socio-economic status, income in first job. Column 4 includes 
both.

When we add university attendance and other mediators (column 2), the effect of
birth order drops and becomes insignificant at 5%, while the coefficient for
university is positive and highly significant. Fluid IQ, height and BMI are also
positive and significant, while self-reported health has the right sign but is
insignificant. Controlling for income instead of university attendance (column
3), again the effect of birth order shrinks and becomes insignificant, while
income has a positive and highly significant effect. Lastly, the same pattern
holds when we control for both university and income (column 4).

```{r tbl-mediation-prop}

# percent of main effect that is mediated by uni:
# uni effect of BO (first stage) * PSEA effect of uni (column 2)
# divided by PSEA effect of BO (column 1)

mediators <- c("university.xTRUE", "first_job_pay.x", "height.x",
                 "fluid_iq.x", "bmi.x", "sr_health.x")

n_mediators <- length(mediators)

compute_mediation_effects <- function (resample = FALSE) {
  mf_pairs_tmp <- mf_pairs_reg
  if (resample) {
    mf_pairs_tmp <- mf_pairs_tmp[sample(nrow(mf_pairs_tmp), replace = TRUE),]
  }
  
  mod_bo_1st_boot <- purrr::map(fml_bo_1st, fixest::feols, 
                                  data = mf_pairs_tmp, notes = FALSE)
  mod_bo_psea_boot <- purrr::map(fml_bo_psea, fixest::feols, 
                                   data = mf_pairs_tmp, notes = FALSE)
  mod_bo_overall_boot <- fixest::feols(fml_bo_psea[[1]], 
                                         data = mf_pairs_tmp, 
                                         notes = FALSE
                                       )
                                  

  # grab out the birth_order.x coefficient:
  mediator_effect_bo <- purrr::map_dbl(mod_bo_1st_boot, 
                                         list(coef, "birth_order.x"))

  psea_effect_mediators <- purrr::map(mod_bo_psea_boot[-1],
                                         ~coef(.x)[mediators])
  psea_effect_mediators <- matrix(unlist(psea_effect_mediators), 
                                    ncol = n_mediators,
                                    byrow = TRUE)
  colnames(psea_effect_mediators) <- mediators
  
  psea_effect_bo <- coef(mod_bo_overall_boot)["birth_order.x"]

  effects_uni <- mediator_effect_bo["University"] * 
    psea_effect_mediators[, "university.xTRUE"] 
  effects_income <- mediator_effect_bo["Income"] * 
    psea_effect_mediators[, "first_job_pay.x"]
  effects_height <- mediator_effect_bo["Height"] * 
    psea_effect_mediators[, "height.x"] 
  effects_fluid_iq <- mediator_effect_bo["Fluid IQ"] * 
    psea_effect_mediators[, "fluid_iq.x"]
  effects_bmi <- mediator_effect_bo["BMI"] * 
    psea_effect_mediators[, "bmi.x"]
  effects_health <- mediator_effect_bo["Health"] * 
    psea_effect_mediators[, "sr_health.x"]
  
  effects_pct <- cbind(
          University = effects_uni, 
          Income     = effects_income, 
          Height     = effects_height, 
          `Fluid IQ` = effects_fluid_iq,
          BMI        = effects_bmi,
          `Self-reported health` = effects_health
        )
  # divide by overall effect re-estimated using data in each of models 2-4
  effects_pct <- effects_pct / psea_effect_bo
  
  effects_pct
}


# for each 1st stage model, run 199 bootstraps and gather coefs
n_reps <- 199

# bootstraps is now 3 x 4 x n_reps
# cis <- apply(bootstraps, 1:2, quantile, c(0.05, 0.95), na.rm = TRUE)
estimates <- compute_mediation_effects(resample = FALSE)

format_est_ci <- function(coef, model) {
  # char <- glue::glue("{percent(estimates[model, coef], 0.1)} ({percent(cis[1, model, coef], 0.1)}, {percent(cis[2,model, coef], 0.1)})")
  char <- estimates[model, coef] * 100
  char[is.na(estimates[model, coef])] <- ""
  char
}

med_matrix <- format_est_ci(c("University", "Income", "Fluid IQ", "Height", 
                                "BMI", "Self-reported health"), 
                              1:3)
med_matrix <- matrix(med_matrix, ncol = 3, byrow = TRUE)
rownames(med_matrix) <- c("University", "Income", "Fluid IQ", "Height", "BMI",
                            "Self-reported health")
colnames(med_matrix) <- paste("Model", 2:4, " (%)")

tbl_note <- paste("Percentage of the effects of birth order in Table", 
"\\ref{tab:tbl-bo-psea}, columns 2 to 4, explained by by each mediating variable.")

as_huxtable(med_matrix, add_colnames = TRUE, add_rownames = "") %>% 
      theme_article() %>% 
      add_footnote(tbl_note) %>% 
      set_escape_contents(final(1), everywhere, FALSE) %>% 
      set_width(0.7) %>% 
      set_col_width(c(.4, .2, .2, .2)) %>% 
      set_align(1, -1, "right") %>% 
      set_align(-1, -1 , ".") %>% 
      set_number_format(-1, -1, 1) %>% 
      set_tb_padding(3) %>% 
      set_bottom_border(1, 1, 0) %>% 
      set_font_size(10) %>%
      set_caption(
        "Percent of birth order effects accounted for by mediators (Great Britain)"
      )  

```


Under the assumptions discussed above, we can estimate the proportion of
the birth order effect that is mediated by these variables. Table
\@ref(tab:tbl-mediation-prop) reports this for each model in columns
2-4. Each estimate is the coefficient of birth order on the mediator,
times the coefficient of the mediator on spouse PSEA, divided by the
coefficient of birth order on spouse PSEA estimated from column 1, i.e.
without mediators. Education explains about 38-55 percent of the effect, much more
than all the other mediators. Income, fluid IQ, height and BMI all explain 
between 6 and 17 percent of the effect, depending on the specification.

These results provide evidence that birth order affects spouse PSEA via
education and income, with education being especially important. The effect
size of birth order is small (a few percent of a standard deviation of PSEA),
but what matters is the effect size of education and income. The effect of
education in particular is quite large as measured here, and since it
also appears to mediate the purely environmental shock of birth order, it
cannot just be due to an unobserved correlation with own genetics.

Our next regressions split up the data into subsets. Cultural stereotypes often
assume that the link between status and genes is not symmetric across the
genders, for example, that males with high SES are particularly likely to marry
attractive spouses. Claim \@ref(claim-men-women-different) showed that these
differences would strengthen the effects of SGAM. To test for this, we
separately regress female spouses' PSEA on male birth order, and male spouses'
PSEA on female birth order. We also rerun regressions among the subset of
individuals who had children. A significant result here will confirm that the
association between status and genetics is carried over into the next
generation.

Table \@ref(tab:tbl-bo-subsets) shows the results. Columns 1 and 2 present
results using birth order of male respondents to predict female spouses' PSEA.
Column 1 shows the regression of birth order plus controls; in column 2, we add
university attendance and non-SES mediators (here, we exclude first job income
so as to keep our N large). Columns 3 and 4 repeat the exercise for female
respondents, using their birth order to predict male spouses' PSEA. The effect of
birth order is imprecisely estimated in these subsets due to the lower sample
size. However, the pattern of coefficient sizes is the same as in the main
regression: the coefficient of birth order is about -0.3 (and very similar 
between the sexes), and adding university attendance reduces the absolute size 
of the birth order effect. Columns 5 and 6 show results from regressions on 
the subsample of couples with children. Here, birth order is significant in the 
base specification, and again, university attendance still seems to mediate the 
birth order effect.


```{r tbl-bo-subsets}

mod_bo_males <- lapply(fml_bo_psea[1:2], fixest::feols, 
                 data = mf_pairs_reg %>% filter(x == "Male"),
                 notes = FALSE
               )
mod_bo_females <- lapply(fml_bo_psea[1:2], fixest::feols, 
                 data = mf_pairs_reg %>% filter(x == "Female"),
                 notes = FALSE
               )

mf_pairs_children <- mf_pairs_reg %>% 
                   filter(n_children.x > 0, n_children.y > 0)
mod_bo_children <- lapply(fml_bo_psea[1:2], fixest::feols, 
                 data = mf_pairs_children,
                 notes = FALSE
               )
# 
# ta_m <-  list(conf.int = FALSE, cluster = 
#                 list(mf_pairs_reg$f.54.0.0.x[mf_pairs_reg$x =="Male"]))
# ta_f <-  list(conf.int = FALSE, cluster = 
#                 list(mf_pairs_reg$f.54.0.0.x[mf_pairs_reg$x == "Female"]))
# ta_ch <-  list(conf.int = FALSE, cluster = list(mf_pairs_children$f.54.0.0.x))

reg_list <- list(
        "Male respondents"     = mod_bo_males[[1]],
        "Male respondents"     = mod_bo_males[[2]],
        "Female respondents"   = mod_bo_females[[1]],
        "Female respondents"   = mod_bo_females[[2]],
        "With children"        = mod_bo_children[[1]],
        "With children"        = mod_bo_children[[2]]
      )

tbl_note <- paste("Estimates from OLS regressions corresponding to",
"columns 1 and 2 in Table \\ref{{tab:tbl-bo-psea}}, separately for males, females",
"and respondents with children. Spouse PSEA is the dependent variable, and own",
"birth order and university attendance are the main independent variables. PSEA",
"is the polygenic score for educational attainment, which is normalized with",
"mean 0 and standard deviation 1. We include own PSEA, parents’ age at birth (the mean of",
"parent’s ages) and further controls (family size, birth year, and birth month",
"dummies) to ensure the balance of", 
"covariates across birth order. All data is from the UK Biobank for a sample of", 
"UK individuals born between 1935 and 1970.", my_note)

huxreg(reg_list,
         coefs      = reg_coefs[reg_coefs != "first_job_pay.x"],
         statistics = c("N" = "nobs", "$R^2$" = "r.squared"),
         note       = tbl_note,
         stars      = my_stars,
         tidy_args  = tidy_args
       ) %>% 
       set_escape_contents(final(2), everywhere, FALSE) %>% 
       insert_row(after = 17, "Family size dummies", rep("Yes", 6)) %>% 
       insert_row(after = 18, "Birth month dummies", rep("Yes", 6)) %>% 
       insert_row(after = 19, "Birth year dummies", rep("Yes", 6)) %>% 
       set_bottom_border(18:19, everywhere, 0) %>% 
       set_align(18:20, -1, "center") %>% 
       set_font_size(10) %>% 
       set_width(1) %>% 
       set_tb_padding(3) %>% 
       set_lr_padding(3) %>% 
       set_caption("Regressions of spouse PSEA on birth order: subsets (Great Britain)")

```

\FloatBarrier

# Results: Norway

Now we turn to the results from MoBa in Norway. Some of the variables are
different from those for UK Biobank. Spouse PSEA is calculated using summary
statistics from @okbay2022polygenic, aka "EA4", rather than "EA3".[^r2-comparison]
Income is from all sources, reported at age 30, converted to a z score. In particular, 
some low-income individuals may be in continuing education or in relationships
already.[^age-25] Data on IQ and self-reported health is unavailable. The sample
is also younger than UK Biobank, spouse pairs are given rather than constructed,
and all couples have at least one child.

[^r2-comparison]: The $R^2$ on own university attendance is 
`r pretty(r2_own_psea_moba, 3)` for EA4 in MoBa, compared to 
`r pretty(r2_own_psea, 3)` for EA3 in UK Biobank.

[^age-25]: We also tried income at age 25. This gave odd results, with a
significant negative beta on spouse PSEA. A possible reason is that in Norway many
potential high earners are still in higher education at age 25. Income at age 25
correlated only at 0.1 with income at age 30, and was negatively correlated with
educational attainment. Overall, we think income at 30 is more informative as a 
measure of SES.

Tables \@ref(tab:tbl-bo-psea-moba) and \@ref(tab:tbl-mediation-moba) are the
equivalent of Tables \@ref(tab:tbl-bo-psea) and \@ref(tab:tbl-mediation-prop)
for respondents in the MoBa dataset. (Equivalents to Tables
\@ref(tab:tbl-bo-first-stage) and \@ref(tab:tbl-bo-psea-basic) are in the
appendix.) The broad pattern of results is similar to Britain. The larger N
gives higher statistical significance. Effects of all the variables are in the
expected direction, except that the coefficient on income changes sign when
university is also included. In particular, the point estimate of the total
effect of birth order is about twice as high as in Britain, and the estimated
effect of university attendance is also higher. The effect of own PSEA is also
about twice as high. Note that these differences could be driven by the PSEA score 
containing less noise in the Norwegian sample.

Adding university attendance again substantially and significantly reduces 
the effect of birth order, though here, birth order remains independently
significant and substantively large even controlling for university. 
In Norway, however, it is much less clear that income is an important mediator on its own. Adding income barely changes the effect of birth order (column 3). 

Table \@ref(tab:tbl-mediation-moba) computes the percentages of the effects that
are mediated by our variables. The percentage effects of education, when
controlling for income, are about the same in both countries. But the effect of
income in the Norwegian sample is much smaller and very close to zero.

Table \@ref(tab:tbl-bo-subsets-moba) runs our regressions separately for focal
males and focal females only.[^no-kids] As in Britain, coefficients look very
similar across the sexes. The effects of birth order are highly significant in
either sex, and adding the mediators significantly reduces them. Interestingly,
the effect of BMI is about 50% larger for women, as in Britain, and here the
difference is significant. On the other hand, in Norway, there is no difference
between the male and female coefficients on university attendance.

[^no-kids]: We don't run separate regressions for families with children, since
MoBa only includes families with children by design.

Overall, the Norway results show two things. First, they clearly confirm that
birth order affects spouse PSEA, with education a key mediator. Second, they
suggest that the effects of SES in marriage markets vary between the two
countries. Education has a similar effect on spouse PSEA in Norway, but income
has a much smaller effect. Of course this is a loose comparison, since even
conditioning on our controls, samples and measures in the two countries are
different. Still, it is interesting that in Norway, a more egalitarian country
than the UK, income seems to have a smaller effect on spouse PSEA, while own
PSEA, a genetic characteristic, seems to have a higher effect. In terms of our
theory, the genetic contribution to attractiveness $a$ appears larger in Norway. 
This suggests that $a$ is itself endogenous to economic institutions.

```{r tbl-bo-psea-moba}

# Notes. 
# - Would still be good to get income from first employment
#   - In the end we might want Abdel to replicate Norway v. closely...
# - Can we get a subset of actual siblings? Would be so coool....


# Robustness checks todo:
# - balance tests on some pgs?
# - birth order dummies
# - education (can already)
mod_bo_psea_moba <- purrr::map2(export_table3_tidyse,
                                  export_table3_glance,
                                  convert_moba_for_huxreg)
names(mod_bo_psea_moba) <- NULL


tbl_note <- paste("Estimates from OLS regressions with spouse PSEA as dependent",
"variable, and own birth order and mediators (university attendance and income)",
"as the main independent variables. Columns 2-4 correspond to model (7). PSEA is", 
"the polygenic score for educational attainment, which is normalized with mean 0",
"and standard deviation 1. We include own PSEA, mean of parents’ ages at birth,",
"potential non-SES mediators (height and BMI)",
"and further controls (family size, birth year, and birth month dummies)",
"to ensure the balance of",
"covariates across birth order. All data is from the MoBa dataset for a sample of", 
"spouse pairs with a child between 1999 and 2008.", my_note)

huxreg(mod_bo_psea_moba,
         coefs = moba_reg_coefs,
         statistics = c("N" = "nobs", "$R^2$" = "r.squared"),
         note       = tbl_note,
         stars      = my_stars
       ) %>% 
       insert_row(after = 15, "Family size dummies", rep("Yes", 4)) %>% 
       insert_row(after = 16, "Birth month dummies", rep("Yes", 4)) %>% 
       insert_row(after = 17, "Birth year dummies", rep("Yes", 4)) %>% 
       set_bottom_border(16:17, everywhere, 0) %>% 
       set_align(16:18, -1, "center") %>% 
       set_number_format(2:15, -1, 4) %>% 
       set_tb_padding(2) %>% 
       set_width(1) %>% 
       set_font_size(10) %>%
       set_escape_contents(20, 1, FALSE) %>%
       set_caption("Regressions of spouse PSEA (Norway)")
```



```{r tbl-mediation-moba}

mediator_effect_bo <- purrr::map_dbl(export_table1_tidyse, 
                                     function (x) {
                                       x$estimate[x$term == "parity"]
                                     })

mediators <- c("university", "incomez", "height", "bmi")
psea_effect_mediators <- purrr::map(export_table3_tidyse[2:4],
                                        function (x){
                                          rows <- match(mediators, x$term)
                                          x$estimate[rows]
                                        })
psea_effect_mediators <- matrix(unlist(psea_effect_mediators), 4, 3)
rownames(psea_effect_mediators) <- mediators

psea_effect_bo <- export_table3_tidyse$column1 %>% 
                    filter(term == "parity") %>%
                    pull(estimate)

mediated_effects <- mediator_effect_bo * psea_effect_mediators

mediated_effects_pct <- mediated_effects / psea_effect_bo * 100

mediated_effects_pct <- format(mediated_effects_pct, digits = 1) 
mediated_effects_pct <- gsub(".*NA", "", mediated_effects_pct)
colnames(mediated_effects_pct) <- paste("Model ", 2:4, "(%)")
rownames(mediated_effects_pct) <- c("University", "Income", "Height", "BMI")


tbl_note <- paste("Percentage of the effects of birth order in Table", 
"\\ref{tab:tbl-bo-psea-moba}, columns 2 to 4, explained by by each mediating variable.")

as_hux(mediated_effects_pct, add_colnames = TRUE, add_rownames = "") %>%
      theme_article() %>% 
      add_footnote(tbl_note) %>% 
      set_escape_contents(final(1), everywhere, FALSE) %>% 
      set_width(0.7) %>% 
      set_col_width(c(.4, .2, .2, .2)) %>% 
      set_align(1, -1, "right") %>% 
      set_align(-1, -1 , ".") %>% 
      set_number_format(-1, -1, 1) %>% 
      set_tb_padding(3) %>% 
      set_bottom_border(1, 1, 0) %>% 
      set_font_size(10) %>%
      set_caption(
        "Percent of birth order effects accounted for by mediators (Norway)"
      )  
```


```{r tbl-bo-subsets-moba}

mod_bo_subsets_moba <- purrr::map2(export_table5_tidyse,
                                     export_table5_glance, 
                                     convert_moba_for_huxreg)

names(mod_bo_subsets_moba) <- rep(c("Male respondents", "Female respondents"), 
                                    each = 2)

tbl_note <- paste("Estimates from OLS regressions corresponding to",
"columns 1 and 2 in Table \\ref{{tab:tbl-bo-psea}}, separately for males, females",
"and respondents with children. Spouse PSEA is the dependent variable, and own",
"birth order and university attendance are the main independent variables. PSEA",
"is the polygenic score for educational attainment, which is normalized with",
"mean 0 and standard deviation 1. We include own PSEA, parents’ age at birth (the mean of",
"parent’s ages) and further controls (family size, birth year, and birth month",
"dummies) to ensure the balance of", 
"covariates across birth order. All data is from the MoBa dataset for a sample of", 
"spouse pairs with a child between 1999 and 2008.", my_note)

huxreg(mod_bo_subsets_moba,
         coefs = moba_reg_coefs[moba_reg_coefs != "income"],
         statistics = c("N" = "nobs", "$R^2$" = "r.squared"),
         note       = tbl_note,
         stars      = my_stars
       ) %>%
       set_escape_contents(final(1), everywhere, FALSE) %>% 
       insert_row(after = 13, "Family size dummies", rep("Yes", 4)) %>% 
       insert_row(after = 14, "Birth month dummies", rep("Yes", 4)) %>% 
       insert_row(after = 15, "Birth year dummies", rep("Yes", 4)) %>% 
       set_bottom_border(14:15, everywhere, 0) %>% 
       set_align(11:13, -1, "center") %>% 
       set_font_size(10) %>% 
       set_width(1) %>% 
       set_tb_padding(3) %>% 
       set_lr_padding(3) %>% 
       set_escape_contents(18, 1, FALSE) %>%
       set_caption("Regressions of spouse PSEA: subsets (Norway)")
```

\FloatBarrier

# Robustness

Although all children of the same parents have the same polygenic scores
in expectation, it might still be possible that genetics correlates with
birth order within the sample. This could happen in three ways. First,
siblings with high birth order will typically come from larger families
than those with low birth order, and parents of different-sized families
are likely to differ systematically on many dimensions, including
genetics. We controlled for this by including a full set of family size
dummies in the regression. Second, there could be selection bias. For
example, if later siblings with high PSEA, and earlier siblings with low
PSEA, are more likely to enter the sample, then this would bias our
results. Thirdly, parents might choose family size in a way related to
genetics. For example, suppose that when the first child has a phenotype
reflecting a high PSEA, parents are more likely to have a second child.
Then within the subset of two-child families, first children would have
higher-than-average PSEA, while second children would not.

```{r calc-pgs-check}

check_cor <- function (score_name) {
  fml_check <- as.formula(paste0(score_name, ".x ~ factor(n_sibs.x) + birth_order.x"))
  mod_check <- lm(fml_check, mf_pairs_reg)
  mod_check %>% 
        tidy() %>% 
        filter(term == "birth_order.x") %>% 
        mutate(score = score_name)
}

check_cor_within <- function (score_name) {
  fml_check_within <- as.formula(paste0(score_name, 
                        ".x ~ 0 + factor(n_sibs.x) + factor(n_sibs.x):factor(birth_order.x)"))
  mod_check <- lm(fml_check_within, mf_pairs_reg)
  # NB: the model creates NA estimates for when n_sibs <= birth_order
  # since these have 0 rows in the data. I think this is not a problem.
  mod_check %>% 
        tidy() %>% 
        filter(grepl("birth_order", term)) %>% 
        filter(! is.na(estimate)) %>% 
        mutate(score = score_name)
}

drake::loadd(score_names)
score_names <- paste0(score_names, "_resid")
balance_check <- purrr::map_dfr(score_names, check_cor)
balance_check_within <- purrr::map_dfr(score_names, check_cor_within)

if (any(balance_check$p.value < 0.1/33)) stop("Some check p values < 0.1/33; rewrite text!")

n_below_10 <- sum(balance_check$p.value < 0.10)
stopifnot(n_below_10 == 4)
if(any(
        balance_check %>% filter(p.value < 0.10) %>% pull(estimate) > 0.02
      )) stop("Some coefs > 0.02; rewrite text!")

n_within_coefs <- nrow(balance_check_within)
if (any(balance_check_within$p.value < 0.001)) stop("Some within-siblings p values < 0.001; rewrite text!")


max_coef <- max(abs(balance_check$estimate))

mod_bo_own_psea <- lm(EA3.x ~ birth_order.x, 
                      mf_pairs_reg,
                      subset = n_sibs.x==3)
coef_own_psea_3 <- tidy(mod_bo_own_psea)[[2, "estimate"]]
pval_own_psea_3 <- tidy(mod_bo_own_psea)[[2, "p.value"]]
stopifnot(pval_own_psea_3 < 0.1)
stopifnot(pval_own_psea_3 > 0.05)

```

To check for the latter two problems, we run balance tests on 33 different
polygenic scores in the UK Biobank sample.[^10] We regress each score on own
birth order, controlling for family size. No scores were significant at $p <
0.10/33$. Four scores were significant at $p < 0.10$, all with effect sizes of
less than 0.02 per standard deviation. Table \@ref(tab:tbl-bo-psea-pgs) in the
appendix reports
regressions controlling for these scores. Results are almost unchanged. To test
whether polygenic scores might vary across birth orders within a particular
family size, we also regress each score on a full set of birth order dummies,
interacted with a full set of family size dummies. None of the 
`r n_within_coefs` birth order coefficients were significant at $p < 0.001$.
However, among families of size 3, there is a marginally significant positive
correlation of birth order with own PSEA (effect size 
`r pretty(coef_own_psea_3, 3)`, $p = `r pretty(pval_own_psea_3, 2)`$). Table
\@ref(tab:tbl-bo-psea-no3) in the appendix therefore reports regressions with
families of size 3 excluded. Results are substantially unchanged. Of course,
there could still be unmeasured genetic variants which correlate with birth
order in our sample. Nevertheless, a wide set of polygenic scores shows no large
or significant correlation. This makes us more confident that birth order is
indeed exogenous to genetics.

[^10]: Polygenic scores were residualized on the first 100 principal
    components of the genetic data. Scores were for: ADHD, age at
    menarche, age at menopause, agreeableness, age at smoking
    initiation, alcohol use, Alzheimer's, autism, bipolarity, BMI, body
    fat, caffeine consumption, cannabis (ever vs. never), cognitive
    ability, conscientiousness, coronary artery disease, smoking
    (cigarettes per day), type II diabetes, drinks per week, educational
    attainment (EA2 and EA3), anorexia, extraversion, height, hip
    circumference, major depressive disorder, neuroticism, openness,
    smoking cessation, schizophrenia, smoking initiation, waist
    circumference, and waist-to-hip ratio. For full details of score
    construction, see @abdellaoui2019genetic. We also ran similar tests on
    MoBa data, and found no significant associations between birth order
    and polygenic scores.


```{r calc-oster-2019, eval = FALSE}
# Plan:
# 1. pick the max R2 of spousal PSEA, backing it out

# 2. for each SES mediator (university, first_job_pay):
#   run the "short regression" including only BO, the SES mediator and controls
#   run the "medium regression" including these + non-SES mediators
#   using these statistics and the max R2, calculate the delta 
#   (degree of selection into treatment on unobservables, compared to 
#   selection on observables)
#   that would lead the effect of the SES mediator to be zero

# we need the max R2 of spousal PSEA *from an unobserved environmental factor*, 
# given that we know birth order's effect could only be via the environment.
# So, sources of error are (1) limits to assortative mating
# (2) error in measured PSEA (3) only environmental factors can contribute
# to the R2
# 

mod_ea_psea <- fixest::feols(age_fulltime_edu ~ EA3_excl_23andMe_UK, 
                               data = famhist, notes = FALSE)
R2_ea_psea <- fixest::r2(mod_ea_psea, type = "r2")
prop_systematic_psea <- R2_ea_psea/0.4 

true_psea_R2_max <- 0.5
R2max <- prop_systematic_psea * true_psea_R2_max

res_uni <- robomit::o_delta(y = "EA3.y", x = "university.x", 
                            con = "fluid_iq.x + height.x + 
                                   bmi.x + sr_health.x", 
                            m = "EA3.x + par_age_birth.x + factor(n_sibs.x) +
                                 factor(YOB.x) + factor(birth_mon.x)", 
                            R2max = R2max, type = "lm", data = mf_pairs_reg)
res_fjp <- robomit::o_delta(y = "EA3.y", x = "first_job_pay.x", 
                            con = "fluid_iq.x + height.x + bmi.x +
                                   sr_health.x", 
                            m = "EA3.x + par_age_birth.x + factor(n_sibs.x) +
                                 factor(YOB.x) + factor(birth_mon.x)", 
                            R2max = R2max, type = "lm", data = mf_pairs_reg)

delta_uni <- round(res_uni[[1, 2]], 3)
delta_fjp <- round(res_fjp[[1, 2]], 3)

```


Another concern is that our chosen SES mediators might not be exogenous. We have
already seen that birth order affects intelligence, height, BMI
and health. So there might be other unobserved variables which mediate the effect
of birth order on spousal PSEA, and which correlate with education or income,
but which do not themselves capture SES. If so, that would threaten our claim
that education and income are important mediators. However, the effects of
education on spouse PSEA in Table \@ref(tab:tbl-bo-psea), and of birth order on
education in Table \@ref(tab:tbl-bo-first-stage), are both large and highly
significant. In other literature on spouse matching, education is a common,
robust and significant predictor. For these reasons, we think that our results
are unlikely to be driven wholly by other, unobserved mediators.

A last concern is that polygenic scores, including PSEA, contain noise from
correlated environments. (That is, effect sizes of individual SNPs may be
confounded in the underlying statistical analyses used to create polygenic
scores.) It is conceivable that birth order could only affect the noise
component of spouse PSEA, rather than the component which is truly causal for
education. We think this is unlikely for several reasons. First, our polygenic 
scores were calculated using per-SNP summary statistics estimated on non-UK
populations. So they will only include noise that correlates with social
environment insofar as non-causal correlations of SNPs with social status
are the same across different countries.
Second, we have residualized PSEA on 100 principal components of the genetic
data, a standard technique in genetics to avoid confounding causal effects with
population stratification. Third, true causal effects of individual
SNPs are highly correlated (r = 0.74) with their "population effects"
including noise [@young2022mendelian]. Lastly, it is hard to imagine an 
assortative mating process by which people match spouses who have genetic 
variants that correlate with educational attainment, but not genetic variants 
that cause it.

Our main specification is linear in birth order. Tables
\@ref(tab:tbl-bo-psea-dummies) and \@ref(tab:tbl-bo-psea-dummies-moba) in the
appendix run specifications with separate dummies for each birth order. The
pattern that birth order coefficients shrink after controlling for SES mediators
holds robustly across all birth orders in both Britain and Norway.

UK Biobank is not a representative sample of the population. Table
\@ref(tab:tbl-bo-psea-weights) in the appendix weights cases to match the
Biobank's sampling frame. Results are similar to those in the main text. 
Although this is still not representative of the population as a whole, it 
provides some assurance that our results are not driven by volunteering bias.

The appendix reports other robustness checks, including replacing university
attendance with age of leaving full-time education. Overall, while significance
sometimes varies, the pattern of results is remarkably consistent. Birth order is
always negatively associated with spouse PSEA, and this effect is always reduced
in magnitude after adding education as a mediator. Effect sizes are also
consistent, with the exception that they are smaller if we do not control for 
parental age (just as in Table \@ref(tab:tbl-bo-psea-basic)). 


# Conclusion

Our empirical analysis shows that in Great Britain and Norway, two contemporary
developed countries, earlier-born siblings had spouses with
higher PSEA. We also provide evidence that these effects are mediated by
socio-economic status, specifically income and education. We interpret this as
evidence of social-genetic assortative mating (SGAM). 

There are multiple mechanisms by which advantage is transmitted across
generations. High-income parents may invest more in their children's human
capital, transfer wealth via gifts and bequests, model valuable skills, or
provide them with advantageous social networks. They may also pass on causally
relevant genetic variants. This channel has been proposed as a reason for the
surprising persistence of inequality over generations
[@clark2015intergenerational; @clark2023inheritance]. One problem with this 
theory is
that in the absence of assortative mating, genetic variation regresses swiftly
to the mean, with coefficient $r = 0.5$ per generation. Thus to explain long-run
persistence, the genetic theory seems to require very high levels of genetic
assortative mating. SGAM may help to solve this puzzle. Persistence will be
increased if, in addition to genetic assortative mating, high SES itself
attracts "good genes". At the same time, SGAM changes the interpretation of 
genetics. As our model shows, genetic variation is not an exogenous input into 
the social system, but an endogenous outcome -- not a confound for wealth, but a mediator. 

SGAM also provides a new explanation for the genes-SES gradient -- the observed
association of genes with SES -- which is an important cause of educational
and occupational inequality, and perhaps also of health inequalities. The
leading alternative explanation is meritocratic social mobility. Whilst
meritocracy exists in modern capitalist economies, it has been far more limited
in most societies throughout history [@smelser1966social]. On the other hand,
assortative mating is likely to be a cultural universal [@buss1989sex]. Thus,
SGAM predicts that genes-SES gradients should exist in all stratified societies.
In fact, people in many societies have believed that innate traits do vary by
social status.[^quotes] In future, it may be possible to directly test for
genetic differences across social status in ancient DNA samples.

[^quotes]: The appendix contains a selection of relevant historical quotations.

In our model, the association between SES and genetic variation depends on
economic and social institutions. Institutions that make wealth more persistent
across generations also increase the correlation between SES and genetics. If
so, then institutional differences may have long-run effects over generations by
altering the genes-SES gradient. There could be hysteresis, with initial social
differences cumulating over time via their effect on genetic inequality. On the
other hand, while lowering the intergenerational transmission of wealth may
eventually flatten the genes-SES gradient, increases in the level of meritocracy
paradoxically make it steeper, suggesting a deep conflict between meritocracy
and egalitarianism [@young1958rise; @markovits2019meritocracy]. Lastly, the
structure of marriage markets also affects the gradient. Our empirical analysis
suggests that in relatively egalitarian Norway, income plays a less important role in
assortative mating than it does in the UK. However, this is a loose comparison,
and we see careful tests of comparative statics across different contexts as
an important challenge for future work.

The broadest message of this paper is that genetics are a social
outcome. Both popular and scientific discourse often parse genetics as
"nature", in opposition to "nurture" or "environment" [e.g.
@chakravarti2003nature; @plomin2019blueprint]. This reflects the fact that
our individual genetic endowment is fixed at birth, affects our body and brain
through proximate biological mechanisms, and cannot be changed by our social
environment (though genes always interact with the environment to cause
individual outcomes). But the idea that human genetics are natural can be highly
misleading. Humans inherit their genes from their parents, along with other
forms of inheritance such as economic and cultural capital. Human parents, in
turn, form spouse pairs and bear children within social institutions. A person's
genetic inheritance is a social and historical fact about them, not just a
fact of nature. As @marx1844economic wrote, "History is the true natural history of
man". 



Genetic endowments can even be viewed as another form of capital,
alongside human, social and cultural capital: a resource to be sought,
accumulated and competed over. The analysis of this kind of capital is an
exciting area for further research, which will require the contributions of both
economists and geneticists.


# Acknowledgements

The Norwegian Mother, Father and Child
Cohort Study (MoBa) is a population-based pregnancy cohort study conducted by
the Norwegian Institute of Public Health. Participants were recruited from all
over Norway from 1999-2008. The women consented to participation in 41% of the
pregnancies. The cohort includes approximately 114,500 children, 95,200 mothers
and 75,200 fathers. The current study is based on version
12 of the quality-assured data files released for research in 2019. The
establishment of MoBa and initial data collection was based on a license from
the Norwegian Data Protection Agency and approval from The Regional Committees
for Medical and Health Research Ethics. The MoBa cohort is currently regulated
by the Norwegian Health Registry Act. The current study was approved by The
Regional Committees for Medical and Health Research Ethics.

We thank the Norwegian Institute of Public Health (NIPH) for generating high-quality
genomic data. This research is part of the HARVEST collaboration, supported by the
Research Council of Norway (#229624). We also thank the NORMENT Centre for providing
genotype data, funded by the Research Council of Norway (#223273), South East Norway
Health Authorities and Stiftelsen Kristian Gerhard Jebsen. We further thank the Center for
Diabetes Research, the University of Bergen for providing genotype data and performing
quality control and imputation of the data funded by the ERC AdG project
SELECTionPREDISPOSED, Stiftelsen Kristian Gerhard Jebsen, Trond Mohn Foundation,
the Research Council of Norway, the Novo Nordisk Foundation, the University of Bergen,
and the Western Norway Health Authorities.

This study was conducted using UK Biobank resources under application number 
40310. 

AA is supported by the Foundation Volksbond Rotterdam and by ZonMw grant 
849200011 from The Netherlands Organisation for Health Research and Development. 

Code to reproduce is available at https://github.com/hughjonesd/trading-genetics.


\FloatBarrier

\newpage

# Appendix: for online publication

\localtableofcontents
\clearpage

## Proofs

```{r proofs, child = "proofs.Rmd"}

```

\FloatBarrier

\newpage

## More empirical results

Figure \@ref(fig:pic-basic-corr-moba) redoes Figure \@ref(fig:pic-basic-corr)
for the MoBa data, plotting measures of individual SES against spouse PSEA. 
While university attendance gives results similar to the UK, the relationship 
between income decile and spouse PSEA is interestingly nonlinear in both
sexes. Some low-income individuals may be out of the labour market,
either in continuing education or as a stay-at-home spouse.

Table \@ref(tab:tbl-bo-first-stage-moba) shows regressions of birth order on
mediators for the MoBa data. Birth order predicts all the mediators significantly
and with large effect sizes, except for BMI.

Table \@ref(tab:tbl-bo-psea-basic-moba) regresses birth order on spouse PSEA
for the MoBa data, starting with a simple bivariate regression and adding
controls as in Table \@ref(tab:tbl-bo-psea-basic). Interestingly, here, 
the negative effect of birth order is highly significant even before we 
control for parental age. As in Britain, the effect is greatly increased
when we control for parental age.

Table \@ref(tab:tbl-bo-psea-no-par-age) reruns our central regressions,
dropping the control for parents' age at birth. Results show the same
pattern as in the main text: the coefficient for birth order is
negative, but changes sign when university attendance is added as a
potential mediator. However, the birth order effect is smaller overall,
and is never significant. We also ran regressions using father's age
only: results are similar to those in the main text.

Tables \@ref(tab:tbl-bo-psea-dummies) and \@ref(tab:tbl-bo-psea-dummies-moba)
rerun our central regressions estimating a separate coefficient for each position
in the birth order (with firstborn as the baseline). The basic pattern of our
main result is remarkably robust, in both Great Britain and Norway: birth order
coefficients are generally negative, and adding mediators always
causes them to increase towards zero or to change sign. 

We also ran a specification with separate birth order dummies within
each family size. Figure \@ref(fig:pic-bo-psea-interactions) shows 95%
confidence intervals for the birth order coefficients, from the column 2
specification including height and IQ controls but no mediators. Not
surprisingly, coefficients are imprecisely estimated. But most birth
order coefficients are negative compared to the baseline for firstborns.

Table \@ref(tab:tbl-bo-psea-weights) re-estimates Table \@ref(tab:tbl-bo-psea)
using weights from @vanalten2022reweighting. These weight the UK Biobank sample
to match its sampling frame of 40-69 year olds living close to 22 assessment
centres. Although these weights probably bring the sample closer to the UK population, the sampling frame is still not representative of that population: 
for instance, urban areas are oversampled. Results are similar to those in
the main text, although birth order coefficients are absolutely larger in all
the specifications. 

Table \@ref(tab:tbl-bo-psea-pgs) reruns our regressions controlling for
several polygenic scores. Results are very close to those in the main
text.

Table \@ref(tab:tbl-bo-psea-age-fte) and \@ref(tab:tbl-bo-psea-age-fte-moba)
rerun our regressions using age of leaving full-time education as a
measure of educational SES, instead of the university attendance dummy.
Results are similar to those in the main text: controlling for age of
leaving full-time education shrinks the effect of birth order substantially.

Table \@ref(tab:tbl-bo-psea-no3) reruns Table \@ref(tab:tbl-bo-psea)
excluding families of size 3. Results are very similar to those in the
main text.

```{r pic-basic-corr-moba, fig.cap = "Spouse polygenic score for educational attainment (PSEA) against own university attendance and own income at age 30 (Norway). Lines show 95\\% confidence intervals. PSEA is normalized to have mean 0 and variance 1.", fig.subcap = rep("", 4), fig.align = "center", fig.width = 3, fig.height = 3, fig.ncol = 2}



pic_data_norway_uni <- figure4university %>%
                       mutate(
                         University = ifelse(university == 1, "Yes", "No")
                       )
pic_data_norway_income <- figure4incomedec %>%
                          mutate(
                            Income = factor(incomedec)
                          )


pic_cc_norway_uni <- coord_cartesian(ylim = c(-0.3, 0.3))

pic_data_norway_uni %>%
  filter(sex == 1) %>%
  ggplot(aes(University)) + 
    geom_pointrange(aes(y = mean, ymin = lower, ymax = upper)) +
    pic_theme + 
    pic_cc_norway_uni +
    labs(x = "Male spouse university attendance", y = "Female spouse PSEA")

pic_data_norway_uni %>%
  filter(sex == 2) %>%
  ggplot(aes(University)) + 
    geom_pointrange(aes(y = mean, ymin = lower, ymax = upper)) +
    pic_theme + 
    pic_cc_norway_uni +
    labs(x = "Female spouse university attendance", y = "Male spouse PSEA")


pic_cc_norway_income <- coord_cartesian(ylim = c(-0.2, 0.3))

pic_data_norway_income %>%
  filter(sex == 1) %>%
  ggplot(aes(Income)) + 
    geom_pointrange(aes(y = mean, ymin = lower, ymax = upper)) +
    pic_cc_norway_income +
    pic_theme + 
    labs(x = "Male spouse income decile", y = "Female spouse PSEA")

pic_data_norway_income %>%
  filter(sex == 2) %>%
  ggplot(aes(Income)) + 
    geom_pointrange(aes(y = mean, ymin = lower, ymax = upper)) +
    pic_cc_norway_income +
    pic_theme + 
    labs(x = "Female spouse income decile", y = "Male spouse PSEA")


```


```{r tbl-bo-first-stage-moba}

mod_bo_1st_moba <- purrr::map2(export_table1_tidyse,
                                     export_table1_glance, 
                                     convert_moba_for_huxreg)

names(mod_bo_1st_moba) <- c("University", "Income", "Height", "BMI")


tbl_note <- paste("Estimates from OLS regressions with the ",
"mediators (university attendance, income, height, BMI)",
" as dependent",
"variables, and own birth order as the main independent variable. PSEA is the",
"polygenic score for educational attainment, which is normalized with mean 0 and",
"standard deviation 1. We include parents' age at birth",
"and further controls to ensure the balance of covariates across birth", 
"order. All data is from the MoBa dataset for a sample of", 
"spouse pairs with a child between 1999 and 2008.", my_note)

coefs_1st <- c("Birth order", "Own PSEA", "Parents' age at birth")
huxreg(mod_bo_1st_moba, 
         coefs = moba_reg_coefs[coefs_1st],
         statistics = c(
           "N"  = "nobs", 
           "R2" = "r.squared"
          ),
         note = tbl_note,
         stars = my_stars,
         tidy_args = tidy_args
       ) %>% 
       insert_row(after = 7, "Family size dummies", rep("Yes", 4)) %>% 
       insert_row(after = 8, "Birth month dummies", rep("Yes", 4)) %>% 
       insert_row(after = 9, "Birth year dummies", rep("Yes", 4)) %>% 
       set_bottom_border(8:9, everywhere, 0) %>% 
       set_align(8:10, -1, "center") %>% 
       set_number_format(2:7, -1, 4) %>% 
       set_width(1) %>% 
       set_font_size(10) %>%
       set_caption("Regressions of mediators on birth order (Norway)")
       
```


```{r tbl-bo-psea-basic-moba}
mod_bo_psea_basic_moba <- purrr::map2(export_table2_tidyse,
                                     export_table2_glance, 
                                     convert_moba_for_huxreg)

names(mod_bo_psea_basic_moba) <- NULL


tbl_note <- paste("Estimates from OLS regressions with spouse PSEA as dependent",
"variable, and own birth order as the main independent variable. PSEA is the",
"polygenic score for educational attainment, which is normalized with mean 0 and",
"standard deviation 1. We include own PSEA, parents' age at birth (the mean of parents'",
"ages), and further controls (family size, birth year, and birth month dummies)", 
"in columns 2-3 to ensure the balance of covariates",
"across birth order. All data is from the MoBa dataset for a sample of", 
"spouse pairs with a child between 1999 and 2008.", my_note)

coefs_basic <-  c("Birth order", "Own PSEA", "Parents' age at birth")
huxreg(mod_bo_psea_basic_moba, 
         coefs = moba_reg_coefs[coefs_basic],
         statistics = c(
           "N"  = "nobs", 
           "R2" = "r.squared"
          ),
         note = tbl_note,
         stars = my_stars,
         tidy_args = tidy_args
       ) %>% 
       insert_row(after = 7, "Family size dummies", rep("Yes", 3)) %>% 
       insert_row(after = 8, "Birth month dummies", "No", "Yes", "Yes") %>% 
       insert_row(after = 9, "Birth year dummies", "No", "Yes", "Yes") %>% 
       set_bottom_border(8:9, everywhere, 0) %>% 
       set_align(8:10, -1, "center") %>% 
       set_number_format(2:7, -1, 4) %>% 
       set_width(0.9) %>% 
       set_font_size(10) %>%
       set_caption("Regressions of spouse PSEA on birth order (Norway)")


```


```{r tbl-bo-psea-no-par-age}


age_table <- famhist %>% 
      filter(n_sibs >=2, n_sibs <= 6, ! is.na(birth_order)) %>% 
      group_by(n_sibs, birth_order) %>% 
      dplyr::summarize(
        N   = n(), 
        mfa = mean(fath_age_birth, na.rm = TRUE),
        mma = mean(moth_age_birth, na.rm = TRUE),
        .groups = "drop"
      )

fml_bo_no_par_age <- lapply(fml_bo_psea, update, 
                     . ~ . - par_age_birth.x, 
                     rhs = 1
                   )
mod_bo_no_par_age <- lapply(fml_bo_no_par_age, fixest::feols,
                     data  = mf_pairs_reg, 
                     notes = FALSE
                   )

fml_bo_fath_age <- lapply(fml_bo_no_par_age, update, 
                     . ~ . + fath_age_birth.x, 
                     rhs = 1
                   )
mod_bo_fath_age <- lapply(fml_bo_fath_age, fixest::feols,
                     data  = mf_pairs_reg, 
                     notes = FALSE
                   )

huxreg(mod_bo_no_par_age, 
         coefs     = reg_coefs[reg_coefs != "par_age_birth.x"],
         note      = my_note,
         stars     = my_stars,
         tidy_args = tidy_args
       ) %>% 
       insert_row(after = 17, "Family size dummies", rep("Yes", 4)) %>% 
       insert_row(after = 18, "Birth month dummies", rep("Yes", 4)) %>% 
       insert_row(after = 19, "Birth year dummies", rep("Yes", 4)) %>% 
       set_bottom_border(18:19, everywhere, 0) %>% 
       set_align(18:20, -1, "center") %>% 
       set_number_format(2:17, -1, 4) %>% 
       set_font_size(11) %>%
       set_caption("Regressions of spouse PSEA, without controls for parents' age at respondent's birth (Great Britain)")

```


```{r tbl-bo-psea-dummies}

# we use standard lms because fixest::wald breaks sometimes

fml_bo_psea_dummies <- purrr::map(fml_bo_psea, formula, collapse = TRUE)

fml_bo_psea_dummies <- purrr::map(fml_bo_psea_dummies, update, 
                                    . ~ . - birth_order.x + 
                                    factor(birth_order.x))

# use a formula here to preserve the `data` argument
mod_bo_psea_dummies <- purrr::map(fml_bo_psea_dummies, ~lm(.x, 
                         data = mf_pairs_reg))

mod_restricted <- purrr::map(mod_bo_psea_dummies, 
                               ~update(.x, . ~ . - factor(birth_order.x)))

my_vcov <- function (x) sandwich::vcovHC(x, type ="HC1")
wald_tests <- purrr::map2(mod_bo_psea_dummies, mod_restricted,
                          lmtest::waldtest, vcov = my_vcov)
wald_pvals <- purrr::map_dbl(wald_tests, ~ pull(tidy(.x), p.value)[2])


dummy_reg_coefs <- grep("factor\\(birth_order", 
                         names(coef(mod_bo_psea_dummies[[1]])),
                         value = TRUE)
names(dummy_reg_coefs) <- gsub("^.*(\\d).*$", "Birth order \\1",
                                 dummy_reg_coefs)
dummy_reg_coefs <- c(
                      dummy_reg_coefs,
                      "University"  = "university.xTRUE", 
                      "Income"      = "first_job_pay.x",
                      "Own PSEA"     = "EA3.x",
                      "Parents' age at birth" = "par_age_birth.x"
                    )

ht <- huxtable::huxreg(mod_bo_psea_dummies, 
                   coefs     = dummy_reg_coefs,
                   note      = paste(my_note, 
                     "\nGrey background: coefficients are higher than column 1."),
                   stars     = my_stars,
                   tidy_args = tidy_args
                 ) %>% 
       insert_row(after = 19, "Wald p-value, birth order", wald_pvals) %>%
       insert_row(after = 20, "Family size dummies", rep("Yes", 4)) %>% 
       insert_row(after = 21, "Birth month dummies", rep("Yes", 4)) %>% 
       insert_row(after = 22, "Birth year dummies", rep("Yes", 4)) %>% 
       insert_row(after = 23, "Other mediators (IQ, height, BMI, s.-r. health)", 
                    c("No", rep("Yes", 3))) %>% 
       set_bottom_border(21:23, everywhere, 0) %>% 
       set_align(21:24, -1, "center") %>% 
       set_number_format(2:20, -1, 4) %>% 
       set_caption("Regressions of spouse PSEA, separate birth order dummies (Great Britain)") %>% 
       set_tb_padding(2) %>% 
       set_font_size(10) %>% 
       set_width(0.9)

coef_vals <- as.matrix(ht[seq(2, 10, 2), 2:5])
coef_vals <-  readr::parse_number(coef_vals)
dim(coef_vals) <- c(5, 4)
coefs_larger <- coef_vals[1:5, -1] > coef_vals[1:5, 1]
background_color(ht)[seq(2, 10, 2), 3:5] <- ifelse(coefs_larger, "grey90", 
                                                     "white")
ht

```


```{r tbl-bo-psea-dummies-moba}
mod_bo_psea_moba_dummies <- purrr::map2(export_table3dummy_tidyse,
                                  export_table3dummy_glance,
                                  convert_moba_for_huxreg)
names(mod_bo_psea_moba_dummies) <- NULL

moba_dummy_coefs <- paste0("factor(parity)", 2:6)
names(moba_dummy_coefs) <- paste("Birth order", 2:6)
moba_reg_coefs_dummies <- c(moba_dummy_coefs, moba_reg_coefs)
# Sometimes R syntax is awkward:
moba_reg_coefs_dummies <- moba_reg_coefs_dummies[
  names(moba_reg_coefs_dummies) != "Birth order"]

ht <- huxreg(mod_bo_psea_moba_dummies,
             coefs = moba_reg_coefs_dummies,
             statistics = c("N" = "nobs", "$R^2$" = "r.squared"),
             note       = paste(my_note, 
                         "\nGrey background: coefficients are higher than column 1."),
             stars      = my_stars
           ) %>% 
           insert_row(after = 23, "Family size dummies", rep("Yes", 4)) %>% 
           insert_row(after = 24, "Birth month dummies", rep("Yes", 4)) %>% 
           insert_row(after = 25, "Birth year dummies", rep("Yes", 4)) %>% 
           set_bottom_border(24:25, everywhere, 0) %>% 
           set_align(24:26, -1, "center") %>% 
           set_number_format(2:22, -1, 4) %>% 
           set_tb_padding(2) %>% 
           set_width(1) %>% 
           set_font_size(10) %>%
           set_escape_contents(20, 1, FALSE) %>%
           set_caption("Regressions of spouse PSEA, separate birth order dummies (Norway)")


coef_vals <- as.matrix(ht[seq(2, 10, 2), 2:5])
coef_vals <-  readr::parse_number(coef_vals)
dim(coef_vals) <- c(5, 4)
coefs_larger <- coef_vals[1:5, -1] > coef_vals[1:5, 1]
background_color(ht)[seq(2, 10, 2), 3:5] <- ifelse(coefs_larger, "grey90", 
                                                     "white")


ht
```


```{r calc-bo-mediator-interactions}

fml_bo_mediator_interact <- update(fml_bo_psea_dummies[[4]],
                                     . ~ . * factor(birth_order.x)) 
mod_bo_mediator_interact <- fixest::feols(fml_bo_mediator_interact, 
                                            data = mf_pairs_reg, 
                                            note = FALSE)

tidy_bo_mediator <- do.call(broom::tidy, c(list(x = mod_bo_mediator_interact), 
                                           tidy_args))

tidy_bo_mediator$sig <- tidy_bo_mediator$p.value < 0.05
n_interactions <- sum(grepl(":", tidy_bo_mediator$term))
tidy_bo_mediator$sig_corrected <- tidy_bo_mediator$p.value < 0.05/n_interactions
```


```{r pic-bo-psea-interactions, fig.align = "center", fig.cap = "Regressions of spouse PSEA: birth order dummies within different family sizes (Great Britain). Labels show birth order. Lines are 95 per cent confidence intervals. The omitted category is birth order 1."}
 
mf_pairs_reg$BO2 <- 1 * (mf_pairs_reg$birth_order.x == 2)
mf_pairs_reg$BO3 <- 1 * (mf_pairs_reg$birth_order.x == 3)
mf_pairs_reg$BO4 <- 1 * (mf_pairs_reg$birth_order.x == 4)
mf_pairs_reg$BO5 <- 1 * (mf_pairs_reg$birth_order.x == 5)
mf_pairs_reg$BO6 <- 1 * (mf_pairs_reg$birth_order.x == 6)
mf_pairs_reg$nsf <- factor(mf_pairs_reg$n_sibs.x)

fml_bo_psea_full <- update(fml_bo_psea[[1]], . ~ . - birth_order.x + 
                             nsf:(BO2 + BO3 + BO4 + BO5 + BO6))

mod_bo_psea_full <- fixest::feols(fml_bo_psea_full, mf_pairs_reg, 
                                    notes = FALSE)
coefs <- tidy(mod_bo_psea_full, conf.int = TRUE, se = tidy_args$se,
                cluster = tidy_args$cluster)

suppressWarnings({
  coefs[["Family size"]] <- as.numeric(gsub("^nsf(.).*", "\\1", coefs$term))
  coefs[["Birth order"]] <- as.numeric(gsub(".*BO(.)$", "\\1", coefs$term))
})

coefs %<>% filter(`Birth order` %in% 1:8)

ns_by_n_sibs <- as.data.frame(table(mf_pairs_reg$n_sibs.x))
ns_by_n_sibs$Var1 <- as.integer(as.character(ns_by_n_sibs$Var1))

suppressWarnings({
  gg <- ggplot(coefs, aes(estimate, `Family size`, colour = `Birth order`)) + 
          scale_color_continuous(type = "viridis", name = "Birth order") + 
          geom_vline(xintercept = 0, linetype = 2) +
          geom_pointrange(aes(xmin = conf.low, xmax = conf.high), 
                         alpha = 0.7, size = 1.15, fatten = 1,
                         position = position_dodge2(width = 0.15)) + 
          geom_text(aes(label = `Birth order`), nudge_y = 0.2, size = 5) + 
          guides(colour = "none") +
          xlab("Estimate")
  print(gg) # captures the warnings inside the suppressWarnings() call
})

```


```{r tbl-bo-psea-weights}

mod_bo_psea_weights <- purrr::map(fml_bo_psea, fixest::feols, 
                                    data = mf_pairs_reg,
                                    weights = mf_pairs_reg$weights.x, 
                                    notes = FALSE
                                  )


huxreg(mod_bo_psea_weights, 
         coefs     = reg_coefs,
         note      = my_note,
         stars     = my_stars,
         tidy_args = tidy_args
       ) %>% 
       insert_row(after = 19, "Family size dummies", rep("Yes", 4)) %>% 
       insert_row(after = 20, "Birth month dummies", rep("Yes", 4)) %>% 
       insert_row(after = 21, "Birth year dummies", rep("Yes", 4)) %>% 
       set_bottom_border(20:21, everywhere, 0) %>% 
       set_align(18:20, -1, "center") %>% 
       set_number_format(2:17, -1, 4) %>% 
       set_font_size(10) %>%
       set_caption("Regressions of spouse PSEA, weighted to match 
                   UK Biobank sampling frame (Great Britain)")

```


```{r tbl-bo-psea-pgs}

fml_bo_psea_pgs <- fml_bo_psea %>% 
                        purrr::map(update, 
                           . ~ . + 
                           alzheimer_resid.x + 
                           cognitive_ability_resid.x +
                           neuroticism_resid.x +
                           sc_substance_use_resid.x
                        )

mod_bo_psea_pgs <-  purrr::map(fml_bo_psea_pgs, fixest::feols, 
                         data = mf_pairs_reg,
                         notes = FALSE
                       )

huxreg(mod_bo_psea_pgs, 
         coefs = reg_coefs,
         statistics = c(
           "N" = "nobs", 
           "R2" = "r.squared"
         ),
         note = paste(my_note,
                 "\nPolygenic scores: alzheimer's, cognitive ability, neuroticism, substance use."),
         stars = my_stars,
         tidy_args = list(
           conf.int = FALSE, 
           cluster = list(mf_pairs_reg$f.54.0.0.x)
         )
       ) %>% 
       insert_row(after = 19, "Family size dummies", rep("Yes", 4)) %>% 
       insert_row(after = 20, "Birth month dummies", rep("Yes", 4)) %>% 
       insert_row(after = 21, "Birth year dummies", rep("Yes", 4)) %>% 
       insert_row(after = 22, "Polygenic score controls", rep("Yes", 4)) %>% 
       set_bottom_border(20:22, everywhere, 0) %>% 
       set_align(20:23, -1, "center") %>% 
       set_number_format(2:19, -1, 4) %>% 
       set_width(1) %>% 
       set_wrap(final(1), everywhere, TRUE) %>% 
       set_tb_padding(2) %>%
       set_font_size(10) %>%
       set_caption("Regressions of spouse PSEA with controls for polygenic 
                   scores (Great Britain)")
```


```{r tbl-bo-psea-age-fte}

fml_bo_psea_age_fte <- fml_bo_psea[c(1, 2, 4)]
fml_bo_psea_age_fte[2:3] <- lapply(fml_bo_psea_age_fte[2:3], 
                                update, 
                                . ~ . - university.x + age_fulltime_edu.x)

mod_bo_psea_age_fte <- lapply(fml_bo_psea_age_fte, fixest::feols, 
                 data = mf_pairs_reg,
                 notes = FALSE
               )


age_reg_coefs <- reg_coefs
age_reg_coefs[age_reg_coefs == "university.xTRUE"] <- "age_fulltime_edu.x"
names(age_reg_coefs)[age_reg_coefs == "age_fulltime_edu.x"] <- 
      "Age left full-time educ."

huxreg(mod_bo_psea_age_fte, 
         coefs     = age_reg_coefs,
         note      = my_note,
         stars     = my_stars,
         tidy_args = tidy_args
       ) %>% 
       insert_row(after = 19, "Family size dummies", rep("Yes", 3)) %>% 
       insert_row(after = 20, "Birth month dummies", rep("Yes", 3)) %>% 
       insert_row(after = 21, "Birth year dummies", rep("Yes", 3)) %>% 
       set_bottom_border(20:21, everywhere, 0) %>% 
       set_align(20:22, -1, "center") %>% 
       set_number_format(2:19, -1, 4) %>% 
       set_width(0.9) %>% 
       set_font_size(10) %>%
       set_caption("Regressions of spouse PSEA using age of leaving full-time 
                   education (Great Britain)")
```



```{r tbl-bo-psea-age-fte-moba}

mod_bo_psea_age_fte_moba <- purrr::map2(export_table3eduyears_tidyse[c(1, 2, 4)],
                                  export_table3eduyears_glance[c(1, 2, 4)],
                                  convert_moba_for_huxreg)
names(mod_bo_psea_age_fte_moba) <- NULL


moba_reg_coefs_age_fte <- moba_reg_coefs
names(moba_reg_coefs_age_fte)[moba_reg_coefs_age_fte=="university"] <- 
  "Age left fulltime educ."
moba_reg_coefs_age_fte[moba_reg_coefs_age_fte=="university"] <- "eduyears"

huxreg(mod_bo_psea_age_fte_moba, 
         coefs     = moba_reg_coefs_age_fte,
         note      = my_note,
         stars     = my_stars,
         tidy_args = tidy_args
       ) %>% 
       insert_row(after = 15, "Family size dummies", rep("Yes", 3)) %>% 
       insert_row(after = 16, "Birth month dummies", rep("Yes", 3)) %>% 
       insert_row(after = 17, "Birth year dummies", rep("Yes", 3)) %>% 
       set_bottom_border(16:17, everywhere, 0) %>% 
       set_align(16:18, -1, "center") %>% 
       set_number_format(2:14, -1, 4) %>% 
       set_width(0.9) %>% 
       set_font_size(10) %>%
       set_caption("Regressions of spouse PSEA using age of leaving full-time 
                   education (Norway)")
```


```{r tbl-bo-psea-no3}

mod_bo_psea_no3 <- lapply(fml_bo_psea, 
                            fixest::feols, 
                            data = mf_pairs_reg %>% 
                                     filter(n_sibs.x != 3),
                            notes = FALSE
                          )
huxreg(mod_bo_psea_no3, 
         coefs     = reg_coefs,
         note      = my_note,
         stars     = my_stars,
         tidy_args = tidy_args
       ) %>% 
       insert_row(after = 19, "Family size dummies", rep("Yes", 4)) %>% 
       insert_row(after = 20, "Birth month dummies", rep("Yes", 4)) %>% 
       insert_row(after = 21, "Birth year dummies", rep("Yes", 4)) %>% 
       set_bottom_border(20:21, everywhere, 0) %>% 
       set_align(20:22, -1, "center") %>% 
       set_number_format(2:19, -1, 4) %>% 
       set_font_size(10) %>%
       set_caption("Regressions of spouse PSEA, excluding family size 3 
                   (Great Britain)")
```


\FloatBarrier
\newpage

### Testing the UK Biobank spouse pair matching 


```{r calc-validate-pairs}

prop_shared <- calc_prop_shared_children(mf_pairs)
n_one_has_kid <- prop_shared$one
n_both_same_kid <- prop_shared$both

```


Some of our spouse pairs in UK Biobank could be false positives, i.e. people who
are not each others' spouse but simply live in the same postcode. To validate
the accuracy of our pairs, we use genetic relationships. Some respondents in the
UK Biobank sample have a child (inferred from genetic data) who is also in the
sample. Among our spouse pairs, `r n_one_has_kid` have a genetic child of at
least one partner in the sample. For `r n_both_same_kid` of these, at least one
child is the genetic child of both partners. If this subsample is
representative, then about `r percent(n_both_same_kid/n_one_has_kid)` of the
pairs who have had a child, have had a child together. This is a lower bound
estimate, because some of the remaining couples may have had a genetically
shared child who is not in the UK Biobank sample. As a point of comparison, 11%
of families with dependent children included a stepchild in England and Wales in
2011 [@ons2011stepfamilies].

It is still possible that some pairs in our data may not be actual spouses.
These pairs might show a relationship between one partner's phenotype and the
other's genotype. For example, maybe early-born children grow up to live in
richer postcodes, along with people who have higher PSEA scores
[@abdellaoui2019genetic]. This could then bias the results. If the coefficient
for "fake pairs" is absolutely larger (smaller) than for real pairs, then our
results will be biased away from zero (towards zero).

To sign the bias, we create a dataset of "known fake pairs". These are
opposite-sexed pairs who live in the same postcode, but do not share all
the characteristics listed for the real pairs. Specifically, from the
list of characteristics used to create our real pairs (same
homeownership status, same length of time at address, same number of
children, attended same assessment center, attended on same day, husband
reported living with spouse, wife reported living with spouse) the fake
pairs ticked exactly 5 out of 7 boxes.

```{r calc-fake-pairs}

drake::loadd(mf_pairs_fake)
prop_shared_fake <- calc_prop_shared_children(mf_pairs_fake)

n_one <- prop_shared_fake$one
n_both <- prop_shared_fake$both
```

We again use genetic children to confirm that the fake pairs are "real
fakes". Out of `r n_one` genetic children of the fake pairs, only
`r n_both` were children of both parents. Thus, the vast majority of
fake pairs do not appear to be spouses. Table
\@ref(tab:tbl-bo-psea-fake) reruns the regressions of Table
\@ref(tab:tbl-bo-psea-basic) using the fake pairs. Although the
coefficients on birth order are always negative, and significant
when controlling for parent's age, they are always absolutely smaller than the
corresponding coefficient in the main text. This suggests that
any fake pairs remaining in our data will have the effect of biasing our
results towards zero.

```{r tbl-bo-psea-fake}

drake::loadd(mf_fake_twice)
fake_pairs_reg <- regression_subset(mf_fake_twice)
mod_bo_psea_fake <- lapply(fml_bo_psea_base, fixest::feols, 
                 data = fake_pairs_reg,
                 notes = FALSE
               )

huxreg(mod_bo_psea_fake, 
         coefs = c(
           "Birth order"           = "birth_order.x", 
           "Own PSEA"              = "EA3.x",
           "Parents' age at birth" = "par_age_birth.x"
         ),
         statistics = c(
           "N"  = "nobs", 
           "R2" = "r.squared"
          ),
         note = my_note,
         stars = my_stars,
         tidy_args = tidy_args
       ) %>% 
       insert_row(after = 7, "Family size dummies", rep("Yes", 3)) %>% 
       insert_row(after = 8, "Birth month dummies", "No", "Yes", "Yes") %>% 
       insert_row(after = 9, "Birth year dummies", "No", "Yes", "Yes") %>% 
       set_bottom_border(8:9, everywhere, 0) %>% 
       set_align(8:10, -1, "center") %>% 
       set_number_format(2:7, -1, 4) %>% 
       set_width(0.8) %>% 
       set_caption("Regressions of PSEA on birth order: fake pairs (Great Britain)")

```


\FloatBarrier

\clearpage

## Quotations on natural inequality

...your face and figure have nothing of the slave about them, and
proclaim you of noble birth.

-- *Odyssey*, Odysseus to Laertes

Citizens, we shall say to them in our tale, you are brothers, yet God
has framed you differently. Some of you have the power of command, and
in the composition of these he has mingled gold, wherefore also they
have the greatest honour; others he has made of silver, to be
auxiliaries; others again who are to be husbandmen and craftsmen he has
composed of brass and iron; and the species will generally be preserved
in the children. But as all are of the same original stock, a golden
parent will sometimes have a silver son, or a silver parent a golden
son.

-- Plato *Republic*

Nature would like to distinguish between the bodies of freemen and
slaves, making the one strong for servile labor, the other upright, and
although useless for such services, useful for political life in the
arts both of war and peace. But the opposite often happens -- that some
have the souls and others have the bodies of freemen.

-- Aristotle *Politics*

Sons have no richer endowment than the quality

A noble and brave father gives in their begetting.

-- Euripides *Heracleidae*


His head by nature fram'd to wear a crown,

His hands to wield a sceptre....

-- Shakespeare *Henry VI Part 3*

A daughter of a green Grocer, walks the Streets in London dayly with a
baskett of Cabbage Sprouts, Dandelions and Spinage on her head. She is
observed by the Painters to have a beautiful Face, an elegant figure, a
graceful Step and a debonair. They hire her to Sitt. She complies, and
is painted by forty Artists, in a Circle around her. The Scientific Sir
William Hamilton outbids the Painters, Sends her to Schools for a
genteel Education and Marries her. This Lady not only causes the
Tryumphs of the Nile of Copenhagen and Trafalgar, but Seperates Naples
from France and finally banishes the King and Queen from Sicilly. Such
is the Aristocracy of the natural Talent of Beauty.

-- John Adams to Thomas Jefferson, on Emma Hamilton

\newpage

# References
